---
title: "91 · Per-sample set scores (ssGSEA)"
subtitle: "ssGSEA scores, tested on the design and read against the phenotype"
params:
  score_fdr: 0.05
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    code-fold: true
    df-print: kable
    fig-width: 7.5
    fig-height: 5
execute:
  message: false
  warning: false
---

`90_Set_Test_Calibration` asks whether a set moved in a contrast. This stage asks a different
question: how much of each set is present in each sample. The output is one number per set per
sample, a 1,040 × 131 matrix, and that turns a pathway into a variable that can be plotted against
the muscle measurements rather than only reported with a p-value.

Two things are lost in the conversion and have to be said out loud. ssGSEA ranks within a sample,
so the precision weights from limpa do not travel; a protein reconstructed largely from missing
values carries the same rank as a well-measured one. And the scores are relative to the cohort:
adding or removing samples changes them. Neither makes the scores wrong, but both mean the
whole-set tests in `90_Set_Test_Calibration`, not these scores, carry error control for the
contrasts. The main pathway stages score samples with singscore, which is sample-independent;
`92_Method_Comparison` sets the two side by side.

```{r setup}
required <- c("here", "limma", "GSVA", "dplyr", "tibble", "tidyr", "purrr",
              "readr", "ggplot2")
missing <- required[!vapply(required, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing)) stop("Install required R packages: ", paste(missing, collapse = ", "))
library(here)
library(limma)
library(dplyr)
library(tibble)
library(tidyr)
library(purrr)
library(ggplot2)
here::i_am("03_Pathway_Enrichment/91_Sample_Scores_ssGSEA/a_script/91_sample_scores_ssgsea.qmd")

stage <- here("03_Pathway_Enrichment", "91_Sample_Scores_ssGSEA")
out <- file.path(stage, "c_data")
dir.create(out, recursive = TRUE, showWarnings = FALSE)

# The same frozen sets as 90_Set_Test_Calibration, so both stages test exactly one list.
frozen_path <- here("03_Pathway_Enrichment", "90_Set_Test_Calibration", "c_data",
                    "tested_sets.rds")
stopifnot("Render 90_Set_Test_Calibration first" = file.exists(frozen_path))
x <- readRDS(frozen_path)
proteins <- readRDS(here("01_Preprocess", "02_Quantification", "c_data", "proteins.rds"))
d <- readRDS(here("02_Differential_Expression", "01_Design", "c_data", "design.rds"))
design <- d$design
contrasts <- d$contrasts
control <- d$negative_control
targets <- as_tibble(proteins$targets)

stopifnot(identical(colnames(x$gene_matrix), rownames(design)),
          all(vapply(x$gene_sets, \(g) all(g %in% rownames(x$gene_matrix)), logical(1))))
tibble(n_genes = nrow(x$gene_matrix), n_samples = ncol(x$gene_matrix),
       n_sets = length(x$gene_sets), negative_control = control)
```

## Scoring

`gene_matrix` is one fixed representative row per gene, chosen upstream without using any
p-value or fold change, so the same representative serves every contrast.

```{r score}
set.seed(42)
param <- GSVA::ssgseaParam(x$gene_matrix, x$gene_sets, minSize = 1, maxSize = Inf,
                           normalize = TRUE)
scores <- GSVA::gsva(param, verbose = FALSE)
stopifnot(identical(dim(scores), c(length(x$gene_sets), ncol(x$gene_matrix))),
          identical(colnames(scores), colnames(x$gene_matrix)),
          all(is.finite(scores)))
tibble(sets = nrow(scores), samples = ncol(scores),
       score_min = signif(min(scores), 3), score_median = signif(median(scores), 3),
       score_max = signif(max(scores), 3))
```

## Testing the scores on the same design

The scores go into the same fixed-participant design, so a score is compared only within a
person, exactly as the protein-level model does. No precision weights: they did not survive the
ranking. `trend = TRUE` because set scores carry a mean-variance relationship of their own.

The negative control is read first here too. It is the same true null, and a score-level test
that finds pathways between two untrained legs is telling us the same thing a protein-level one
would.

```{r score-tests}
fit <- lmFit(scores, design) |> contrasts.fit(contrasts) |> eBayes(trend = TRUE)

score_results <- map(colnames(contrasts), \(cn) {
  topTable(fit, coef = cn, number = Inf, adjust.method = "none", sort.by = "none") |>
    rownames_to_column("set_id") |>
    transmute(set_id, contrast = cn, diff = logFC, t, p = P.Value)
}) |>
  list_rbind() |>
  separate_wider_delim(set_id, "|", names = c("database", "pathway"), cols_remove = FALSE) |>
  mutate(padj = p.adjust(p, "BH"), .by = c(contrast, database))

score_results |>
  summarise(sets = n(), sig = sum(padj < params$score_fdr),
            median_p = round(median(p), 3),
            p_below_05 = round(mean(p < 0.05), 3), .by = contrast) |>
  arrange(match(contrast, c(control, setdiff(colnames(contrasts), control))))
```

```{r control-check}
score_results |>
  filter(contrast == control, padj < params$score_fdr) |>
  slice_min(padj, n = 15) |>
  transmute(database, pathway, diff = signif(diff, 3), padj = signif(padj, 3))
```

Sets above are false positives by construction. Compare the count with
`90_Set_Test_Calibration`: if the score-level test is looser than `fry` on the same null, its
results on the interaction inherit that looseness.

```{r hallmark-scores}
score_results |>
  filter(database == "Hallmark", padj < params$score_fdr) |>
  arrange(contrast, padj) |>
  transmute(contrast, pathway, diff = signif(diff, 3), padj = signif(padj, 3))
```

## Against the phenotype

This is what per-sample scores are for. A set whose change tracks the change in muscle
cross-sectional area is worth more than a set with a small p-value, because the phenotype is the
outcome the study was designed around.

The unit is a leg: one score change (T2 − T1) and one CSA change per leg. Correlation is computed
**within each treatment separately**, so each person contributes one leg to each correlation and
the observations are independent. Pooling both legs would put the same participant in twice.

Only sets that the score-level test flagged are correlated. Correlating all 1,040 against the
phenotype would be several thousand tests chasing a handful of real relationships.

```{r phenotype}
pheno <- readr::read_csv(here("00_Input", "phenotype.csv"), show_col_types = FALSE) |>
  mutate(d_csa = vl_csa_post_cm2 - vl_csa_pre_cm2) |>
  select(participant, leg, treatment, d_csa)

score_long <- as_tibble(scores, rownames = "set_id") |>
  pivot_longer(-set_id, names_to = "sample_id", values_to = "score") |>
  left_join(select(targets, sample_id, participant, leg, treatment, timepoint),
            by = "sample_id")

delta <- score_long |>
  pivot_wider(id_cols = c(set_id, participant, leg, treatment),
              names_from = timepoint, values_from = score) |>
  filter(!is.na(T1), !is.na(T2)) |>
  mutate(d_score = T2 - T1) |>
  inner_join(pheno, by = c("participant", "leg", "treatment"))

tested_sets <- score_results |>
  filter(padj < params$score_fdr, contrast != control) |>
  distinct(set_id) |>
  pull(set_id)

correlations <- delta |>
  filter(set_id %in% tested_sets) |>
  summarise(
    n = n(),
    rho = cor(d_score, d_csa, method = "spearman"),
    p = cor.test(d_score, d_csa, method = "spearman", exact = FALSE)$p.value,
    .by = c(set_id, treatment)
  ) |>
  mutate(padj = p.adjust(p, "BH"), .by = treatment) |>
  separate_wider_delim(set_id, "|", names = c("database", "pathway"), cols_remove = FALSE)

tibble(sets_correlated = n_distinct(correlations$set_id),
       legs_per_treatment = paste(range(correlations$n), collapse = "-"),
       sig_BFR = sum(correlations$padj < 0.05 & correlations$treatment == "BFR"),
       sig_HLRT = sum(correlations$padj < 0.05 & correlations$treatment == "HLRT"))
```

```{r top-correlations}
correlations |>
  slice_min(padj, n = 20) |>
  arrange(padj) |>
  transmute(treatment, database, pathway, n, rho = round(rho, 3), padj = signif(padj, 3))
```

```{r correlation-plot}
top <- correlations |> slice_min(padj, n = 4) |> pull(set_id) |> unique()
if (length(top)) {
  delta |>
    filter(set_id %in% top) |>
    mutate(pathway = sub("^[^|]+\\|", "", set_id)) |>
    ggplot(aes(d_score, d_csa, colour = treatment)) +
    geom_point(alpha = 0.8) +
    geom_smooth(method = "lm", se = FALSE, linewidth = 0.6, formula = y ~ x) +
    facet_wrap(~pathway, scales = "free_x", labeller = label_wrap_gen(34)) +
    labs(x = "change in set score (T2 - T1)", y = expression(Delta ~ "VL CSA (cm"^2 * ")"),
         colour = NULL) +
    theme_minimal() +
    theme(legend.position = "top")
}
```

A correlation here is a relationship between two changes measured on the same leg, not evidence
that the pathway caused the growth. Both would move together under any shared upstream cause,
including how much that leg was loaded.

## Alternative: every Hallmark set, unfiltered

The analysis above correlates only the sets whose mean score changed with training. That filter
keeps the number of tests down, but it selects on the average, and a set can track growth without
moving on average: up in the legs that grew most, down in those that grew least, near zero overall.
Those are exactly the sets a correlation exists to find, and the filter removes them before they
are tested.

So the same correlation is run here on all Hallmark sets, with no filter. Hallmark was declared the
primary collection before any test ran, so the set list is fixed in advance and not chosen after
seeing results, and at around forty sets per arm the FDR correction costs little. Reactome and
GO:BP stay exploratory.

```{r hallmark-all}
hallmark_ids <- grep("^Hallmark\\|", unique(delta$set_id), value = TRUE)

correlations_hallmark <- delta |>
  filter(set_id %in% hallmark_ids) |>
  summarise(
    n = n(),
    rho = cor(d_score, d_csa, method = "spearman"),
    p = cor.test(d_score, d_csa, method = "spearman", exact = FALSE)$p.value,
    .by = c(set_id, treatment)
  ) |>
  mutate(padj = p.adjust(p, "BH"), .by = treatment) |>
  separate_wider_delim(set_id, "|", names = c("database", "pathway"), cols_remove = FALSE)

correlations_hallmark |>
  summarise(sets = n(), sig_FDR = sum(padj < 0.05), raw_p_below_05 = sum(p < 0.05),
            expected_by_chance = round(0.05 * n(), 1), smallest_FDR = signif(min(padj), 3),
            .by = treatment)
```

```{r hallmark-all-table}
correlations_hallmark |>
  arrange(treatment, p) |>
  slice_head(n = 10, by = treatment) |>
  transmute(treatment, pathway = sub("^HALLMARK_", "", pathway), n,
            rho = round(rho, 3), p = signif(p, 3), padj = signif(padj, 3))
```

Whether a set tracks growth differently in the two arms is a separate question from whether it
tracks growth in either arm. Being significant in one arm and not the other is not evidence of a
difference, so the arms are compared directly, with Fisher's z on the two correlations.

```{r hallmark-all-difference}
arm_difference <- correlations_hallmark |>
  select(set_id, database, pathway, treatment, rho, n) |>
  pivot_wider(names_from = treatment, values_from = c(rho, n)) |>
  mutate(
    z = (atanh(rho_HLRT) - atanh(rho_BFR)) / sqrt(1 / (n_HLRT - 3) + 1 / (n_BFR - 3)),
    p_difference = 2 * pnorm(-abs(z)),
    padj_difference = p.adjust(p_difference, "BH")
  )

arm_difference |>
  arrange(p_difference) |>
  slice_head(n = 10) |>
  transmute(pathway = sub("^HALLMARK_", "", pathway),
            rho_BFR = round(rho_BFR, 3), rho_HLRT = round(rho_HLRT, 3),
            p_difference = signif(p_difference, 3), padj_difference = signif(padj_difference, 3))
```

## Export

```{r export}
saveRDS(scores, file.path(out, "set_scores.rds"))
readr::write_csv(score_results, file.path(out, "score_contrast_results.csv"))
readr::write_csv(correlations, file.path(out, "score_phenotype_correlations.csv"))
readr::write_csv(correlations_hallmark,
                 file.path(out, "score_phenotype_correlations_hallmark_all.csv"))
readr::write_csv(arm_difference, file.path(out, "score_phenotype_arm_difference_hallmark.csv"))
tibble(file = list.files(out), bytes = file.size(file.path(out, list.files(out))))
```

```{r session}
sessionInfo()
```
