---
title: "01 · Filtering"
subtitle: "DIA-NN report to a filtered precursor matrix"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    df-print: kable
execute:
  message: false
---

`readDIANN()` reads the DIA-NN report, applies q-value cutoffs and returns log2 precursor
intensities. Every step after it removes rows and never columns: annotation artefacts of
the search FASTA, precursors shared between protein groups, compound groups, and proteins
skeletal muscle does not express. No sample is dropped and no missing value is filtered.

```{r setup}
pacman::p_load(here, limpa, dplyr, stringr, purrr, readr, tibble)

CONTAMINANT_FAMILIES <- str_c(
  "^(IGH|IGK|IGL|JCHAIN$|KRT|SPRR|HRNR$|DCD$|SBSN$|CSTA$|HB[ABDGMQZ][0-9]?$|AHSP$)"
)

# From a review of all 3,750 gene symbols in this dataset (Claude, claude-sonnet-5,
# 2026-09-04), judged on HPA myonuclei single-cell expression rather than whole-muscle
# tissue. Whole-tissue values are themselves blood-contaminated, which is why haemoglobin
# reads 544 nTPM in skeletal muscle. 93 of the 239 symbols it returned collapse into the
# family patterns above; every symbol those patterns match was independently flagged.
CONTAMINANTS <- c(
  "A1BG", "AFM", "AGT", "AHSG", "ALB", "AMBP", "APCS", "APOA1", "APOA2", "APOA4", "APOB",
  "APOC1", "APOC2", "APOC3", "APOD", "APOE", "APOF", "APOH", "APOL1", "APOL2", "APOM",
  "ARHGDIB", "AZGP1", "BCAM", "BTD", "C1QB", "C1QC", "C1R", "C1RL", "C1S", "C2", "C3",
  "C4A", "C4B", "C4BPA", "C4BPB", "C5", "C6", "C7", "C8A", "C8B", "C8G", "C9", "CA1",
  "CFB", "CFD", "CFH", "CFHR2", "CFHR5", "CFI", "CLEC3B", "CLU", "CP", "CPB2", "CPN1",
  "CPN2", "CTSG", "DEFA1", "DMTN", "ELANE", "EPB42", "F10", "F11", "F12", "F13A1", "F13B",
  "F2", "F9", "FCN3", "FERMT3", "FETUB", "FGA", "FGB", "FGG", "FYB1", "GC", "GIMAP1",
  "GIMAP4", "GMFG", "GP1BA", "GPLD1", "GYPC", "HABP2", "HP", "HPR", "HPX", "HRG",
  "IGFALS", "ITIH1", "ITIH3", "KEL", "KLKB1", "KNG1", "LBP", "LCAT", "LCP1", "LPA",
  "LRG1", "LTF", "LYZ", "MASP1", "MPO", "MPP1", "NCF1", "ORM1", "ORM2", "PF4", "PGLYRP2",
  "PLG", "PLTP", "PON1", "PPBP", "PROS1", "PRTN3", "PZP", "RBP4", "RHCE", "RNASE3",
  "S100A8", "S100A9", "SAA1", "SAA4", "SELENOP", "SERPINA1", "SERPINA3", "SERPINA4",
  "SERPINA5", "SERPINA6", "SERPINA7", "SERPINC1", "SERPIND1", "SERPINF2", "SERPING1",
  "SHBG", "SLC4A1", "SPTA1", "TF", "TTR", "TUBB1", "VTN", "VWF"
)

# Asserted to survive every filter below. The first two rows are muscle proteins the
# annotation repair exists to rescue. The rest are blood-abundant but muscle-expressed on
# HPA myonuclei data: ANK1 is small ankyrin-1, a sarcoplasmic reticulum protein, and the
# review did flag A2M, BLVRB, STOM and TALDO1, which carry 0.004% of signal between them.
MUST_KEEP <- c(
  "MB", "ACTA1", "ALDOA", "TPM2", "ENO1", "GPI", "CKM", "MYH1",
  "MYH2", "MYH7", "TTN", "DES", "PYGM", "CASQ1", "PVALB", "MYH13",
  "ANK1", "SPTB", "ITIH2", "ITIH4", "SERPINF1", "CAT", "CA2", "CA3",
  "A2M", "BLVRB", "STOM", "TALDO1"
)
```

## Read

`Protein.Ids` is not one of `readDIANN()`'s default `annotation.columns`, and the repair
below cannot run without it. It is added to the defaults, not substituted for them, so
`Protein.Names` still comes through for later annotation.

Everything else stays as shipped. `q.cutoffs = 0.01` over `Q.Value`, `Lib.Q.Value` and
`Lib.PG.Q.Value` is the set the vignette recommends for a match-between-runs search.
`intensity.column = "Precursor.Normalised"` is DIA-NN's normalized column, which is why
nothing here normalizes again. `log = TRUE` means `y$E` arrives on the log2 scale, so never
log it a second time.

```{r read}
#| cache: true
y <- readDIANN(
  here("00_Input", "report.parquet"),
  annotation.columns = c(
    "Protein.Group", "Protein.Ids", "Protein.Names",
    "Genes", "Proteotypic"
  )
)
dim(y)
```

## Map runs to samples

Columns come back named by DIA-NN `Run`. Matching them case-insensitively against
`metadata$run` does three jobs at once: it renames columns to `sample_id`, it absorbs a run
the instrument wrote `Cl_MDR_7E-T2` against `CL_MDR_7E-T2` in the sheet, and it drops the
discarded S08 injection, which is in the report but not the sheet.

```{r map-samples}
metadata <- read_csv(here("00_Input", "metadata.csv"), show_col_types = FALSE)
i <- match(str_to_lower(metadata$run), str_to_lower(colnames(y)))
stopifnot(!any(is.na(i)))

runs_dropped <- setdiff(colnames(y), colnames(y)[i])
y <- y[, i]
colnames(y) <- metadata$sample_id
y$targets <- as.data.frame(metadata)

signal_start <- sum(2^y$E, na.rm = TRUE)
c(runs_in_report = length(i) + length(runs_dropped), samples_kept = ncol(y))
runs_dropped
```

## Check what survived the mapping

`metadata.csv` decides which runs enter the matrix, so a failed injection is excluded by
name rather than by rule. This chunk drops nothing; it asserts that the exclusion list was
right. A failed injection identifies a fraction of the cohort's precursors while its median
observed intensity runs *high*, because only the brightest peaks survive.

```{r run-qc}
depth <- colSums(!is.na(y$E))
run_qc <- tibble(
  sample_id = colnames(y),
  precursors = depth,
  median_log2 = apply(y$E, 2, median, na.rm = TRUE),
  pct_missing = 100 * colMeans(is.na(y$E))
)
stopifnot(all(depth > 0.5 * median(depth)))
arrange(run_qc, precursors) |> head(5)
```

## Repair the annotation

The search FASTA held every protein twice, plain and `cRAP-`-prefixed, so myoglobin is
reported as `P02144;cRAP-P02144` with `Proteotypic = 0`. Dropping every group that mentions
`cRAP` would take CKM, MB, ALDOA, CAT, DES and FABP3 with it. What is wanted is to drop only
the groups where *every* member is cRAP, then strip the prefix from the rest.

The order matters, because `Genes` carries the prefix too. Six cRAP-only symbols appear
here, and one of them is `cRAP-MB`. Strip the prefix first and it becomes `MB`, which
`MUST_KEEP` requires to survive, so ten precursors of a contaminant standard would merge
into human myoglobin and nobody would see it. `cRAP-CSN1S1` is not in `CONTAMINANTS`
either, and two cRAP-only rows carry no gene symbol at all, so they match nothing and would
survive as a group standing for no real protein. The contaminant list cannot do this job.

Stripping also repairs `Proteotypic`, which is the whole reason the muscle survives:
`filterNonProteotypicPeptides()` believes the flag, and on the unrepaired annotation it
deletes 13 of the 28 `MUST_KEEP` symbols.

```{r repair}
members <- str_split(y$genes$Protein.Ids, ";")
crap_only <- map_lgl(members, \(m) all(str_starts(m, "cRAP-")))
drop <- crap_only | str_starts(y$genes$Protein.Group, "rev_")
n_crap_only <- sum(drop)
pct_crap_only <- 100 * sum(2^y$E[drop, ], na.rm = TRUE) / signal_start

n_mentions_crap <- sum(str_detect(y$genes$Protein.Ids, "cRAP-"))
pct_mentions_crap <- 100 * sum(2^y$E[str_detect(y$genes$Protein.Ids, "cRAP-"), ],
  na.rm = TRUE
) / signal_start

y <- y[!drop, ]

strip_crap <- function(x) {
  str_split(str_remove_all(x, "cRAP-"), ";") |>
    map_chr(\(z) str_c(unique(z), collapse = ";"))
}
y$genes <- y$genes |>
  mutate(across(c(Protein.Group, Protein.Ids, Protein.Names, Genes), strip_crap),
    Proteotypic = as.integer(!str_detect(Protein.Ids, ";"))
  )

stopifnot(!any(str_detect(unlist(y$genes), "cRAP")))
```

Groups mentioning `cRAP` at all: `r format(n_mentions_crap, big.mark = ",")` precursors
carrying `r round(pct_mentions_crap, 2)`% of signal. Groups with no human accession, which
is what was removed: `r format(n_crap_only, big.mark = ",")` precursors carrying
`r round(pct_crap_only, 2)`%.

## limpa's precursor filters

Both run on the repaired annotation, in the vignette's order. `filterSingletonPeptides()`
is deliberately skipped: the vignette calls it optional and recommends keeping all proteins
to retain maximum information. Nothing here filters on missingness either, because the
vignette is explicit that precursors need no missing-value filtering, and those gaps are
exactly what `dpc()` recovers.

```{r limpa-filters}
tally <- function(y, step) {
  tibble(
    step = step,
    precursors = nrow(y),
    protein_groups = n_distinct(y$genes$Protein.Group),
    pct_signal = 100 * sum(2^y$E, na.rm = TRUE) / signal_start
  )
}
log_start <- tally(y, "0 repaired")

y <- filterNonProteotypicPeptides(y)
log_proteotypic <- tally(y, "1 shared peptides")

y <- filterCompoundProteins(y)
log_compound <- tally(y, "2 compound protein groups")
```

## Contaminants

A needle biopsy bleeds and passes through skin, so a large share of signal here is blood,
plasma, skin and antibody protein that muscle fibres do not express. It goes before
quantification because `dpc()` fits one curve across every precursor and `dpcQuant()`
estimates its prior from the global data, so contaminants left in place shape the
missing-value model that every muscle protein is then quantified through.
`filterCompoundProteins()` removes groups holding several *proteins*, but a single protein
can still be annotated with several *gene symbols*, so a plain symbol test is not enough. A
group counts as contamination only when every one of its symbols does, which is what keeps
myoglobin when DIA-NN reports it alongside a second name.

```{r contaminants}
members <- str_split(str_to_upper(y$genes$Genes), ";")
n_multi <- sum(lengths(members) > 1)

flagged <- function(g) g %in% CONTAMINANTS | str_detect(g, CONTAMINANT_FAMILIES)
contaminant <- vapply(members, \(g) length(g) > 0L && all(flagged(g)), logical(1))

removed <- tibble(
  symbol = y$genes$Genes[contaminant],
  precursors = 1L,
  signal = rowSums(2^y$E[contaminant, ], na.rm = TRUE)
) |>
  summarise(precursors = n(), pct_signal = 100 * sum(signal) / signal_start, .by = symbol) |>
  mutate(matched_by = if_else(symbol %in% CONTAMINANTS, "symbol", "family")) |>
  arrange(desc(pct_signal))

y <- y[!contaminant, ]
log_contaminants <- tally(y, "3 contaminants")

head(removed, 10)
```

`r format(n_multi, big.mark = ",")` of the surviving groups list more than one gene symbol,
which is why the per-member rule is needed. `r nrow(removed)` distinct symbol strings were
removed, carrying `r round(sum(removed$pct_signal), 2)`% of the starting signal.

::: {.callout-note}
## Blood index, not computed
Nothing here measures how much of each sample's signal was blood. Biopsy technique varies
between legs and between visits, and blood takes MS duty cycle away from muscle protein, so
the amount could differ between groups and confound the comparison. Someone should check
whether it does before write-up. The haemoglobin rows are still in the matrix at this point
in the notebook, which is where that measurement would have to be taken.
:::

## Guard

One assertion, run after every filter. It exists in order to fail: it catches a broken
repair, which deletes muscle through the `Proteotypic` flag, and an over-reaching
contaminant rule, which deletes it directly.

```{r guard}
survivors <- unlist(str_split(str_to_upper(y$genes$Genes), ";"))
stopifnot(all(MUST_KEEP %in% survivors))
```

## Outputs

`precursors_filtered.rds` is read by `01_Preprocess/02_Quantification`. Nothing reads the
two csv files; they are the record of what this stage did.

```{r write}
filter_log <- bind_rows(log_start, log_proteotypic, log_compound, log_contaminants)

saveRDS(y, here("01_Preprocess", "01_Filtering", "c_data", "precursors_filtered.rds"))
write_csv(filter_log, here("01_Preprocess", "01_Filtering", "c_data", "filter_log.csv"))
write_csv(removed, here("01_Preprocess", "01_Filtering", "c_data", "contaminants_removed.csv"))

filter_log
```

```{r session}
sessionInfo()
```
