---
title: "01 · Filtering"
subtitle: "DIA-NN report to a filtered precursor matrix"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    df-print: kable
    fig-width: 7
    fig-height: 5
    fig-dpi: 150
    fig-align: center
execute:
  message: false
---

`readDIANN()` reads the DIA-NN report, applies q-value cutoffs and returns log2 precursor
intensities. Every step after it removes rows and never columns: annotation artefacts of the
search FASTA, precursors shared between protein groups, compound groups, and proteins skeletal
muscle does not express. No sample is dropped and no missing value is filtered.

```{r setup}
pacman::p_load(here, limpa, dplyr, stringr, purrr, readr, writexl)

# Intensities arrive on the log2 scale, so every share-of-signal figure un-logs first.
total_signal <- function(m) sum(2^m, na.rm = TRUE)

step_summary <- function(elist, step, signal_start) {
  tibble(
    step = step,
    precursors = nrow(elist),
    protein_groups = n_distinct(elist$genes$Protein.Group),
    pct_signal = 100 * total_signal(elist$E) / signal_start
  )
}

# Anchored at both ends: a bare IGH/IGK/IGL/KRT/SPRR prefix also matches IGHMBP2, IGLON5 and
# KRTCAP2, which muscle expresses. Built from a vector so a lost "|" cannot widen it.
contaminant_families <- str_c(
  "^(",
  str_c(
    c(
      "IG[HKL][VDJCAGMEL]([0-9][0-9A-Z-]*)?",
      "KRT([0-9]|AP)[0-9A-Z-]*",
      "SPRR[0-9][0-9A-Z]*",
      "HB[ABDGMQZ][0-9]?"
    ),
    collapse = "|"
  ),
  ")$"
)

# Screened on HPA myonuclei single-cell expression rather than whole-muscle tissue, because
# tissue values are themselves blood-contaminated: haemoglobin reads 544 nTPM in skeletal muscle.
contaminant_symbols <- c(
  "JCHAIN", "HRNR", "DCD", "SBSN", "CSTA", "AHSP",
  "A1BG", "AFM", "AGT", "AHSG", "ALB", "AMBP", "APCS", "APOA1", "APOA2", "APOA4", "APOB",
  "APOC1", "APOC2", "APOC3", "APOD", "APOE", "APOF", "APOH", "APOL1", "APOL2", "APOM",
  "ARHGDIB", "AZGP1", "BCAM", "BTD", "C1QB", "C1QC", "C1R", "C1RL", "C1S", "C2", "C3",
  "C4A", "C4B", "C4BPA", "C4BPB", "C5", "C6", "C7", "C8A", "C8B", "C8G", "C9", "CA1",
  "CFB", "CFD", "CFH", "CFHR2", "CFHR5", "CFI", "CLEC3B", "CLU", "CP", "CPB2", "CPN1",
  "CPN2", "CTSG", "DEFA1", "DMTN", "ELANE", "EPB42", "F10", "F11", "F12", "F13A1", "F13B",
  "F2", "F9", "FCN3", "FERMT3", "FETUB", "FGA", "FGB", "FGG", "FYB1", "GC", "GIMAP1",
  "GIMAP4", "GMFG", "GP1BA", "GPLD1", "GYPC", "HABP2", "HP", "HPR", "HPX", "HRG",
  "IGFALS", "ITIH1", "ITIH3", "KEL", "KLKB1", "KNG1", "LBP", "LCAT", "LCP1", "LPA",
  "LRG1", "LTF", "LYZ", "MASP1", "MPO", "MPP1", "NCF1", "ORM1", "ORM2", "PF4", "PGLYRP2",
  "PLG", "PLTP", "PON1", "PPBP", "PROS1", "PRTN3", "PZP", "RBP4", "RHCE", "RNASE3",
  "S100A8", "S100A9", "SAA1", "SAA4", "SELENOP", "SERPINA1", "SERPINA3", "SERPINA4",
  "SERPINA5", "SERPINA6", "SERPINA7", "SERPINC1", "SERPIND1", "SERPINF2", "SERPING1",
  "SHBG", "SLC4A1", "SPTA1", "TF", "TTR", "TUBB1", "VTN", "VWF"
)

# Must survive every filter: muscle proteins the repair rescues, and blood-abundant proteins
# HPA myonuclei data show in muscle. ANK1 is small ankyrin-1, a sarcoplasmic reticulum protein.
must_keep <- c(
  "MB", "ACTA1", "ALDOA", "TPM2", "ENO1", "GPI", "CKM", "MYH1",
  "MYH2", "MYH7", "TTN", "DES", "PYGM", "CASQ1", "PVALB", "MYH13",
  "ANK1", "SPTB", "ITIH2", "ITIH4", "SERPINF1", "CAT", "CA2", "CA3",
  "A2M", "BLVRB", "STOM", "TALDO1"
)
```

## Read

`Protein.Ids` is not one of `readDIANN()`'s default `annotation.columns`, and the repair below
cannot run without it. It is added to the defaults, not substituted for them, so `Protein.Names`
still comes through for later annotation.

Everything else stays at the defaults. `q.cutoffs = 0.01` over `Q.Value`, `Lib.Q.Value` and
`Lib.PG.Q.Value` is the set the vignette recommends for a match-between-runs search.
`intensity.column = "Precursor.Normalised"` is DIA-NN's normalised column, which is why nothing
here normalises again. `log = TRUE` means `y$E` arrives on the log2 scale, so never log it twice.

```{r read}
y <- readDIANN(
  here("00_Input", "report.parquet"),
  annotation.columns = c(
    "Protein.Group", "Protein.Ids", "Protein.Names",
    "Genes", "Proteotypic"
  )
)
dim(y)
```

## Map runs to samples

Columns come back named by DIA-NN `Run`. Matching them case-insensitively against `metadata$run`
renames columns to `sample_id`, absorbs a run the instrument wrote `Cl_MDR_7E-T2` against
`CL_MDR_7E-T2` in the sheet, and drops the discarded S08 injection, which is in the report but
not the sheet.

```{r map-samples}
metadata <- read_csv(here("00_Input", "metadata.csv"), show_col_types = FALSE)
i <- match(str_to_lower(metadata$run), str_to_lower(colnames(y)))
# An unmatched run would index as NA and produce a silent all-NA sample column.
stopifnot(!anyNA(i))
runs_dropped <- setdiff(colnames(y), colnames(y)[i])
y <- y[, i]
colnames(y) <- metadata$sample_id
# A data.frame, not the tibble: limma's EList subsetting uses drop-style indexing on
# $targets, which a tibble does not honour.
y$targets <- data.frame(metadata, row.names = metadata$sample_id)

# The denominator for every percentage below. Captured after sample mapping, so dropped
# runs never count, and never recomputed, so the step percentages stay comparable.
signal_start <- total_signal(y$E)
c(runs_in_report = length(i) + length(runs_dropped), samples_kept = ncol(y))
runs_dropped
```

## Repair the annotation

The contaminant database appended to the search FASTA repeats proteins the human FASTA already
lists, so myoglobin is reported as `P02144;cRAP-P02144` with `Proteotypic = 0`. Dropping every
group that mentions `cRAP` would take CKM, MB, ALDOA, CAT, DES and FABP3 with it. Only the groups
where every member is cRAP go; the rest have the prefix stripped.

Order matters, because `Genes` carries the prefix too. One of the six cRAP-only symbols here is
`cRAP-MB`: strip first and it becomes `MB`, which `must_keep` requires to survive, so precursors
of a contaminant standard would merge into human myoglobin unnoticed. Two more cRAP-only rows
carry no gene symbol at all, so no symbol list could catch them either.

Stripping also repairs `Proteotypic`, which is why the muscle survives.
`filterNonProteotypicPeptides()` believes the flag, and on the unrepaired annotation it deleted 13
of the `must_keep` symbols when that was last measured.

```{r repair}
members <- str_split(y$genes$Protein.Ids, ";")
crap_only <- map_lgl(members, \(m) all(str_starts(m, "cRAP-")))
is_decoy <- str_starts(y$genes$Protein.Group, "rev_")
mentions_crap <- str_detect(y$genes$Protein.Ids, "cRAP-")
removed_rows <- crap_only | is_decoy

removal_rows <- list(
  "cRAP-only group" = crap_only, "reversed decoy" = is_decoy, "mentions cRAP at all" = mentions_crap
)
removal_counts <- tibble(
  reason = names(removal_rows),
  precursors = map_int(removal_rows, sum),
  pct_signal = 100 * map_dbl(removal_rows, \(r) total_signal(y$E[r, ])) / signal_start
)

y <- y[!removed_rows, ]

y$genes <- y$genes |>
  mutate(
    across(
      c(Protein.Group, Protein.Ids, Protein.Names, Genes),
      \(x) {
        map_chr(
          str_split(str_remove_all(x, "cRAP-"), ";"),
          \(ids) str_c(unique(ids), collapse = ";")
        )
      }
    ),
    # mutate() evaluates in order, so this reads the de-duplicated Protein.Ids above; splitting
    # the call breaks it. mentions_crap was taken before the prefix was stripped.
    Proteotypic = if_else(
      mentions_crap[!removed_rows],
      as.integer(!str_detect(Protein.Ids, ";")),
      as.integer(Proteotypic)
    )
  )

removal_counts
```

Two faults share this chunk. Most groups mentioning `cRAP` keep a human accession and are
repaired rather than removed, so only the cRAP-only ones go. Reversed decoys are separate: the
search FASTA already contained them, so DIA-NN scored them as targets and its own decoy column
reads zero for every row. They carry `r signif(removal_counts$pct_signal[2], 2)`% of the starting
signal, but they inflate the protein count and would surface as nonsense identifiers in any
enrichment.

## limpa's precursor filters

Both run on the repaired annotation, in the vignette's order. `filterSingletonPeptides()` is
skipped: the vignette calls it optional and recommends keeping all proteins to retain maximum
information. Nothing here filters on missingness either, because the vignette says precursors
need no missing-value filtering, and those gaps are what `dpc()` recovers.

```{r limpa-filters}
summary_repaired <- step_summary(y, "repaired", signal_start)

y <- filterNonProteotypicPeptides(y)
summary_proteotypic <- step_summary(y, "shared peptides removed", signal_start)

y <- filterCompoundProteins(y)
summary_compound <- step_summary(y, "compound groups removed", signal_start)
```

## Contaminants

A needle biopsy bleeds and passes through skin, so a large share of signal here is blood, plasma,
skin and antibody protein that muscle fibres do not express. Removal happens before quantification
because `dpc()` fits one curve across every precursor and `dpcQuant()` estimates its prior from
the global data, so contaminants left in place would shape the missing-value model that every
muscle protein is then quantified through.

`contaminant_symbols` holds 147 gene symbols and `contaminant_families` four family patterns.
`filterCompoundProteins()` removes groups holding several proteins, but one protein can still
carry several gene symbols, so a test on the first symbol fails both ways. A group counts as
contamination only when every one of its symbols does, which keeps myoglobin when DIA-NN reports
it alongside a second name.

```{r contaminants}
gene_symbols <- str_split(str_to_upper(y$genes$Genes), ";")
n_multi <- sum(lengths(gene_symbols) > 1)

by_symbol <- map_lgl(gene_symbols, \(g) all(g %in% contaminant_symbols))
contaminant <- map_lgl(gene_symbols, \(g) {
  isTRUE(all(g %in% contaminant_symbols | str_detect(g, contaminant_families)))
})

removed <- tibble(
  symbol = y$genes$Genes[contaminant],
  matched_by = if_else(by_symbol[contaminant], "symbol", "family"),
  signal = rowSums(2^y$E[contaminant, , drop = FALSE], na.rm = TRUE)
) |>
  summarise(
    precursors = n(), pct_signal = 100 * sum(signal) / signal_start,
    .by = c(symbol, matched_by)
  ) |>
  arrange(desc(pct_signal))

# Uneven contamination across the four treatment-by-timepoint cells would land on a contrast.
contaminant_fraction <- colSums(2^y$E[contaminant, , drop = FALSE], na.rm = TRUE) /
  colSums(2^y$E, na.rm = TRUE)
round(tapply(contaminant_fraction, y$targets$group, median), 4)

y <- y[!contaminant, ]
summary_contaminants <- step_summary(y, "contaminants removed", signal_start)

head(removed, 20)
```

`r format(n_multi, big.mark = ",")` of the groups tested list more than one gene symbol, which is
why the per-member rule is needed. `r nrow(removed)` distinct symbol strings were removed,
carrying `r round(sum(removed$pct_signal), 2)`% of the starting signal.

::: callout-note
## What the contaminant fraction does and does not settle

The medians above compare the four cells and show a gross imbalance in how bloody the biopsies
were. They do not show a small one. Nothing adjusts for the fraction and no test is run on it.
Blood takes MS duty cycle away from muscle protein, and biopsy technique varies between legs and
between visits. `02_Differential` estimates one quality weight per sample; the weight absorbs
part of this and is not aimed at it.
:::

## Guard

All 28 symbols in `must_keep` must survive. The assertion runs after the last filter. It catches a
broken repair, which deletes muscle through the `Proteotypic` flag, and an over-reaching
contaminant rule, which deletes it directly.

```{r guard}
survivors <- unlist(str_split(str_to_upper(y$genes$Genes), ";"))
stopifnot(all(must_keep %in% survivors))
```

## Outputs

`precursors_filtered.rds` is read by `01_Preprocess/02_Quantification`. `01_filter.xlsx` is the
record of what this stage did; its first sheet lists the others.

```{r write}
filter_log <- bind_rows(
  summary_repaired, summary_proteotypic, summary_compound, summary_contaminants
)
sheets <- list(filter_log = filter_log, contaminants_removed = removed)
overview <- data.frame(
  sheet = names(sheets),
  rows = sapply(sheets, nrow),
  columns = sapply(sheets, ncol),
  description = c(
    "Rows, groups and signal after each filter",
    "Removed contaminants and their signal"
  )
)

out <- here("01_Preprocess", "01_Filtering", "c_data")
saveRDS(y, file.path(out, "precursors_filtered.rds"))
write_xlsx(c(list(overview = overview), sheets), file.path(out, "01_filter.xlsx"))

filter_log
```

```{r session}
sessionInfo()
```
