---
title: "01 · Filtering"
subtitle: "DIA-NN report to a filtered precursor matrix"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    df-print: kable
    fig-width: 7
    fig-height: 5
    fig-dpi: 150
    fig-align: center
execute:
  message: false
---

`readDIANN()` reads the DIA-NN report, applies q-value cutoffs and returns log2 precursor
intensities. Every later step removes rows, never columns: annotation artefacts of the search FASTA,
precursors shared between protein groups, compound groups, and proteins skeletal muscle does not
express. Nothing is filtered on missing values.

```{r setup}
pacman::p_load(here, limpa, dplyr, stringr, purrr, readr, writexl)

# Intensities arrive on the log2 scale, so every share-of-signal figure un-logs first.
total_signal <- function(m) sum(2^m, na.rm = TRUE)

step_summary <- function(elist, step) {
  tibble(
    step = step,
    precursors = nrow(elist),
    protein_groups = n_distinct(elist$genes$Protein.Group),
    pct_signal = 100 * total_signal(elist$E) / signal_start
  )
}

# Anchored at both ends: a bare IGH/IGK/IGL/KRT/SPRR prefix also matches IGHMBP2, IGLON5 and
# KRTCAP2, which muscle expresses. Built from a vector so a lost "|" cannot widen it.
contaminant_families <- str_c("^(", str_flatten(c(
  "IG[HKL][VDJCAGMEL]([0-9][0-9A-Z-]*)?", "KRT([0-9]|AP)[0-9A-Z-]*",
  "SPRR[0-9][0-9A-Z]*", "HB[ABDGMQZ][0-9]?"
), "|"), ")$")

# Screened on HPA myonuclei single-cell expression rather than whole-muscle tissue, because
# tissue values are themselves blood-contaminated: haemoglobin reads 544 nTPM in skeletal muscle.
contaminant_symbols <- c(
  "JCHAIN", "HRNR", "DCD", "SBSN", "CSTA", "AHSP",
  "A1BG", "AFM", "AGT", "AHSG", "ALB", "AMBP", "APCS", "APOA1", "APOA2", "APOA4", "APOB",
  "APOC1", "APOC2", "APOC3", "APOD", "APOE", "APOF", "APOH", "APOL1", "APOL2", "APOM",
  "ARHGDIB", "AZGP1", "BCAM", "BTD", "C1QB", "C1QC", "C1R", "C1RL", "C1S", "C2", "C3",
  "C4A", "C4B", "C4BPA", "C4BPB", "C5", "C6", "C7", "C8A", "C8B", "C8G", "C9", "CA1",
  "CFB", "CFD", "CFH", "CFHR2", "CFHR5", "CFI", "CLEC3B", "CLU", "CP", "CPB2", "CPN1",
  "CPN2", "CTSG", "DEFA1", "DMTN", "ELANE", "EPB42", "F10", "F11", "F12", "F13A1", "F13B",
  "F2", "F9", "FCN3", "FERMT3", "FETUB", "FGA", "FGB", "FGG", "FYB1", "GC", "GIMAP1",
  "GIMAP4", "GMFG", "GP1BA", "GPLD1", "GYPC", "HABP2", "HP", "HPR", "HPX", "HRG",
  "IGFALS", "ITIH1", "ITIH3", "KEL", "KLKB1", "KNG1", "LBP", "LCAT", "LCP1", "LPA",
  "LRG1", "LTF", "LYZ", "MASP1", "MPO", "MPP1", "NCF1", "ORM1", "ORM2", "PF4", "PGLYRP2",
  "PLG", "PLTP", "PON1", "PPBP", "PROS1", "PRTN3", "PZP", "RBP4", "RHCE", "RNASE3",
  "S100A8", "S100A9", "SAA1", "SAA4", "SELENOP", "SERPINA1", "SERPINA3", "SERPINA4",
  "SERPINA5", "SERPINA6", "SERPINA7", "SERPINC1", "SERPIND1", "SERPINF2", "SERPING1",
  "SHBG", "SLC4A1", "SPTA1", "TF", "TTR", "TUBB1", "VTN", "VWF"
)

# Must survive every filter: muscle proteins the repair rescues, and blood-abundant proteins
# HPA myonuclei data show in muscle. ANK1 is small ankyrin-1, a sarcoplasmic reticulum protein.
must_keep <- c(
  "MB", "ACTA1", "ALDOA", "TPM2", "ENO1", "GPI", "CKM", "MYH1",
  "MYH2", "MYH7", "TTN", "DES", "PYGM", "CASQ1", "PVALB", "MYH13",
  "ANK1", "SPTB", "ITIH2", "ITIH4", "SERPINF1", "CAT", "CA2", "CA3",
  "A2M", "BLVRB", "STOM", "TALDO1"
)
```

## Read

The repair below needs `Protein.Ids`, which `readDIANN()` does not read by default, so it is added
to the default `annotation.columns` and `Protein.Names` still comes through.

Everything else is default: `q.cutoffs = 0.01` over `Q.Value`, `Lib.Q.Value` and `Lib.PG.Q.Value`,
as the vignette recommends for a match-between-runs search;
`intensity.column = "Precursor.Normalised"`, already normalised by DIA-NN, so this stage does not
normalise again; and `log = TRUE`, so `y$E` is already log2.

```{r read}
y <- readDIANN(
  here("00_Input", "report.parquet"),
  annotation.columns = c(
    "Protein.Group", "Protein.Ids", "Protein.Names",
    "Genes", "Proteotypic"
  )
)
dim(y)
```

## Map runs to samples

Columns arrive named by DIA-NN `Run`. A case-insensitive match against `metadata$run` renames them
to `sample_id`, absorbs `Cl_MDR_7E-T2` (written `CL_MDR_7E-T2` in the sheet), and drops the
discarded S08 injection, which is in the report but not the sheet.

```{r map-samples}
metadata <- read_csv(here("00_Input", "metadata.csv"), show_col_types = FALSE)
i <- match(str_to_lower(metadata$run), str_to_lower(colnames(y)))
# An unmatched run would index as NA and produce a silent all-NA sample column.
stopifnot(!anyNA(i))
setdiff(colnames(y), colnames(y)[i])
y <- y[, i]
colnames(y) <- metadata$sample_id
# A data.frame, not the tibble: limma's EList subsetting uses drop-style indexing on
# $targets, which a tibble does not honour.
y$targets <- data.frame(metadata, row.names = metadata$sample_id)

# The denominator for every percentage below. Captured after sample mapping, so dropped
# runs never count, and never recomputed, so the step percentages stay comparable.
signal_start <- total_signal(y$E)
```

## Repair the annotation

The contaminant database appended to the search FASTA repeats proteins the human FASTA already
lists, so myoglobin is reported as `P02144;cRAP-P02144` with `Proteotypic = 0`. Dropping every group
that mentions `cRAP` would take CKM, MB, ALDOA, CAT, DES and FABP3 with it, so only groups where
every member is cRAP go, and the rest have the prefix stripped.

Order matters because `Genes` carries the prefix too. Stripping first would turn `cRAP-MB`, one of
the six cRAP-only symbols here, into `MB`, which `must_keep` protects, and a contaminant standard
would merge into human myoglobin unnoticed. Two more cRAP-only rows carry no gene symbol, so no
symbol list could catch them.

Stripping also repairs `Proteotypic`. `filterNonProteotypicPeptides()` trusts that flag, and on the
unrepaired annotation it deleted 13 `must_keep` symbols when last measured.

```{r repair}
members <- str_split(y$genes$Protein.Ids, ";")
crap_only <- map_lgl(members, \(m) all(str_starts(m, "cRAP-")))
is_decoy <- str_starts(y$genes$Protein.Group, "rev_")
mentions_crap <- str_detect(y$genes$Protein.Ids, "cRAP-")
removed_rows <- crap_only | is_decoy

reasons <- list(
  "cRAP-only group" = crap_only, "reversed decoy" = is_decoy, "mentions cRAP at all" = mentions_crap
)
removal_counts <- tibble(
  reason = names(reasons),
  precursors = map_int(reasons, sum),
  pct_signal = 100 * map_dbl(reasons, \(r) total_signal(y$E[r, ])) / signal_start
)

y <- y[!removed_rows, ]

y$genes <- y$genes |>
  mutate(
    across(
      c(Protein.Group, Protein.Ids, Protein.Names, Genes),
      \(x) str_split(str_remove_all(x, "cRAP-"), ";") |> map_chr(\(m) str_flatten(unique(m), ";"))
    ),
    # mutate() evaluates in order, so this reads the de-duplicated Protein.Ids above; splitting
    # the call breaks it. mentions_crap was taken before the prefix was stripped.
    Proteotypic = if_else(
      mentions_crap[!removed_rows],
      as.integer(!str_detect(Protein.Ids, ";")),
      as.integer(Proteotypic)
    )
  )

removal_counts
```

The chunk also drops reversed decoys. The search FASTA already contained them, so DIA-NN scored them
as targets and its decoy column reads zero throughout. They carry
`r signif(removal_counts$pct_signal[removal_counts$reason == "reversed decoy"], 2)`% of the starting signal, yet inflate the protein count
and would surface as nonsense identifiers in enrichment.

## limpa's precursor filters

Both run on the repaired annotation, in the vignette's order. `filterSingletonPeptides()` is
skipped, since the vignette calls it optional and recommends keeping all proteins. Nothing is
filtered on missingness: the vignette says precursors need none, and `dpc()` models those gaps.

```{r limpa-filters}
filter_log <- step_summary(y, "repaired")
y <- filterNonProteotypicPeptides(y)
filter_log <- bind_rows(filter_log, step_summary(y, "shared peptides removed"))
y <- filterCompoundProteins(y)
filter_log <- bind_rows(filter_log, step_summary(y, "compound groups removed"))
```

## Contaminants

A needle biopsy bleeds and passes through skin, so much of the signal is blood, plasma, skin and
antibody protein that muscle fibres do not express. It is removed before quantification because
`dpc()` fits one curve across every precursor and `dpcQuant()` takes its prior from the global data,
so contaminants would shape the model every muscle protein is quantified through.

`contaminant_symbols` holds 147 gene symbols and `contaminant_families` four family patterns. One
protein can carry several gene symbols even after `filterCompoundProteins()`, so testing the first
symbol fails both ways. A group is contamination only when every symbol on it is, which keeps
myoglobin when DIA-NN lists a second name.

```{r contaminants}
gene_symbols <- str_split(str_to_upper(y$genes$Genes), ";")
n_multi <- sum(lengths(gene_symbols) > 1)

by_symbol <- map_lgl(gene_symbols, \(g) all(g %in% contaminant_symbols))
contaminant <- map_lgl(gene_symbols, \(g) {
  isTRUE(all(g %in% contaminant_symbols | str_detect(g, contaminant_families)))
})

contaminants_removed <- tibble(
  symbol = y$genes$Genes[contaminant],
  matched_by = if_else(by_symbol[contaminant], "symbol", "family"),
  signal = rowSums(2^y$E[contaminant, ], na.rm = TRUE)
) |>
  summarise(
    precursors = n(), pct_signal = 100 * sum(signal) / signal_start,
    .by = c(symbol, matched_by)
  ) |>
  arrange(desc(pct_signal))

# Uneven contamination across the four treatment-by-timepoint cells would land on a contrast.
contaminant_fraction <- colSums(2^y$E[contaminant, ], na.rm = TRUE) /
  colSums(2^y$E, na.rm = TRUE)
round(tapply(contaminant_fraction, y$targets$group, median), 4)

y <- y[!contaminant, ]
filter_log <- bind_rows(filter_log, step_summary(y, "contaminants removed"))

head(contaminants_removed, 20)
```

`r format(n_multi, big.mark = ",")` tested groups list more than one symbol. `r nrow(contaminants_removed)`
distinct symbol strings were removed, carrying `r round(sum(contaminants_removed$pct_signal), 2)`% of the
starting signal.

::: callout-note
## What the contaminant fraction does and does not settle

The medians above can reveal a gross imbalance in blood content between the four cells, not a small
one. Nothing adjusts for the fraction or tests it. Blood takes MS duty cycle from muscle protein,
and biopsy technique varies between legs and visits. The per-sample quality weight in
`02_Differential` absorbs part of this without targeting it.
:::

## Guard

After the last filter, all 28 `must_keep` symbols must survive. That catches a broken repair, which
deletes muscle through `Proteotypic`, and an over-reaching contaminant rule, which deletes it
directly.

```{r guard}
survivors <- unlist(str_split(str_to_upper(y$genes$Genes), ";"))
stopifnot(all(must_keep %in% survivors))
```

## Outputs

`01_Preprocess/02_Quantification` reads `precursors_filtered.rds`. `01_filter.xlsx` records what
this stage did, with an overview as its first sheet.

```{r write}
sheets <- list(filter_log = filter_log, contaminants_removed = contaminants_removed)
overview <- data.frame(
  sheet = names(sheets), rows = map_int(sheets, nrow), columns = map_int(sheets, ncol),
  description = c(
    "Rows, groups and signal after each filter",
    "Removed contaminants and their signal"
  )
)

out <- here("01_Preprocess", "01_Filtering", "c_data")
saveRDS(y, file.path(out, "precursors_filtered.rds"))
write_xlsx(c(list(overview = overview), sheets), file.path(out, "01_filter.xlsx"))

filter_log
```

```{r session}
sessionInfo()
```
