---
title: "01 · Filtering"
subtitle: "DIA-NN report to a filtered precursor matrix"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    df-print: kable
execute:
  message: false
---

`readDIANN()` reads the DIA-NN report, applies q-value cutoffs and returns log2 precursor
intensities. Every step after it removes rows and never columns: annotation artefacts of
the search FASTA, precursors shared between protein groups, compound groups, and proteins
skeletal muscle does not express. No sample is dropped and no missing value is filtered.

```{r setup}
library(here)
library(limpa)
library(dplyr)
library(stringr)
library(purrr)
library(readr)
library(tibble)

# Intensities arrive on the log2 scale, so every share-of-signal figure un-logs first.
total_signal <- function(m) sum(2^m, na.rm = TRUE)

step_summary <- function(elist, step, signal_start) {
  tibble(
    step = step,
    precursors = nrow(elist),
    protein_groups = n_distinct(elist$genes$Protein.Group),
    pct_signal = 100 * total_signal(elist$E) / signal_start
  )
}

# Four gene families, each anchored at both ends. Anchoring matters: an unanchored
# IGH/IGK/IGL/KRT/SPRR prefix also matches IGHMBP2, IGLON5 and KRTCAP2, which muscle
# expresses. Built from a vector so a dropped "|" at a line wrap cannot silently
# change the pattern.
CONTAMINANT_FAMILIES <- str_c(
  "^(",
  str_c(
    c(
      "IG[HKL][VDJCAGMEL]([0-9][0-9A-Z-]*)?",
      "KRT([0-9]|AP)[0-9A-Z-]*",
      "SPRR[0-9][0-9A-Z]*",
      "HB[ABDGMQZ][0-9]?"
    ),
    collapse = "|"
  ),
  ")$"
)

# Screened on HPA myonuclei single-cell expression rather than whole-muscle tissue,
# because tissue values are themselves blood-contaminated: haemoglobin reads 544 nTPM in
# skeletal muscle. See 01_Preprocess/01_Filtering/README.md for the screening and the
# alternatives rejected.
CONTAMINANTS <- c(
  "JCHAIN", "HRNR", "DCD", "SBSN", "CSTA", "AHSP",
  "A1BG", "AFM", "AGT", "AHSG", "ALB", "AMBP", "APCS", "APOA1", "APOA2", "APOA4", "APOB",
  "APOC1", "APOC2", "APOC3", "APOD", "APOE", "APOF", "APOH", "APOL1", "APOL2", "APOM",
  "ARHGDIB", "AZGP1", "BCAM", "BTD", "C1QB", "C1QC", "C1R", "C1RL", "C1S", "C2", "C3",
  "C4A", "C4B", "C4BPA", "C4BPB", "C5", "C6", "C7", "C8A", "C8B", "C8G", "C9", "CA1",
  "CFB", "CFD", "CFH", "CFHR2", "CFHR5", "CFI", "CLEC3B", "CLU", "CP", "CPB2", "CPN1",
  "CPN2", "CTSG", "DEFA1", "DMTN", "ELANE", "EPB42", "F10", "F11", "F12", "F13A1", "F13B",
  "F2", "F9", "FCN3", "FERMT3", "FETUB", "FGA", "FGB", "FGG", "FYB1", "GC", "GIMAP1",
  "GIMAP4", "GMFG", "GP1BA", "GPLD1", "GYPC", "HABP2", "HP", "HPR", "HPX", "HRG",
  "IGFALS", "ITIH1", "ITIH3", "KEL", "KLKB1", "KNG1", "LBP", "LCAT", "LCP1", "LPA",
  "LRG1", "LTF", "LYZ", "MASP1", "MPO", "MPP1", "NCF1", "ORM1", "ORM2", "PF4", "PGLYRP2",
  "PLG", "PLTP", "PON1", "PPBP", "PROS1", "PRTN3", "PZP", "RBP4", "RHCE", "RNASE3",
  "S100A8", "S100A9", "SAA1", "SAA4", "SELENOP", "SERPINA1", "SERPINA3", "SERPINA4",
  "SERPINA5", "SERPINA6", "SERPINA7", "SERPINC1", "SERPIND1", "SERPINF2", "SERPING1",
  "SHBG", "SLC4A1", "SPTA1", "TF", "TTR", "TUBB1", "VTN", "VWF"
)

# Asserted to survive every filter below. The first two rows are muscle proteins the
# annotation repair exists to rescue. The rest are blood-abundant but muscle-expressed on
# HPA myonuclei data: ANK1 is small ankyrin-1, a sarcoplasmic reticulum protein, and the
# review did flag A2M, BLVRB, STOM and TALDO1, which carry 0.004% of signal between them.
MUST_KEEP <- c(
  "MB", "ACTA1", "ALDOA", "TPM2", "ENO1", "GPI", "CKM", "MYH1",
  "MYH2", "MYH7", "TTN", "DES", "PYGM", "CASQ1", "PVALB", "MYH13",
  "ANK1", "SPTB", "ITIH2", "ITIH4", "SERPINF1", "CAT", "CA2", "CA3",
  "A2M", "BLVRB", "STOM", "TALDO1"
)
```

## Read

`Protein.Ids` is not one of `readDIANN()`'s default `annotation.columns`, and the repair
below cannot run without it. It is added to the defaults, not substituted for them, so
`Protein.Names` still comes through for later annotation.

Everything else stays as shipped. `q.cutoffs = 0.01` over `Q.Value`, `Lib.Q.Value` and
`Lib.PG.Q.Value` is the set the vignette recommends for a match-between-runs search.
`intensity.column = "Precursor.Normalised"` is DIA-NN's normalized column, which is why
nothing here normalizes again. `log = TRUE` means `y$E` arrives on the log2 scale, so never
log it a second time.

```{r read}
#| cache: true
y <- readDIANN(
  here("00_Input", "report.parquet"),
  annotation.columns = c(
    "Protein.Group", "Protein.Ids", "Protein.Names",
    "Genes", "Proteotypic"
  )
)
dim(y)
```

## Map runs to samples

Columns come back named by DIA-NN `Run`. Matching them case-insensitively against
`metadata$run` does three jobs at once: it renames columns to `sample_id`, it absorbs a run
the instrument wrote `Cl_MDR_7E-T2` against `CL_MDR_7E-T2` in the sheet, and it drops the
discarded S08 injection, which is in the report but not the sheet.

```{r map-samples}
metadata <- read_csv(here("00_Input", "metadata.csv"), show_col_types = FALSE)
i <- match(str_to_lower(metadata$run), str_to_lower(colnames(y)))
runs_dropped <- setdiff(colnames(y), colnames(y)[i])
y <- y[, i]
colnames(y) <- metadata$sample_id
# as.data.frame, not the tibble: limma's EList subsetting uses drop-style indexing on
# $targets, which a tibble does not honour.
y$targets <- as.data.frame(metadata)
rownames(y$targets) <- metadata$sample_id

# The denominator for every percentage below. Captured after sample mapping, so dropped
# runs never count, and never recomputed, so the step percentages stay comparable.
signal_start <- total_signal(y$E)
c(runs_in_report = length(i) + length(runs_dropped), samples_kept = ncol(y))
runs_dropped
```

## Repair the annotation

The search FASTA held every protein twice, plain and `cRAP-`-prefixed, so myoglobin is
reported as `P02144;cRAP-P02144` with `Proteotypic = 0`. Dropping every group that mentions
`cRAP` would take CKM, MB, ALDOA, CAT, DES and FABP3 with it. What is wanted is to drop only
the groups where *every* member is cRAP, then strip the prefix from the rest.

The order matters, because `Genes` carries the prefix too. Six cRAP-only symbols appear
here, and one of them is `cRAP-MB`. Strip the prefix first and it becomes `MB`, which
`MUST_KEEP` requires to survive, so ten precursors of a contaminant standard would merge
into human myoglobin and nobody would see it. `cRAP-CSN1S1` is not in `CONTAMINANTS`
either, and two cRAP-only rows carry no gene symbol at all, so they match nothing and would
survive as a group standing for no real protein. The contaminant list cannot do this job.

Stripping also repairs `Proteotypic`, which is the whole reason the muscle survives:
`filterNonProteotypicPeptides()` believes the flag, and on the unrepaired annotation it
deletes 13 of the 28 `MUST_KEEP` symbols.

```{r repair}
members <- str_split(y$genes$Protein.Ids, ";")
crap_only <- map_lgl(members, \(m) all(str_starts(m, "cRAP-")))
is_decoy <- str_starts(y$genes$Protein.Group, "rev_")
mentions_crap <- str_detect(y$genes$Protein.Ids, "cRAP-")
drop <- crap_only | is_decoy

removal_counts <- tibble(
  reason = c("cRAP-only group", "reversed decoy", "mentions cRAP at all"),
  precursors = c(sum(crap_only), sum(is_decoy), sum(mentions_crap)),
  pct_signal = 100 * c(
    total_signal(y$E[crap_only, ]),
    total_signal(y$E[is_decoy, ]),
    total_signal(y$E[mentions_crap, ])
  ) / signal_start
)

y <- y[!drop, ]

# Captured before the across() below overwrites Protein.Ids. Once the prefix is gone
# there is no way to tell which rows the repair touched.
crap_mention <- str_detect(y$genes$Protein.Ids, "cRAP-")

y$genes <- y$genes |>
  mutate(
    across(
      c(Protein.Group, Protein.Ids, Protein.Names, Genes),
      \(x) map_chr(str_split(str_remove_all(x, "cRAP-"), ";"),
                   \(ids) str_c(unique(ids), collapse = ";"))
    ),
    # mutate() evaluates in order, so this reads the de-duplicated Protein.Ids the
    # line above just wrote. Splitting these into two mutate() calls breaks it.
    Proteotypic = if_else(
      crap_mention,
      as.integer(!str_detect(Protein.Ids, ";")),
      as.integer(Proteotypic)
    )
  )
```

```{r removal-counts}
removal_counts
```

Two separate problems share this chunk. Most groups mentioning `cRAP` keep a human accession
and are repaired rather than removed. Only the cRAP-only ones go. Reversed decoy sequences
are a different fault: the search FASTA already contained them, so DIA-NN scored them as
targets and its own decoy column reads zero for every row. They carry almost no signal, but
they inflate the protein count and would appear as nonsense identifiers in any enrichment.

## limpa's precursor filters

Both run on the repaired annotation, in the vignette's order. `filterSingletonPeptides()`
is deliberately skipped: the vignette calls it optional and recommends keeping all proteins
to retain maximum information. Nothing here filters on missingness either, because the
vignette is explicit that precursors need no missing-value filtering, and those gaps are
exactly what `dpc()` recovers.

```{r limpa-filters}
summary_repaired <- step_summary(y, "repaired", signal_start)

y <- filterNonProteotypicPeptides(y)
summary_proteotypic <- step_summary(y, "shared peptides removed", signal_start)

y <- filterCompoundProteins(y)
summary_compound <- step_summary(y, "compound groups removed", signal_start)
```

## Contaminants

A needle biopsy bleeds and passes through skin, so a large share of signal here is blood,
plasma, skin and antibody protein that muscle fibres do not express. It goes before
quantification because `dpc()` fits one curve across every precursor and `dpcQuant()`
estimates its prior from the global data, so contaminants left in place shape the
missing-value model that every muscle protein is then quantified through.
`filterCompoundProteins()` removes groups holding several *proteins*, but a single protein
can still be annotated with several *gene symbols*, so a plain symbol test is not enough. A
group counts as contamination only when every one of its symbols does, which is what keeps
myoglobin when DIA-NN reports it alongside a second name.

```{r contaminants}
members <- str_split(str_to_upper(y$genes$Genes), ";")
n_multi <- sum(lengths(members) > 1)

contaminant <- map_lgl(members, \(g) {
  isTRUE(all(g %in% CONTAMINANTS | str_detect(g, CONTAMINANT_FAMILIES)))
})

removed <- tibble(
  symbol = y$genes$Genes[contaminant],
  signal = rowSums(2^y$E[contaminant, , drop = FALSE], na.rm = TRUE)
) |>
  summarise(precursors = n(), pct_signal = 100 * sum(signal) / signal_start, .by = symbol) |>
  mutate(
    matched_by = if_else(
      map_lgl(str_split(str_to_upper(symbol), ";"), \(g) all(g %in% CONTAMINANTS)),
      "symbol", "family"
    )
  ) |>
  arrange(desc(pct_signal))

# Uneven contamination between T1 and T2 would be a systematic shift landing on the two time
# contrasts, not noise. The per-sample medians are checked in 02_Quantification, on the protein
# matrix, which is where limpa places any normalisation.
contaminant_fraction <- colSums(2^y$E[contaminant, ], na.rm = TRUE) /
  colSums(2^y$E, na.rm = TRUE)
tapply(contaminant_fraction, y$targets$timepoint, median)

y <- y[!contaminant, ]
summary_contaminants <- step_summary(y, "contaminants removed", signal_start)

head(removed, 20)
```


`r format(n_multi, big.mark = ",")` of the surviving groups list more than one gene symbol,
which is why the per-member rule is needed. `r nrow(removed)` distinct symbol strings were
removed, carrying `r round(sum(removed$pct_signal), 2)`% of the starting signal.

::: {.callout-note}
## Blood index, not computed
Nothing here measures how much of each sample's signal was blood. Biopsy technique varies
between legs and between visits, and blood takes MS duty cycle away from muscle protein, so
the amount could differ between groups and confound the comparison. Someone should check
whether it does before write-up. The haemoglobin rows are still in the matrix at this point
in the notebook, which is where that measurement would have to be taken.
:::

## Guard

One assertion, run after every filter. It exists in order to fail: it catches a broken
repair, which deletes muscle through the `Proteotypic` flag, and an over-reaching
contaminant rule, which deletes it directly.

```{r guard}
survivors <- unlist(str_split(str_to_upper(y$genes$Genes), ";"))
stopifnot(all(MUST_KEEP %in% survivors))

removed |>
  summarise(groups = n(), pct_signal = round(sum(pct_signal), 1))
```

## Outputs

`precursors_filtered.rds` is read by `01_Preprocess/02_Quantification`. Nothing reads the
two csv files; they are the record of what this stage did.

```{r write}
filter_log <- bind_rows(
  summary_repaired, summary_proteotypic, summary_compound, summary_contaminants
)

saveRDS(y, here("01_Preprocess", "01_Filtering", "c_data", "precursors_filtered.rds"))
write_csv(filter_log, here("01_Preprocess", "01_Filtering", "c_data", "filter_log.csv"))
write_csv(removed, here("01_Preprocess", "01_Filtering", "c_data", "contaminants_removed.csv"))

filter_log
```

```{r session}
sessionInfo()
```
