---
title: "02 · Differential expression"
subtitle: "Five contrasts on the protein matrix"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    df-print: kable
execute:
  message: false
---

limpa's sequence for a contrast-based analysis is `dpcDE`, `contrasts.fit`, `eBayes`, `topTable`.
`dpcDE()` is the only entry point that reads the standard errors `dpcQuant()` produced, so it
replaces `lmFit()` and nothing else may stand in for it.

```{r setup}
library(here)
library(limpa)
library(limma)
library(dplyr)
library(purrr)
library(readr)
library(tibble)

proteins <- readRDS(here("01_Preprocess", "02_Quantification", "c_data", "proteins.rds"))
d <- readRDS(here("02_Differential_Expression", "01_Design", "c_data", "design.rds"))

stopifnot(identical(rownames(d$design), colnames(proteins$E)))
```

The assertion checks the design rows are the same samples in the same order as the matrix columns.
limma compares only their counts.

`proteins.rds` arrives normalised; `01_Preprocess/02_Quantification` applies cyclic loess and
reports what it corrected.

## Fit

`sample.weights = TRUE` estimates one quality weight per sample. Biopsies differed in how much
blood they carried, and removing the contaminant rows does not undo what that did to the sample
underneath.

```{r fit}
#| fig-height: 4
fit <- dpcDE(proteins, d$design, sample.weights = TRUE, plot = TRUE)
fit <- contrasts.fit(fit, d$contrasts)
fit <- eBayes(fit, robust = TRUE)
```

Effect size the interaction could detect at 80% power, two-sided 0.05. Without this, "null" says
nothing.

```{r detectable-effect}
se <- fit$stdev.unscaled[, "interaction"] * sqrt(fit$s2.post)
round(quantile(se * (qt(0.975, fit$df.total) + qt(0.80, fit$df.total)), c(0.5, 0.9)), 3)
```

## Results

Benjamini-Hochberg runs within each contrast and never across them. The five share participants
and the interaction is a difference of two of the others, so pooling them would not control
anything that can be stated.

```{r results}
results <- map(colnames(d$contrasts), \(cn) {
  topTable(fit, coef = cn, number = Inf, adjust.method = "BH", sort.by = "p") |>
    rownames_to_column("protein") |>
    mutate(contrast = cn, .before = 1)
}) |>
  list_rbind() |>
  mutate(pi_score = P.Value^abs(logFC))

contrast_summary <- results |>
  summarise(n_tested = n(),
            n_fdr_05 = sum(adj.P.Val < 0.05),
            n_fdr_10 = sum(adj.P.Val < 0.10),
            min_adj_p = min(adj.P.Val),
            .by = contrast)
contrast_summary
```

`sort.by = "p"` is passed because `topTable()` defaults to sorting on the B statistic, which
orders differently.

## Is a contrast empty, or is it being tested conservatively

Three numbers together, because each misleads alone. `propTrueNull()` estimates the null
proportion from the p-value distribution. Median |t| is 0.674 under a calibrated null, so a value
below that means the tests are conservative. `p_below_0.2` is 0.2 under a uniform null, so a value
below that means the distribution rises toward 1 rather than being flat.

Read the between-leg contrasts against `BFR_vs_HLRT_at_T1`, not against theory. That contrast is a
true null by construction, and whatever it returns is what a null looks like in this design.

```{r null-calibration}
results |>
  summarise(prop_true_null = round(propTrueNull(P.Value), 3),
            median_abs_t = round(median(abs(t)), 3),
            p_below_0.2 = round(mean(P.Value < 0.2), 3), .by = contrast)
```


## Significant and large enough

`treat()` tests against a fold-change floor rather than against zero, and unlike filtering on
logFC afterwards it controls an error rate. It is stringent by construction: `lfc = log2(1.2)`
tests the null that |logFC| is at most 0.263, against a median interaction effect of 0.087. Three
floors are reported so the counts are read as a property of the threshold, not of the data.

```{r treat}
map(c(1.1, 1.15, 1.2), \(fc) {
  ft <- treat(fit, lfc = log2(fc))
  tibble(fold_change = fc, contrast = colnames(fit$coefficients),
         n_q_05 = map_int(colnames(fit$coefficients),
                          \(cn) sum(topTreat(ft, coef = cn, number = Inf)$adj.P.Val < 0.05)))
}) |>
  list_rbind()
```

## The pi score

`pi_score` is `P.Value^abs(logFC)`, equal to `10^(-pi)` for the score of Xiao et al. (2014, PMID
22321699), so it ranks as the published `-log10(p) * |log2FC|` does. It controls no error rate, so
it ranks within a contrast and selects nothing.

It rewards large `|logFC|`, and here the largest fold changes come from the proteins with the
fewest detected precursors, whose values lean on DPC recovery. The observation count is reported
alongside so that is visible.

```{r pi-ranking}
mean_obs <- rowMeans(proteins$other$n.observations)

results |>
  filter(contrast == "interaction") |>
  slice_min(pi_score, n = 20) |>
  mutate(mean_obs = round(mean_obs[protein], 2)) |>
  select(protein, Genes, logFC, adj.P.Val, pi_score, NPrec, mean_obs)
```

## Outputs

Written before the negative control is checked, so a failure leaves the evidence on disk.

```{r write}
out <- here("02_Differential_Expression", "02_Differential", "c_data")
write_csv(results, file.path(out, "protein_contrasts_long.csv"))
write_csv(contrast_summary, file.path(out, "contrast_summary.csv"))
saveRDS(fit, file.path(out, "fit.rds"))
```

## Negative control

Two untrained legs of the same person should not differ. If they do, the cause is upstream and
the failure is the finding.

```{r negative-control}
nc <- filter(results, contrast == d$negative_control)
stopifnot(nrow(nc) == nrow(proteins))
sum(nc$adj.P.Val < 0.05)

slice_head(nc, n = 5) |>
  select(protein, Genes, logFC, P.Value, adj.P.Val)
```

`nrow(nc)` is asserted because `sum(logical(0))` is zero, so a contrast name matching nothing
would look like a clean control. The hit count is printed rather than asserted: if two untrained
legs of one person ever differ, that is the most informative result this pipeline can produce and
it belongs in the report, not in a halted render.

```{r session}
sessionInfo()
```
