---
title: "02b · Normalisation report"
subtitle: "Does the protein matrix need correcting, and of what kind"
format:
  html:
    toc: true
    embed-resources: true
    df-print: kable
execute:
  message: false
---

limpa performs no normalisation and does not require it: DIA-NN has usually done some already.
Where it is needed, limpa's stated order puts it between `dpcQuant()` and `dpcDE()`. This report
decides the question for this matrix; `02_differential.qmd` applies the answer as one line.

The diagnostic is the MA plot. For one sample, every protein is compared against the average of all
samples: the vertical axis is that difference, the horizontal axis is how abundant the protein is.
A flat cloud at zero needs nothing. A cloud offset from zero needs a per-sample shift, which
`scale` supplies. A cloud that tilts or curves needs a correction that varies with abundance, which
only `cyclicloess` supplies. `quantile` forces samples to share a distribution, which fixes the
offset but not the tilt.

```{r setup}
library(here)
library(limpa)
library(limma)
library(dplyr)
library(purrr)
library(tibble)

proteins <- readRDS(here("01_Preprocess", "02_Quantification", "c_data", "proteins.rds"))

matrices <- map(set_names(c("none", "scale", "quantile", "cyclicloess")), \(m) {
  if (m == "none") proteins$E else normalizeBetweenArrays(proteins$E, method = m)
})
```

## The four most distorted samples, before and after

```{r ma-plots}
#| fig-height: 8
curvature <- function(E) {
  a <- rowMeans(E)
  apply(E, 2, \(x) diff(range(lowess(a, x - a, f = 0.5)$y)))
}
worst <- order(curvature(proteins$E), decreasing = TRUE)[1:4]

par(mfrow = c(4, 2), mar = c(4, 4, 2, 1))
for (j in worst) {
  for (m in c("none", "cyclicloess")) {
    E <- matrices[[m]]
    plotMA(E, array = j, main = paste(colnames(E)[j], m), ylim = c(-3, 3))
    lines(lowess(rowMeans(E), E[, j] - rowMeans(E), f = 0.5), col = "red", lwd = 2)
  }
}
par(mfrow = c(1, 1))
```

The red line is the fitted trend. Left column uncorrected, right column after `cyclicloess`. A line
that slopes or bends is the distortion this report exists to detect.

## How much curvature every sample carries

```{r curvature}
boxplot(map(matrices, curvature), ylab = "MA trend width, log2")
```

One value per sample: how far its red line travels from end to end. Zero means the sample
disagrees with the cohort by a constant, which a shift fixes. Large means it disagrees more at one
end of the abundance range, which a shift cannot fix.

## Is the distortion technical or biological

```{r association}
tibble(
  curvature = curvature(proteins$E),
  timepoint = proteins$targets$timepoint,
  treatment = proteins$targets$treatment
) |>
  summarise(median_curvature = round(median(curvature), 3), .by = c(timepoint, treatment))
```

If curvature tracked treatment or timepoint, correcting it could erase the effect under study, and
limpa's guidance is to be cautious. If it does not, the distortion is unstructured and correcting
it removes noise rather than signal.

```{r session}
sessionInfo()
```
