---
title: "02 · Set tests"
subtitle: "fry for significance, singscore for where each sample sits"
params:
  collection: combined
  set_fdr: 0.05
  signature_fdr: 0.05
  signature_max: 200
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    df-print: kable
    fig-width: 7
    fig-height: 5
    fig-dpi: 150
    fig-align: center
execute:
  message: false
---

`01_Gene_Sets` ranks proteins and asks fgsea whether a set clusters at one end. That test treats
the ranked proteins as exchangeable, and it calls sets significant on a contrast that cannot
contain signal. This stage runs the two tests that do read the design.

`fry()` asks whether a set moved, using the participant design and limpa's precision weights.
It is a rotation test, so correlated members do not inflate it. `singscore` asks where each
sample sits on each set, which is the number you can plot against the ultrasound measurements.
Neither replaces the other: one gives a p-value, the other gives 131 numbers per set.

```{r setup}
library(here)
library(limma)
library(dplyr)
library(tibble)
library(tidyr)
library(purrr)
library(ggplot2)
here::i_am("03_Pathway_Enrichment/02_Set_Tests/a_script/02_set_tests.qmd")
source(here("03_Pathway_Enrichment", "00_Functions", "00_functions.R"))

stage <- here("03_Pathway_Enrichment", "02_Set_Tests")
out <- file.path(stage, "c_data")
dir.create(out, recursive = TRUE, showWarnings = FALSE)

inputs <- stage_inputs(
  proteins = "01_Preprocess/02_Quantification/c_data/proteins.rds",
  fit = "02_Differential_Expression/02_Differential/c_data/fit.rds",
  design = "02_Differential_Expression/01_Design/c_data/design.rds",
  pathway = "03_Pathway_Enrichment/01_Gene_Sets/c_data/pathway_inputs.rds",
  phenotype = "00_Input/phenotype.csv"
)
proteins <- inputs$proteins
fit <- inputs$fit
d <- inputs$design
pathway <- inputs$pathway

protein_ids <- rownames(proteins$E)
contrast_names <- colnames(d$contrasts)
training_contrasts <- c("BFR_post_vs_pre", "HLRT_post_vs_pre")
sets <- pathway$set_lists[[params$collection]]
set_indices <- set_row_indices(sets, pathway$protein_map, protein_ids)

# set_indices is rebuilt here rather than taken from the handoff, because it must index the
# matrix this notebook actually passes to fry. A mismatch would test the wrong rows silently.
stopifnot(
  identical(colnames(proteins$E), rownames(d$design)),
  identical(rownames(fit$coefficients), protein_ids),
  identical(dim(fit$EList$weights), dim(proteins$E)),
  !any(map_lgl(set_indices, anyNA))
)
tibble(
  collection = params$collection, n_sets = length(sets),
  n_proteins = nrow(proteins$E), n_samples = ncol(proteins$E)
)
```

## Does fry need a block term

`01_Design` put participant in the design as a fixed term, so every contrast is already tested
within one person. What survives that is the correlation between the two samples of the *same
leg*, and `fry()` would take it through `block` and `correlation`.

Measured on the matrix fry is about to test, not carried across from anywhere else. A correlation
belongs to the matrix it was estimated on: this project once handed fry the singscore consensus
of 0.29 instead of the protein-level 0.24, and 29 sets crossed the FDR line on the strength of it.

```{r block-check}
leg_id <- paste(proteins$targets$participant, proteins$targets$leg, sep = "_")
rho <- duplicateCorrelation(proteins$E, d$design, block = leg_id)$consensus.correlation
round(c(within_leg_correlation = rho), 4)
```

Near zero, as it was at the design stage, so `fry()` runs without `block`. If this ever rose, the
fix is to pass `block = leg_id, correlation = rho` below — but read the number first, because
nothing here branches on it.

## fry

One call per contrast: `fry()` takes a single contrast vector, not a matrix. It reads
`fit$EList$weights`, which is what limpa's per-observation standard errors became, so a protein
reconstructed largely from missing precursors counts for less here too.

```{r fry}
fry_results <- run_fry_all(
  proteins$E, set_indices, d$design, d$contrasts,
  weights = fit$EList$weights
) |>
  mutate(collection = params$collection, .before = 1)

fry_results |>
  summarise(
    tested = n(),
    up = sum(FDR < params$set_fdr & Direction == "Up"),
    down = sum(FDR < params$set_fdr & Direction == "Down"),
    min_fdr = signif(min(FDR), 3), .by = contrast
  )
```

## The negative control

`r d$negative_control` compares two legs of one person before either was trained. A set test that
respects the design should find nothing here. fgsea did not; this is the comparison that matters.

```{r control}
fry_results |>
  filter(contrast == d$negative_control) |>
  slice_min(PValue, n = 5) |>
  select(set_id, NGenes, Direction, PValue, FDR)
```

## Where each sample sits

`singscore` ranks each sample's proteins and scores every set against those ranks, giving one
number per set per sample. It carries no p-value and no design; `simpleScore()` returns a score
and a dispersion, and the permutation route in `generateNull()` handles one set per call at a
thousand re-scorings each, answering a per-sample question that never sees the contrast. The
scores are the useful output, and fry above supplies the significance.

```{r singscore}
scores <- score_samples(proteins$E, pathway$protein_map, sets)
dim(scores)
```

Set-level scores are a different matrix from the protein one, so their intra-block correlation is
a different number and belongs only to fits run on them. It is recorded here and used nowhere
else in this notebook.

```{r set-correlation}
rho_sets <- duplicateCorrelation(scores, d$design, block = leg_id)$consensus.correlation
round(c(set_score_correlation = rho_sets, protein_correlation = rho), 4)
```

## Scores against the phenotype

`phenotype.csv` is one row per leg, which is why it is joined on participant and leg rather than
on sample. Each leg contributes two samples, so the change in a set's score across the two
timepoints is what pairs with a change in cross-sectional area.

```{r phenotype}
score_change <- scores |>
  as.data.frame() |>
  rownames_to_column("set_id") |>
  pivot_longer(-set_id, names_to = "sample_id", values_to = "score") |>
  left_join(proteins$targets, by = "sample_id") |>
  select(set_id, participant, leg, timepoint, score) |>
  pivot_wider(names_from = timepoint, values_from = score) |>
  mutate(score_change = T2 - T1) |>
  inner_join(inputs$phenotype, by = c("participant", "leg")) |>
  mutate(csa_change = vl_csa_post_cm2 - vl_csa_pre_cm2)

# Every set against the same outcome, so the ranking is the point, not any single p-value.
csa_correlation <- score_change |>
  filter(!is.na(csa_change), !is.na(score_change)) |>
  summarise(
    legs = n(),
    r = cor(score_change, csa_change, method = "spearman"), .by = set_id
  ) |>
  arrange(desc(abs(r)))

slice_head(csa_correlation, n = 10)
```

No multiplicity correction is applied to that table and none is implied: it ranks sets by how
their movement tracks hypertrophy, across `r nrow(csa_correlation)` sets, and any one row would
need its own test.

## A signature from one contrast, tested on another

Take the proteins that moved with high-load training, then ask whether restriction moved the same
ones. Building the signature from `HLRT_post_vs_pre` and testing it on `BFR_post_vs_pre` keeps the
two independent; building and testing on the same contrast would be circular and always
significant.

The same signature is tested on `r d$negative_control`, where it must find nothing. Read that row
before either of the others.

```{r signature}
signature <- pathway$protein_results |>
  filter(
    contrast == "HLRT_post_vs_pre", selected, !is.na(gene),
    adj.P.Val < params$signature_fdr
  ) |>
  slice_min(adj.P.Val, n = params$signature_max) |>
  mutate(direction = if_else(logFC > 0, "up", "down"))

signature_index <- signature |>
  split(~direction) |>
  map(\(part) match(part$protein, protein_ids))

c(up = length(signature_index$up), down = length(signature_index$down))
```

```{r signature-test}
map(set_names(c("BFR_post_vs_pre", d$negative_control, "interaction")), \(contrast) {
  fry(
    y = proteins$E, index = signature_index, design = d$design,
    contrast = d$contrasts[, contrast], weights = fit$EList$weights,
    sort = "none"
  ) |>
    rownames_to_column("direction")
}) |>
  list_rbind(names_to = "tested_on") |>
  select(tested_on, direction, NGenes, Direction, PValue, FDR)
```

Measured here, the control does not come back empty: the up half returns p = 0.030 on two
untrained legs of one person. It stays above FDR 0.05, but only just, and the adjustment runs over
two tests. A signature selected for large effects may simply be selecting the proteins with the
widest baseline spread between legs, which no amount of cross-contrast testing fixes. The
interaction row below rests on that control and should not be read as a result.

The fixed sets above do not share the problem, since they were defined before any contrast was
seen, and fry returns nothing on the control across all of them.

Every sample also gets a score on that signature, up and down together, so the response can be
plotted per participant rather than read only as a p-value.

```{r signature-scores}
gene_map <- filter(pathway$protein_map, selected)
signature_matrix <- proteins$E[gene_map$protein, ]
rownames(signature_matrix) <- gene_map$gene
signature_score <- singscore::simpleScore(
  singscore::rankGenes(signature_matrix),
  upSet = signature$gene[signature$direction == "up"],
  downSet = signature$gene[signature$direction == "down"]
) |>
  rownames_to_column("sample_id") |>
  left_join(proteins$targets, by = "sample_id")

ggplot(signature_score, aes(timepoint, TotalScore, group = participant)) +
  geom_line(alpha = 0.3) +
  geom_point(size = 1) +
  facet_wrap(~treatment) +
  labs(
    x = NULL, y = "high-load signature score",
    caption = "signature defined on HLRT_post_vs_pre; each line is one participant's leg"
  ) +
  theme_minimal()
```

## Outputs

```{r write}
write_stage_outputs(
  out, "02_set_tests",
  sheets = list(
    fry = fry_results,
    set_scores = rownames_to_column(as.data.frame(scores), "set_id"),
    score_vs_csa = csa_correlation,
    signature = signature,
    signature_scores = signature_score
  ),
  rds = list(set_tests = list(
    fry = fry_results, scores = scores, score_change = score_change,
    csa_correlation = csa_correlation, signature = signature,
    signature_score = signature_score
  )),
  csv = list(fry_results = fry_results),
  manifest = inputs$manifest, parameters = params,
  packages = c("here", "limma", "limpa", "singscore", "dplyr", "tibble", "tidyr", "purrr"),
  notes = list(
    protein_correlation = rho, set_score_correlation = rho_sets,
    block = paste(
      "fry run without block: participant is fixed in the design and the",
      "residual within-leg correlation is near zero"
    ),
    signature = "Defined on HLRT_post_vs_pre, tested on BFR_post_vs_pre and the negative control"
  )
)
```

```{r session}
sessionInfo()
```
