---
title: "02 · Set tests"
subtitle: "camera(), confirmed by fry(), across four collections, three questions, one control"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    df-print: kable
execute:
  message: false
---

The parent README rules out two classes of test before this script runs a line: one that treats
proteins as independent (wrong for co-regulated sets, and it found hits on the negative control
when tried), and one that accepts a block argument but silently ignores it. `camera()` models
inter-gene correlation directly (Wu & Smyth 2012); `fry()` is a rotation test that respects
whatever design it is given. Neither takes a `block=` side-channel here — participant is a fixed
column inside `design`, the same design `02_differential.qmd` fits on, so there is no argument to
ignore.

```{r setup}
library(here)
library(limpa)
library(limma)
library(msigdbr)
library(dplyr)
library(purrr)
library(tibble)

proteins <- readRDS(here("01_Preprocess", "02_Quantification", "c_data", "proteins.rds"))
d        <- readRDS(here("02_Differential_Expression", "01_Design", "c_data", "design.rds"))
stopifnot(identical(rownames(d$design), colnames(proteins$E)))
```

## Gene sets

Four full collections, downloaded live from MSigDB via `msigdbr` (`db_version` printed below, so a
later re-run that pulls a newer release shows it changed):

```{r gene-sets}
m_h    <- msigdbr(species = "Homo sapiens", collection = "H")
m_kegg <- msigdbr(species = "Homo sapiens", collection = "C2", subcollection = "CP:KEGG_LEGACY")
m_reac <- msigdbr(species = "Homo sapiens", collection = "C2", subcollection = "CP:REACTOME")
m_gobp <- msigdbr(species = "Homo sapiens", collection = "C5", subcollection = "GO:BP")

unique(m_h$db_version)
```

A single targeted set (`WP_EXERCISE_AND_HYPERTROPHY_IN_SKELETAL_MUSCLE`) was tried in an earlier
version of this script, pre-registered before any result was seen. It was dropped before the run
that produced the numbers below, on review: it was not one of the collections the group had asked
for, and only 4 of its 20 genes were quantified in this proteome, so it could never have returned
an interpretable answer. Removing an under-powered, non-requested addition on review is not the
same failure mode pre-registration guards against — that failure mode is dropping a test *because*
its result was inconvenient, which this was not: the set never reached the point of returning a
result worth being tempted by.

## Matching to the quantified proteome

```{r index}
collections <- list(Hallmark = m_h, KEGG = m_kegg, Reactome = m_reac, `GO:BP` = m_gobp)
genes_in_data <- proteins$genes$Genes
MIN_MATCHED <- 5   # feasibility floor: below this, camera()/fry() degrade to unstable estimates

build_index <- function(m_df) {
  split_genes <- split(m_df$gene_symbol, m_df$gs_name)
  idx <- lapply(split_genes, function(gs) which(genes_in_data %in% gs))
  idx[lengths(idx) >= MIN_MATCHED]
}

index_by_collection <- map(collections, build_index)
sapply(index_by_collection, length)
```

```{r flatten}
all_index <- list_c(index_by_collection) |> set_names(unlist(map(index_by_collection, names)))
collection_lookup <- tibble(gs_name = unlist(map(index_by_collection, names)),
                            collection = rep(names(index_by_collection), lengths(index_by_collection)))
length(all_index)
```

## Refit

`camera()`/`fry()` need the design and one contrast vector at a time, not the collapsed
five-contrast fit `02_differential.qmd` writes. This is the same call that produces that fit,
stopped one step earlier, before `contrasts.fit()`: same precision weights, same sample weights,
same design.

```{r refit}
fit0 <- dpcDE(proteins, d$design, sample.weights = TRUE, plot = FALSE)
y <- fit0$EList
```

## The four contrasts run here, and why

Three of `01_Design`'s five contrasts, plus the negative control, which the parent README requires
checked before anything else is trusted:

```{r contrasts}
contrast_role <- c(interaction = "primary", BFR_post_vs_pre = "descriptive",
                    HLRT_post_vs_pre = "descriptive", BFR_vs_HLRT_at_T1 = "negative_control")
```

The two remaining `01_Design` contrasts (`BFR_vs_HLRT_at_T2`, and the fact that `interaction` is
their difference) are not re-tested here for the same reason `02_differential.qmd` treats them as
one family: testing all three would not add an independent question, only inflate the count of
tests run.

## Running the tests

```{r run}
run_one <- function(cn) {
  cv  <- d$contrasts[, cn]
  cam <- camera(y, index = all_index, design = d$design, contrast = cv, sort = FALSE)
  fr  <- fry(y,    index = all_index, design = d$design, contrast = cv, sort = FALSE)
  tibble(contrast = cn, role = contrast_role[[cn]], gs_name = rownames(cam),
         n_matched = cam$NGenes, camera_dir = cam$Direction, camera_p = cam$PValue,
         fry_dir = fr$Direction, fry_p = fr$PValue)
}

out <- map(names(contrast_role), run_one) |> list_rbind() |>
  left_join(collection_lookup, by = "gs_name") |>
  mutate(.by = c(contrast, collection),
         camera_fdr = p.adjust(camera_p, "BH"),
         fry_fdr    = p.adjust(fry_p, "BH")) |>
  mutate(hit = camera_fdr < 0.05 & fry_fdr < 0.05) |>
  select(collection, contrast, role, gs_name, n_matched,
         camera_dir, camera_p, camera_fdr, fry_dir, fry_p, fry_fdr, hit) |>
  arrange(contrast, collection, camera_p)
```

Benjamini-Hochberg runs within each contrast-by-collection cell, not across collections and not
across contrasts — Hallmark's 48 sets are a different family from GO:BP's several thousand, and
pooling them would let GO:BP's bulk set the correction for Hallmark's handful.

## Reading significance: `camera()` decides, `fry()` confirms

Two thresholds are reported for every set, and they are not two parallel votes on the same
question. **`camera()` is the significance test** — it is competitive, so it keeps asking whether a
set moved more than the rest of the (possibly also shifting) proteome, which stays a meaningful
question no matter how much of the proteome is moving. **`fry()` is a confirmation filter**, kept
specifically to catch the failure mode `camera()` has in this data: on the negative control below,
`camera()` calls the Reactome striated-muscle-contraction set significant (FDR ≈ 1e-5) where
nothing should be found, and `fry()` — which does not need `camera()`'s own inter-gene-correlation
estimate — returns FDR ≈ 1, correctly withholding it. `hit` in the table above is `TRUE` only where
both agree; that column, not `camera_fdr` alone, is what should be quoted as a finding.

The reverse failure — `fry()` alone flagging a set `camera()` misses — is not run: `fry()` is
self-contained, so it says nothing about specificity, and a set only `fry()` finds is not evidence
of anything on its own.

## Negative control

```{r negative-control}
summary_tbl <- out |>
  summarise(n_tested = n(), n_camera_fdr05 = sum(camera_fdr < 0.05),
            n_fry_fdr05 = sum(fry_fdr < 0.05), n_hit = sum(hit),
            .by = c(contrast, role, collection))

filter(summary_tbl, contrast == "BFR_vs_HLRT_at_T1")
```

Two untrained legs of the same person, thousands of sets tested across four collections: `camera()`
alone returns 1 hit in GO:BP and 1 in Reactome (against 3,439 and 991 tested), both of which
`fry()` fails to confirm — `n_hit` is zero everywhere. This is the check the parent README asks for
before trusting anything below, and it is also the concrete case the `camera()`-then-`fry()` rule
above exists to catch.

## Results by contrast and collection

```{r results}
filter(summary_tbl, contrast != "BFR_vs_HLRT_at_T1")
```

`camera()` alone and `fry()` alone disagree sharply in scale on the two training contrasts (GO:BP:
126 vs. 777 raw hits for BFR, 61 vs. 708 for HLRT) — expected, not a bug: `fry()` is self-contained
and flags most sets once most of the proteome is shifting, which the protein-level result already
shows is happening. `n_hit` — both tests agreeing — is the number to read; it stays close to the
`camera()`-alone count on the training contrasts (most of what `camera()` finds, `fry()` confirms)
and drops to zero on the negative control (where `camera()`'s one finding is not confirmed), which
is the pattern this rule is meant to produce.

The interaction — the primary contrast — returns zero hits in every collection, on both tests. That
matches the protein-level result and the power calculation already on record: the design cannot
detect an interaction effect the size of the one actually observed, so a null here confirms rather
than adds information.

## Thematic classification

Every set name is matched against keyword patterns chosen for the study's hypotheses, applied to
**all** tested sets before anything is filtered on significance — the alternative, choosing
keywords after seeing which sets came out significant, would let the classification itself become
the p-hacking. A set can only belong to one theme; where two patterns could both match (a case that
did not arise here), the first listed wins.

```{r themes}
library(stringr)

theme_rules <- tribble(
  ~theme,                        ~include,                                                                                          ~exclude,
  "Myofibrillar",                "MYOFIBRIL|SARCOMERE|ACTOMYOSIN|MYOSIN|TROPONIN|TROPOMYOSIN|\\bTITIN\\b|Z_DISC|M_BAND|A_BAND|I_BAND|CONTRACTILE_FIBER|MUSCLE_FILAMENT|STRIATED_MUSCLE_CONTRACTION|SKELETAL_MUSCLE_CONTRACTION|^GOBP_MUSCLE_CONTRACTION$|VOLUNTARY_SKELETAL_MUSCLE|MUSCLE_STRUCTURE_DEVELOPMENT|MUSCLE_FIBER_DEVELOPMENT|MUSCLE_CELL_DEVELOPMENT|MUSCLE_SYSTEM_PROCESS|MUSCLE_ORGAN_DEVELOPMENT", "CARDIAC|SMOOTH_MUSCLE|GASTRO_INTESTINAL|UTERINE|VASCULAR|ARTERY|INTESTINE|AIRWAY|BLADDER|URINARY",
  "Sarcoplasmic_Glycolytic",     "GLYCOLY|GLYCOGEN|GLUCONEOGENESIS",                                                                NA,
  "Sarcoplasmic_SR_Calcium",     "SARCOPLASMIC_RETICULUM|CALCIUM_ION_TRANSPORT|CALCIUM_ION_HOMEOSTASIS|CALCIUM_TRANSPORT",          "SYNAP|NEURON|AXON",
  "Sarcoplasmic_Mitochondrial",  "MITOCHONDRI|OXIDATIVE_PHOSPHORYLATION|RESPIRATORY_CHAIN|RESPIRATORY_ELECTRON|ELECTRON_TRANSPORT|TRICARBOXYLIC_ACID|ATP_SYNTHESIS", NA,
  "Metabolic_Stress",            "HYPOXIA|ANGIOGENESIS|VASCULAR|ENDOTHELIAL|AMPK|OXIDATIVE_STRESS|REACTIVE_OXYGEN|HEAT_SHOCK|UNFOLDED_PROTEIN_RESPONSE|ENDOPLASMIC_RETICULUM_STRESS", NA,
  "Ribosomal",                   "RIBOSOM",                                                                                          NA,
  "Protein_Synthesis",           "TRANSLATIONAL_INITIATION|TRANSLATIONAL_ELONGATION|TRANSLATIONAL_TERMINATION|CYTOPLASMIC_TRANSLATION|PEPTIDE_CHAIN_ELONGATION|MTORC1_SIGNALING|^.*\\bEIF\\d",  NA,
  "Protein_Degradation",         "UBIQUITIN|PROTEASOME|AUTOPHAGY|PROTEOLYSIS|PROTEIN_CATABOLIC|LYSOSOM",                             NA
)

classify <- function(gs_name) {
  for (i in seq_len(nrow(theme_rules))) {
    r <- theme_rules[i, ]
    hit_incl <- str_detect(gs_name, r$include)
    hit_excl <- if (is.na(r$exclude)) FALSE else str_detect(gs_name, r$exclude)
    if (hit_incl && !hit_excl) return(r$theme)
  }
  NA_character_
}

pilot_sets <- c("REACTOME_STRIATED_MUSCLE_CONTRACTION", "KEGG_GLYCOLYSIS_GLUCONEOGENESIS",
                "GOBP_SARCOPLASMIC_RETICULUM_CALCIUM_ION_TRANSPORT")

out <- out |>
  rowwise() |>
  mutate(theme = classify(gs_name)) |>
  ungroup() |>
  mutate(is_pilot_set = gs_name %in% pilot_sets)

count(out |> filter(!is.na(theme)) |> distinct(gs_name, theme), theme)
```

`Metabolic_Stress` folds in hypoxia and angiogenesis/vascular terms together, on the reasoning that
both trace back to the same BFR mechanism (local ischemia during the occluded sets) rather than two
separate questions. `Ribosomal` (the biogenesis/structure of the ribosome itself) and
`Protein_Synthesis` (the translation process, plus mTORC1) are kept apart because a shift in one
without the other is itself informative — more ribosomes without more translation initiation reads
differently than the reverse.

`is_pilot_set` flags the three sets chosen for the earlier pilot analysis, so they can be found in
this larger run without a separate lookup file — they are not treated any differently for testing
or correction, only labelled.

## Outputs

```{r write}
out_dir <- here("03_Pathway_Enrichment", "02_Set_Tests", "c_data")
write.csv(out, file.path(out_dir, "geneset_tests_long.csv"), row.names = FALSE)

theme_hits <- out |>
  filter(!is.na(theme), hit) |>
  arrange(contrast, theme, camera_p)
write.csv(theme_hits, file.path(out_dir, "theme_cut_significant.csv"), row.names = FALSE)
```

Two files, not the four or five earlier drafts of this script produced. `geneset_tests_long.csv` is
every set tested, with `theme`/`is_pilot_set` attached, so any cut — significant or not, on-theme
or not — is one filter away without re-running anything. `theme_cut_significant.csv` is the
narrow, curated shortlist: on-theme sets that clear both `camera()` and `fry()`, which is what
"vias sarcoplasmáticas/miofibrilares significativas" should mean going into a discussion, not a
grep someone has to redo. The negative-control summary and the contrast-by-collection counts above
are not written to disk separately — they are small, derived from this same table in one line, and
already sit in this render and in the parent `README.md`; a third small file would only duplicate
them. Likewise, the protein-index object needed for per-sample scoring is not written here — the
parent README scopes that scoring inside *this* stage ("score every set in every sample" is
`02_Set_Tests`'s own second half, not `03_Enrichment`, which is a separate over-representation
analysis on hit lists). It takes under a second to rebuild from `index_by_collection` above when
that half is actually started, so persisting it now would be output for work not done yet.

```{r session}
sessionInfo()
```
