---
title: "02 · Quantification"
subtitle: "Detection probability curve, protein roll-up, detection filter, normalisation"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    df-print: kable
    fig-width: 7
    fig-height: 5
    fig-dpi: 150
    fig-align: center
execute:
  message: false
---

Half of a DIA matrix is empty, and fainter precursors go missing more often. Following limpa's
vignette, `dpc()` fits that as a curve and `dpcQuant()` uses it to give every protein a value in
every sample without imputation.

```{r setup}
pacman::p_load(here, limpa, limma, tibble, purrr, writexl)

y <- readRDS(here("01_Preprocess", "01_Filtering", "c_data", "precursors_filtered.rds"))

checkpoints <- here("01_Preprocess", "02_Quantification", "c_data", "quant_runs")
sensitivity_file <- file.path(checkpoints, "proteins_slope_fitted.rds")
proteins <- readRDS(file.path(checkpoints, "proteins_slope_0.7.rds"))
```

## Missingness

```{r ave-vs-mis}
plotAveVsMis(y)
```

A downward trend means detection depends on intensity, which the DPC fits. A flat cloud would
mean near-random missingness, where limpa gains less over simpler methods.

## The curve

```{r dpc}
dpcest <- dpc(y$E)
plotDPC(dpcest)

dpc_parameters <- tibble(intercept = dpcest$dpc[1], slope = dpcest$dpc[2])
round(dpc_parameters, 3)
```

`dpc()` defaults to `model = "cn"`, which returns `dpcCN()`, the complete-normal estimator the
limpa FAQ recommends for noisy data. `subset = 2000` fits a systematic sample stratified by
missingness and mean intensity rather than all `r format(nrow(y), big.mark = ",")` rows. `robust`
is *"Only used if `model='on'"*, so here the subsample supplies the robustness.

The slope of `r round(dpcest$dpc[2], 3)` sits below the 0.7-to-0.9 band the FAQ calls typical for
DIA-NN with match-between-runs, inside the 0.1-to-1.0 range it calls usable.

Quantification therefore uses the FAQ's preset: "If `dpcCN()` doesn't give a value in that
range, it would be reasonable to run `dpcQuant()` with a preset value `dpc.slope=0.7`."

The FAQ says a slope set too low makes the analysis more conservative. Here the reverse held: the
lower fitted slope gave smaller standard errors and more hits. The preset is the documented
setting and, on this data, the cautious one, chosen before any downstream count was consulted.

::: {.callout-note}
The visual fit is not diagnostic: `plotDPC()` draws a curve in *unobserved* intensity against
proportions computed on *observed* intensity, so curve and points disagree by construction. Judge
the slope.
:::

## Roll-up

`dpcQuant()` fits an additive precursor-plus-sample model per protein, maximising a posterior
that scores observed values by squared residual and missing values by their DPC probability. It
returns `$other$standard.error`, which `dpcDE()` reads as a precision weight, so a protein built
largely from missing precursors gets a wide standard error and counts for less.

The roll-up takes about 100 minutes, so `a_script/02_quantify_run.R` runs it and this notebook
loads the result. Rerun the script by hand when the precursor matrix changes.

`r format(nrow(proteins), big.mark = ",")` protein groups come from
`r format(nrow(y), big.mark = ",")` precursors, with no missing values; each value carries its own
standard error.

## Detection filter

`filterByDetection()` runs between `dpcQuant()` and DE. For a small experiment its help sets
`n.samples` to the smallest group size: `BFR_T2`, 32 samples, one short because S06_L_T2 is
missing.

It counts detections across all samples, so the bar is cohort-wide and does not guarantee
detection within any one group.

```{r detection}
group_sizes <- table(proteins$targets$group)
detected <- filterByDetection(proteins, n.samples = min(group_sizes))
proteins <- proteins[detected, ]

group_sizes
c(before = length(detected), kept = sum(detected), dropped = sum(!detected))
```

On an earlier run, dropped proteins had a mean SE of 1.56 against 0.66 for those kept, and 0.15
detected precursors per sample against 7.48; the typical sample measured nothing for them.
limpa's default threshold of 3 recovered 371, one of which reached BH < 0.05 in any contrast,
while the larger test set tightened BH enough to push 14 and 29 significant proteins above 0.05 on
the two training contrasts.

## Normalisation

DIA-NN normalised the precursors (`Precursor.Normalised`), but that leaves the protein roll-up
unnormalised, and limpa normalises nothing itself.

A per-sample shift fixes the spread of per-sample medians but not the MA curve width, which
cyclic loess targets. The MA curve fits a sample's difference from the all-sample average against
abundance: flat means a constant offset, travelling means the offset depends on abundance.

```{r normalise}
ma_width <- function(m) {
  a <- rowMeans(m)
  apply(m, 2, \(x) diff(range(lowess(a, x - a, f = 0.5)$y)))
}
spread <- function(m) diff(range(apply(m, 2, median)))

unnormalised <- proteins$E
widths <- ma_width(unnormalised)
before <- c(ma_width = median(widths), median_spread = spread(unnormalised))
proteins$E <- normalizeBetweenArrays(proteins$E, method = "cyclicloess")
after <- c(ma_width = median(ma_width(proteins$E)), median_spread = spread(proteins$E))

round(rbind(before, after), 3)
```

Top: every sample's intensity density. Bottom: the sample whose curve travelled furthest before
normalisation, against the all-sample average.

```{r normalise-plots}
#| fig-height: 8
worst <- which.max(widths)
par(mfrow = c(2, 2))
plotDensities(unnormalised, legend = FALSE, main = "Before cyclic loess")
plotDensities(proteins$E, legend = FALSE, main = "After cyclic loess")
plotMD(unnormalised, column = worst, main = paste(colnames(unnormalised)[worst], "before"))
plotMD(proteins$E, column = worst, main = paste(colnames(proteins$E)[worst], "after"))
```

The curvature is unrelated to treatment or timepoint, so cyclic loess is not absorbing the tested
effect. `proteins.rds`, written after this step, is the matrix the model is fitted to.

::: {.callout-warning}
## The standard errors stay on the old scale

Cyclic loess remaps `$E` non-linearly and leaves `$other$standard.error` on the unnormalised
scale, which `dpcDE()` reads as though it were the new one. Smyth recommends quantile or cyclic-loess normalisation of limpa
protein quantifications when sample differences are technical
([Bioconductor support, 2026](https://support.bioconductor.org/p/9163187/#9163221)). The mapping
is close to linear, but it is an approximation carried into every later stage.
:::

## What survives

```{r quality}
protein_quality <- tibble(
  protein = rownames(proteins),
  gene = proteins$genes$Genes,
  n_precursors = proteins$genes$NPrec,
  prop_observed = proteins$genes$PropObs,
  mean_log2 = rowMeans(proteins$E),
  mean_standard_error = rowMeans(proteins$other$standard.error),
  mean_n_observations = rowMeans(proteins$other$n.observations)
)
summary(protein_quality[c("mean_log2", "mean_standard_error", "prop_observed")])
```

The installed version writes `NPrec`; the help page calls it `NPeptides`. A high mean standard
error with few observations marks an estimate drawn from the DPC.

Points are coloured by participant. 32 of 33 participants contribute all four cells, so
between-participant variation dominates; the design's participant term removes it. Check any
sample far from its own participant.

```{r mds}
mds <- plotMDSUsingSEs(proteins, pch = 16, col = as.integer(factor(proteins$targets$participant)))
r2 <- function(f) summary(lm(mds$x ~ factor(f)))$r.squared
round(map_dbl(proteins$targets[c("participant", "group")], r2), 3)
```

## Slope sensitivity

The sensitivity run quantifies against the whole fitted curve, intercept included, where the
primary run passes `dpc.slope = 0.7` and lets limpa fit the rest. Both get the same detection
filter and normalisation. The run with the larger mean standard error is the more conservative.

The median absolute abundance difference follows the table. It measures how far values move,
not whether contrasts change, since within-participant comparisons cancel a shared shift.

```{r slope-sensitivity}
#| eval: !expr file.exists(sensitivity_file)
sensitivity <- readRDS(sensitivity_file)
sensitivity <- sensitivity[filterByDetection(sensitivity, n.samples = min(group_sizes)), ]
sensitivity$E <- normalizeBetweenArrays(sensitivity$E, method = "cyclicloess")

shared <- intersect(rownames(proteins), rownames(sensitivity))
tibble(
  slope = c(0.7, round(dpcest$dpc[2], 3)),
  role = c("primary", "sensitivity"),
  n_proteins = c(nrow(proteins), nrow(sensitivity)),
  n_shared = length(shared),
  mean_standard_error = round(c(
    mean(proteins$other$standard.error[shared, ]),
    mean(sensitivity$other$standard.error[shared, ])
  ), 3)
)

round(median(abs(proteins$E[shared, ] - sensitivity$E[shared, ])), 4)
```

This section renders only when `c_data/quant_runs/proteins_slope_fitted.rds` exists, which costs
another 100 minutes to produce:

```sh
Rscript 01_Preprocess/02_Quantification/a_script/02_quantify_run.R --sensitivity
```

`02_Differential` refits the five contrasts under both slopes and counts the hits.

## Outputs

`proteins.rds` holds abundance, standard error and detected-precursor count per protein and
sample. Every sub-stage of `02_Differential_Expression` reads it, as do `00_build_gene_sets`,
`01_run_fgsea_and_fry`, `04_run_singscore` and `05_classify_and_associate_sets`. Keep the `.rds`:
abundances alone lose the standard errors.

```{r write}
sheets <- list(dpc_parameters = dpc_parameters, protein_quality = protein_quality)
overview <- data.frame(
  sheet = names(sheets), rows = map_int(sheets, nrow), columns = map_int(sheets, ncol),
  description = c(
    "Fitted DPC intercept and slope",
    "Per-protein coverage, abundance and standard error"
  )
)

out <- here("01_Preprocess", "02_Quantification", "c_data")
saveRDS(proteins, file.path(out, "proteins.rds"))
write_xlsx(c(list(overview = overview), sheets), file.path(out, "02_quantify.xlsx"))
```

```{r session}
sessionInfo()
```
