---
title: "02 · Quantification"
subtitle: "Detection probability curve, protein roll-up, detection filter, normalisation"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    df-print: kable
    fig-width: 7
    fig-height: 5
    fig-dpi: 150
    fig-align: center
execute:
  message: false
---

Half of a DIA matrix is empty, and not at random: faint precursors go missing more often
than bright ones. `dpc()` fits that as a curve, and `dpcQuant()` uses it to give every
protein a value in every sample without imputing anything. This stage is the limpa
vignette's own sequence, unaltered.

```{r setup}
library(here)
library(limpa)
library(limma)
library(dplyr)
library(tibble)
library(writexl)

precursors_file <- here("01_Preprocess", "01_Filtering", "c_data",
                        "precursors_filtered.rds")
y <- readRDS(precursors_file)
dim(y)

checkpoints <- here("01_Preprocess", "02_Quantification", "c_data", "dpcQuant")
primary_file <- file.path(checkpoints, "proteins_slope_0.7.rds")
sensitivity_file <- file.path(checkpoints, "proteins_slope_fitted.rds")
```

## The missingness relationship

```{r ave-vs-mis}
plotAveVsMis(y)
```

A downward trend means detection depends on intensity and the DPC has something to fit. A
flat cloud would mean missingness is closer to random, where limpa still works but gains
less over simpler methods.

## The curve

```{r dpc}
dpcest <- dpc(y$E)
plotDPC(dpcest)

dpc_parameters <- tibble(
  intercept = dpcest$dpc[1],
  slope = dpcest$dpc[2]
)
round(dpc_parameters, 3)
```

`dpc()` defaults to `model = "cn"` and its help says it then returns `dpcCN()`, the
complete-normal estimator the limpa FAQ recommends for noisy data. Its `subset = 2000`
default fits on a systematic sample stratified by missingness and mean intensity rather
than all `r format(nrow(y), big.mark = ",")` rows. limpa documents `robust` as *"Only used
if `model='on'"*, so it does nothing here and the subsample supplies the robustness.

The fit comes out at `r round(dpcest$dpc[2], 3)`, below the 0.7-to-0.9 band the FAQ calls typical
for DIA-NN searched with match-between-runs, though inside the 0.1-to-1.0 range it calls usable.
The vignette notes the slope comes out low when peptides vary a lot, which is what a
within-subject design with heavy missingness looks like.

Quantification therefore uses a preset 0.7, which is the rule limpa's FAQ gives for this exact
situation: "If `dpcCN()` doesn't give a value in that range, it would be reasonable to run
`dpcQuant()` with a preset value `dpc.slope=0.7`." The fitted value is carried into a sensitivity
run at the end. The choice follows the documented rule, not any downstream count.

One caveat belongs on the record. The FAQ says a slope set too low makes the analysis more
conservative. On this data the reverse held when it was last measured: the lower fitted slope gave
smaller standard errors and more hits than the preset. The preset is still both the documented
setting and the cautious one here, but the direction is not what the documentation predicts.

::: {.callout-note}
The visual fit is not the diagnostic. `plotDPC()` draws a curve in *unobserved* intensity
against proportions computed on *observed* intensity, so some discrepancy is structural.
Judge the slope.
:::

## Roll-up

`dpcQuant()` fits an additive precursor-plus-sample model per protein, maximising a
posterior that scores observed values by squared residual and missing values by their DPC
probability. The returned `$other$standard.error` is why this project uses limpa: `dpcDE()`
reads it as a precision weight, so a protein reconstructed largely from missing precursors
gets a value and a wide one. Handing `$E` to `lmFit()` discards that and treats every
protein as equally reliable.

The call itself lives in `a_script/dpc_quant.R`, not here, because it costs about 100 minutes.
This notebook only loads what that script wrote, so rendering is always cheap and can never
start a long run by accident. Rerun the script by hand when the precursor matrix changes.

```{r quantify}
if (!file.exists(primary_file)) {
  stop("No dpcQuant checkpoint. Run:\n",
    "  Rscript 01_Preprocess/02_Quantification/a_script/dpc_quant.R")
}
proteins <- readRDS(primary_file)$object
```

`r format(nrow(proteins), big.mark = ",")` protein groups from
`r format(nrow(y), big.mark = ",")` precursors. The matrix has no missing values; what varies
is how certain each value is.

## Detection filter

`filterByDetection()`'s help fixes the order: it runs after `dpcQuant()` and before DE, and
it returns a logical vector rather than an object. For a small experiment the help suggests
setting `n.samples` to the smallest group size. Here that is `BFR_T2` with 32 samples, one
short because nobody acquired S06's left leg at T2.

The function does not read `group`. It counts detections across all samples and keeps a
protein when that count clears the threshold, so the smallest group size sets a
cohort-wide bar rather than guaranteeing detection within any one group.

```{r detection}
group_sizes <- table(proteins$targets$group)
keep <- filterByDetection(proteins, n.samples = min(group_sizes))
proteins <- proteins[keep, ]

group_sizes
c(before = length(keep), kept = sum(keep), dropped = sum(!keep))
```

The proteins this drops are not borderline. Measured on an earlier run, they carried a mean
standard error of 1.56 against 0.66 for those kept, and 0.15 detected precursors per sample
against 7.48, so the typical sample measured nothing at all for them. Relaxing the threshold to
limpa's default of 3 recovered 371 of them, of which exactly one reached BH < 0.05 in any
contrast, while the extra tests cost 14 and 29 proteins on the two training contrasts.

## Normalisation

limpa places normalisation after `dpcQuant()` and before `dpcDE()` and performs none itself,
because DIA-NN has usually normalised the precursors already. What matters is which kind is
needed, and two numbers separate the cases.

The spread in per-sample medians is what a single per-sample shift would fix. The MA curve width
is what only cyclic loess fixes: comparing a sample against the average of all samples gives a
difference per protein, and fitting that against abundance gives a curve. A flat curve means the
sample disagrees by a constant. A curve that travels means the disagreement depends on abundance.
Both numbers are reported before and after so the correction can be judged rather than assumed.

```{r normalise}
ma_width <- function(E) {
  a <- rowMeans(E)
  apply(E, 2, \(x) diff(range(lowess(a, x - a, f = 0.5)$y)))
}
spread <- function(E) diff(range(apply(E, 2, median)))

before <- c(ma_width = median(ma_width(proteins$E)), median_spread = spread(proteins$E))
proteins$E <- normalizeBetweenArrays(proteins$E, method = "cyclicloess")
after <- c(ma_width = median(ma_width(proteins$E)), median_spread = spread(proteins$E))

round(rbind(before, after), 3)
```

The curvature is unrelated to treatment or timepoint, so cyclic loess is not absorbing the effect
being tested. `proteins.rds` is written after this, so it is the matrix the model is fitted to.

::: {.callout-warning}
## An approximation, stated plainly

Cyclic loess remaps `$E` non-linearly while `$other$standard.error` is left untouched. After
normalisation the stored errors describe the unnormalised scale and `dpcDE()` reads them as though
they described the new one. Smyth recommends quantile and cyclic loess for this data regardless,
and the mapping is close enough to linear that the discrepancy is small, but it is an
approximation rather than an exact correspondence.
:::

## What survives

```{r quality}
protein_quality <- tibble(
  protein = rownames(proteins),
  gene = proteins$genes$Genes,
  n_precursors = proteins$genes$NPrec,
  prop_observed = proteins$genes$PropObs,
  mean_log2 = rowMeans(proteins$E),
  mean_standard_error = rowMeans(proteins$other$standard.error),
  mean_n_observations = rowMeans(proteins$other$n.observations)
)
summary(protein_quality[c("mean_log2", "mean_standard_error", "prop_observed")])
```

`NPrec` is what the installed version writes; the help page calls it `NPeptides`. A protein
with a high mean standard error and few observations is one whose estimate leans on the DPC
rather than on data.

Points are coloured by participant, and the variance each factor explains in the leading dimension
is printed below the plot. Thirty-two of the 33 participants contribute all four cells, so
between-participant variation dominates and the participant term in the design is what removes
it. A sample sitting
far from its own participant is worth investigating; an absence of group structure is not.

```{r mds}
#| fig-height: 5
mds <- plotMDSUsingSEs(proteins,
  pch = 16,
  col = as.integer(factor(proteins$targets$participant))
)

r2 <- function(f) summary(lm(mds$x ~ factor(f)))$r.squared
round(c(
  participant = r2(proteins$targets$participant),
  group = r2(proteins$targets$group)
), 3)
```

## Slope sensitivity

Quantification is repeated at the fitted slope and compared against the primary run, with the same
detection filter and normalisation applied so the slope is the only difference. `dpcDE()` reads
the standard errors as precision weights, so whichever run carries the larger mean is the more
conservative setting.

The median absolute difference in protein abundance follows the table. It measures how far values
move, not whether that movement survives a contrast: every comparison here is within one
participant, so a shift common to a participant's samples cancels in the subtraction.

```{r slope-fitted}
#| eval: !expr file.exists(sensitivity_file)
sensitivity <- readRDS(sensitivity_file)$object
```

```{r slope-sensitivity}
#| eval: !expr file.exists(sensitivity_file)
sensitivity <- sensitivity[filterByDetection(sensitivity, n.samples = min(group_sizes)), ]
sensitivity$E <- normalizeBetweenArrays(sensitivity$E, method = "cyclicloess")

shared <- intersect(rownames(proteins), rownames(sensitivity))
tibble(
  slope = c(0.7, round(dpcest$dpc[2], 3)),
  role = c("primary", "sensitivity"),
  n_proteins = c(nrow(proteins), nrow(sensitivity)),
  n_shared = length(shared),
  mean_standard_error = round(c(
    mean(proteins$other$standard.error[shared, ]),
    mean(sensitivity$other$standard.error[shared, ])
  ), 3)
)

round(median(abs(proteins$E[shared, ] - sensitivity$E[shared, ])), 4)
```

This section renders only when `c_data/dpcQuant/proteins_slope_fitted.rds` is present, since
computing it costs about another 100 minutes. To produce it once:

```sh
Rscript 01_Preprocess/02_Quantification/a_script/dpc_quant.R --sensitivity
```

Afterwards it loads like the primary run does.

Refitting the contrasts under each slope is a separate exercise and is not done here; the counts
it produced are recorded in this stage's README.

## Outputs

`proteins.rds` is read by both stages of `02_Differential_Expression`.

```{r write}
out <- here("01_Preprocess", "02_Quantification", "c_data")
saveRDS(proteins, file.path(out, "proteins.rds"))
write_xlsx(
  list(dpc_parameters = dpc_parameters, protein_quality = protein_quality),
  file.path(out, "02_quantify.xlsx")
)
```

```{r session}
sessionInfo()
```
