---
title: "02 · Quantification"
subtitle: "Detection probability curve, protein roll-up, detection filter, normalisation"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    df-print: kable
execute:
  message: false
---

Half of a DIA matrix is empty, and not at random: faint precursors go missing more often
than bright ones. `dpc()` fits that as a curve, and `dpcQuant()` uses it to give every
protein a value in every sample without imputing anything. This stage is the limpa
vignette's own sequence, unaltered.

```{r setup}
library(here)
library(limpa)
library(limma)
library(dplyr)
library(readr)
library(tibble)

y <- readRDS(here("01_Preprocess", "01_Filtering", "c_data", "precursors_filtered.rds"))
dim(y)
```

## The missingness relationship

```{r ave-vs-mis}
plotAveVsMis(y)
```

A downward trend means detection depends on intensity and the DPC has something to fit. A
flat cloud would mean missingness is closer to random, where limpa still works but gains
less over simpler methods.

## The curve

```{r dpc}
dpcest <- dpc(y$E)
dpcest_on <- dpcON(y$E, robust = TRUE)
plotDPC(dpcest)

dpc_parameters <- tibble(
  estimator = c("complete normal", "observed normal"),
  intercept = c(dpcest$dpc[1], dpcest_on$dpc[1]),
  slope = c(dpcest$dpc[2], dpcest_on$dpc[2])
)
dpc_parameters
```

`dpc()` defaults to `model = "cn"` and its help says it then returns `dpcCN()`, the
complete-normal estimator the limpa FAQ recommends for noisy data. Its `subset = 2000`
default fits on a systematic sample stratified by missingness and mean intensity rather
than all `r format(nrow(y), big.mark = ",")` rows. limpa documents `robust` as *"Only used
if `model='on'"*, so it does nothing here and the subsample supplies the robustness.

We report the slope as fitted and never substitute a preset. The FAQ puts the usable range at
0.1 to 1.0 and calls 0.7 to 0.9 typical for DIA-NN searched with match-between-runs. The
vignette adds that the slope comes out low when peptides vary a lot, which is what a
within-subject design with heavy missingness looks like. limpa's own published DIA-NN case
study fitted 0.59 and used it. A shallow slope also recovers less from missing values, so it
errs toward finding nothing rather than finding too much.

::: {.callout-note}
The visual fit is not the diagnostic. `plotDPC()` draws a curve in *unobserved* intensity
against proportions computed on *observed* intensity, so some discrepancy is structural.
Judge the slope.
:::

## Roll-up

`dpcQuant()` fits an additive precursor-plus-sample model per protein, maximising a
posterior that scores observed values by squared residual and missing values by their DPC
probability. The returned `$other$standard.error` is why this project uses limpa: `dpcDE()`
reads it as a precision weight, so a protein reconstructed largely from missing precursors
gets a value and a wide one. Handing `$E` to `lmFit()` discards that and treats every
protein as equally reliable.

```{r quantify}
#| cache: true
#| cache.extra: !expr tools::md5sum(here::here("01_Preprocess", "01_Filtering", "c_data", "precursors_filtered.rds"))
elapsed <- system.time(
  proteins <- dpcQuant(y, "Protein.Group", dpc = dpcest)
)
```

`r format(nrow(proteins), big.mark = ",")` protein groups from
`r format(nrow(y), big.mark = ",")` precursors in
`r round(elapsed[["elapsed"]] / 60, 1)` minutes. The matrix has no missing values; what
varies is how certain each value is.

## Detection filter

`filterByDetection()`'s help fixes the order: it runs after `dpcQuant()` and before DE, and
it returns a logical vector rather than an object. For a small experiment the help suggests
setting `n.samples` to the smallest group size. Here that is `BFR_T2` with 32 samples, one
short because nobody acquired S06's left leg at T2.

The function does not read `group`. It counts detections across all samples and keeps a
protein when that count clears the threshold, so the smallest group size sets a
cohort-wide bar rather than guaranteeing detection within any one group.

```{r detection}
group_sizes <- table(proteins$targets$group)
keep <- filterByDetection(proteins, n.samples = min(group_sizes))
proteins <- proteins[keep, ]

group_sizes
c(before = length(keep), kept = sum(keep), dropped = sum(!keep))
```

## Normalisation

limpa places normalisation after `dpcQuant()` and before `dpcDE()` and performs none itself, because
DIA-NN has usually normalised the precursors already. What matters is which kind is needed. Comparing
a sample against the average of all samples gives a difference per protein; fitting that against
abundance gives a curve. A flat curve means the sample disagrees by a constant, which a per-sample
shift corrects. A curve that travels means the disagreement depends on abundance, and only cyclic
loess corrects that.

Here the curve travels 0.45 log2 in the median sample against 0.13 for the spread in per-sample
medians, and the curvature is unrelated to treatment or timepoint.

```{r normalise}
ma_width <- function(E) {
  a <- rowMeans(E)
  apply(E, 2, \(x) diff(range(lowess(a, x - a, f = 0.5)$y)))
}

before <- median(ma_width(proteins$E))
proteins$E <- normalizeBetweenArrays(proteins$E, method = "cyclicloess")

c(before = round(before, 3), after = round(median(ma_width(proteins$E)), 3))
```

`proteins.rds` is written after this, so it is the matrix the model is fitted to.

## What survives

```{r quality}
protein_quality <- tibble(
  protein = rownames(proteins),
  gene = proteins$genes$Genes,
  n_precursors = proteins$genes$NPrec,
  prop_observed = proteins$genes$PropObs,
  mean_log2 = rowMeans(proteins$E),
  mean_standard_error = rowMeans(proteins$other$standard.error),
  mean_n_observations = rowMeans(proteins$other$n.observations)
)
summary(protein_quality[c("mean_log2", "mean_standard_error", "prop_observed")])
```

`NPrec` is what the installed version writes; the help page calls it `NPeptides`. A protein
with a high mean standard error and few observations is one whose estimate leans on the DPC
rather than on data.

```{r mds}
#| fig-height: 5
plotMDSUsingSEs(proteins,
  labels = proteins$targets$group,
  col = as.integer(factor(proteins$targets$group))
)
```

Samples are not expected to separate by group. Every participant contributes all four cells,
so between-participant variation dominates the leading dimensions and the participant term
in the design is what removes it. A sample sitting far from everything else is worth
investigating; a lack of group structure is not.

## Outputs

`proteins.rds` is read by both stages of `02_Differential_Expression`.

```{r write}
saveRDS(proteins, here("01_Preprocess", "02_Quantification", "c_data", "proteins.rds"))
write_csv(
  dpc_parameters,
  here("01_Preprocess", "02_Quantification", "c_data", "dpc_parameters.csv")
)
write_csv(
  protein_quality,
  here("01_Preprocess", "02_Quantification", "c_data", "protein_quality.csv")
)
```

```{r session}
sessionInfo()
```
