---
title: "02 · Quantification"
subtitle: "Detection probability curve, protein roll-up, detection filter, normalisation"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    df-print: kable
execute:
  message: false
---

Half of a DIA matrix is empty, and not at random: faint precursors go missing more often
than bright ones. `dpc()` fits that as a curve, and `dpcQuant()` uses it to give every
protein a value in every sample without imputing anything. This stage is the limpa
vignette's own sequence, unaltered.

```{r setup}
library(here)
library(limpa)
library(limma)
library(dplyr)
library(readr)
library(tibble)

y <- readRDS(here("01_Preprocess", "01_Filtering", "c_data", "precursors_filtered.rds"))
dim(y)
```

## The missingness relationship

```{r ave-vs-mis}
plotAveVsMis(y)
```

A downward trend means detection depends on intensity and the DPC has something to fit. A
flat cloud would mean missingness is closer to random, where limpa still works but gains
less over simpler methods.

## The curve

```{r dpc}
dpcest <- dpc(y$E)
plotDPC(dpcest)

dpc_parameters <- tibble(
  intercept = dpcest$dpc[1],
  slope = dpcest$dpc[2]
)
round(dpc_parameters, 3)
```

`dpc()` defaults to `model = "cn"` and its help says it then returns `dpcCN()`, the
complete-normal estimator the limpa FAQ recommends for noisy data. Its `subset = 2000`
default fits on a systematic sample stratified by missingness and mean intensity rather
than all `r format(nrow(y), big.mark = ",")` rows. limpa documents `robust` as *"Only used
if `model='on'"*, so it does nothing here and the subsample supplies the robustness.

The fit comes out at 0.488, below the 0.7-to-0.9 band the FAQ calls typical for DIA-NN searched
with match-between-runs, though inside the 0.1-to-1.0 range it calls usable. The vignette notes
the slope comes out low when peptides vary a lot, which is what a within-subject design with
heavy missingness looks like.

Quantification therefore uses a preset 0.7, which is the rule limpa's FAQ gives for this exact
situation: "If `dpcCN()` doesn't give a value in that range, it would be reasonable to run
`dpcQuant()` with a preset value `dpc.slope=0.7`." The fitted value is reported as a sensitivity
check below. The choice follows the documented rule, not any downstream count.

One caveat belongs on the record. The FAQ says a slope set too low makes the analysis more
conservative. Here the reverse happened: the lower fitted slope produced smaller standard errors
and more hits than the preset. The preset remains both the documented setting and the cautious one
on this data, but the direction is not what the documentation predicts.

::: {.callout-note}
The visual fit is not the diagnostic. `plotDPC()` draws a curve in *unobserved* intensity
against proportions computed on *observed* intensity, so some discrepancy is structural.
Judge the slope.
:::

## Roll-up

`dpcQuant()` fits an additive precursor-plus-sample model per protein, maximising a
posterior that scores observed values by squared residual and missing values by their DPC
probability. The returned `$other$standard.error` is why this project uses limpa: `dpcDE()`
reads it as a precision weight, so a protein reconstructed largely from missing precursors
gets a value and a wide one. Handing `$E` to `lmFit()` discards that and treats every
protein as equally reliable.

```{r quantify}
#| cache: true
#| cache.extra: !expr tools::md5sum(here::here("01_Preprocess", "01_Filtering", "c_data", "precursors_filtered.rds"))
elapsed <- system.time(
  proteins <- dpcQuant(y, "Protein.Group", dpc.slope = 0.7)
)
```

`r format(nrow(proteins), big.mark = ",")` protein groups from
`r format(nrow(y), big.mark = ",")` precursors in
`r round(elapsed[["elapsed"]] / 60, 1)` minutes. The matrix has no missing values; what
varies is how certain each value is.

## Detection filter

`filterByDetection()`'s help fixes the order: it runs after `dpcQuant()` and before DE, and
it returns a logical vector rather than an object. For a small experiment the help suggests
setting `n.samples` to the smallest group size. Here that is `BFR_T2` with 32 samples, one
short because nobody acquired S06's left leg at T2.

The function does not read `group`. It counts detections across all samples and keeps a
protein when that count clears the threshold, so the smallest group size sets a
cohort-wide bar rather than guaranteeing detection within any one group.

```{r detection}
group_sizes <- table(proteins$targets$group)
keep <- filterByDetection(proteins, n.samples = min(group_sizes))
proteins <- proteins[keep, ]

group_sizes
c(before = length(keep), kept = sum(keep), dropped = sum(!keep))
```

## Normalisation

limpa places normalisation after `dpcQuant()` and before `dpcDE()` and performs none itself, because
DIA-NN has usually normalised the precursors already. What matters is which kind is needed. Comparing
a sample against the average of all samples gives a difference per protein; fitting that against
abundance gives a curve. A flat curve means the sample disagrees by a constant, which a per-sample
shift corrects. A curve that travels means the disagreement depends on abundance, and only cyclic
loess corrects that.

Here the curve travels 0.45 log2 in the median sample against 0.13 for the spread in per-sample
medians, and the curvature is unrelated to treatment or timepoint.

```{r normalise}
ma_width <- function(E) {
  a <- rowMeans(E)
  apply(E, 2, \(x) diff(range(lowess(a, x - a, f = 0.5)$y)))
}

before <- median(ma_width(proteins$E))
proteins$E <- normalizeBetweenArrays(proteins$E, method = "cyclicloess")

c(before = round(before, 3), after = round(median(ma_width(proteins$E)), 3))
```

`proteins.rds` is written after this, so it is the matrix the model is fitted to.

## What survives

```{r quality}
protein_quality <- tibble(
  protein = rownames(proteins),
  gene = proteins$genes$Genes,
  n_precursors = proteins$genes$NPrec,
  prop_observed = proteins$genes$PropObs,
  mean_log2 = rowMeans(proteins$E),
  mean_standard_error = rowMeans(proteins$other$standard.error),
  mean_n_observations = rowMeans(proteins$other$n.observations)
)
summary(protein_quality[c("mean_log2", "mean_standard_error", "prop_observed")])
```

`NPrec` is what the installed version writes; the help page calls it `NPeptides`. A protein
with a high mean standard error and few observations is one whose estimate leans on the DPC
rather than on data.

```{r mds}
#| fig-height: 5
plotMDSUsingSEs(proteins,
  pch = 16,
  col = as.integer(factor(proteins$targets$participant))
)
```

One colour per participant, because that is what structures the plot: participant accounts for
52% of the leading dimension against 3% for group. Every participant contributes all four cells,
so between-participant variation dominates and the participant term in the design is what removes
it. A sample sitting far from its own participant is worth investigating; an absence of group
structure is not.

## Slope sensitivity

Quantification is repeated at the fitted 0.488 and compared against the primary run, with the
same detection filter and normalisation applied so the slope is the only difference.

Both slopes keep the same 3,042 proteins. The preset raises the mean standard error from 0.53 to
0.66, so it is the more conservative setting: `dpcDE()` reads those errors as precision weights,
and larger errors mean less confidence, not more.

The median absolute difference in protein abundance is printed below. It measures how far values
move, not whether that movement survives a contrast. Every comparison here is within one
participant, so a shift common to a participant's samples cancels in the subtraction.

```{r slope-fitted}
#| cache: true
#| cache.extra: !expr tools::md5sum(here::here("01_Preprocess", "01_Filtering", "c_data", "precursors_filtered.rds"))
fitted <- dpcQuant(y, "Protein.Group", dpc = dpcest)
```

```{r slope-sensitivity}
fitted <- fitted[filterByDetection(fitted, n.samples = min(group_sizes)), ]
fitted$E <- normalizeBetweenArrays(fitted$E, method = "cyclicloess")

shared <- intersect(rownames(proteins), rownames(fitted))
tibble(
  slope = c(0.7, round(dpcest$dpc[2], 3)),
  role = c("primary", "sensitivity"),
  n_proteins = c(nrow(proteins), nrow(fitted)),
  n_shared = length(shared),
  mean_standard_error = round(c(mean(proteins$other$standard.error[shared, ]),
                                mean(fitted$other$standard.error[shared, ])), 3)
)

round(median(abs(proteins$E[shared, ] - fitted$E[shared, ])), 4)
```

## Outputs

`proteins.rds` is read by both stages of `02_Differential_Expression`.

```{r write}
saveRDS(proteins, here("01_Preprocess", "02_Quantification", "c_data", "proteins.rds"))
write_csv(
  dpc_parameters,
  here("01_Preprocess", "02_Quantification", "c_data", "dpc_parameters.csv")
)
write_csv(
  protein_quality,
  here("01_Preprocess", "02_Quantification", "c_data", "protein_quality.csv")
)
```

```{r session}
sessionInfo()
```
