---
title: "02 · Quantification"
subtitle: "Detection probability curve, protein roll-up, detection filter"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    df-print: kable
execute:
  message: false
---

Half of a DIA matrix is empty, and not at random: faint precursors go missing more often
than bright ones. `dpc()` fits that as a curve, and `dpcQuant()` uses it to give every
protein a value in every sample without imputing anything. This stage is the limpa
vignette's own sequence, unaltered.

```{r setup}
pacman::p_load(here, limpa, dplyr, readr, tibble)
set.seed(20260904)

y <- readRDS(here("01_Preprocess", "01_Filtering", "c_data", "precursors_filtered.rds"))
dim(y)
```

## The missingness relationship

```{r ave-vs-mis}
plotAveVsMis(y)
```

A downward trend means detection depends on intensity and the DPC has something to fit. A
flat cloud would mean missingness is closer to random, where limpa still works but gains
less over simpler methods.

## The curve

```{r dpc}
dpcest <- dpc(y$E)
plotDPC(dpcest)

dpc_parameters <- tibble(
  parameter = c("intercept (b0)", "slope (b1)"),
  estimate = as.numeric(dpcest$dpc)
)
stopifnot(is.finite(dpcest$dpc[2]), dpcest$dpc[2] > 0)
dpc_parameters
```

`dpc()` defaults to `model = "cn"` and its help says it then returns `dpcCN()`, the
complete-normal estimator the limpa FAQ recommends for noisy data. Its `subset = 2000`
default fits on a systematic sample stratified by missingness and mean intensity rather
than all `r format(nrow(y), big.mark = ",")` rows. `robust` is documented as *"Only used
if `model='on'"*, so it is inactive here and robustness comes from the subsample.

The slope is reported as fitted and never replaced by a preset. The FAQ puts the acceptable
range at 0.1–1.0 and calls 0.7–0.9 typical for DIA-NN searched with match-between-runs; the
vignette adds that the slope is under-estimated when per-peptide variability is large,
which is what a within-subject design with heavy missingness looks like. limpa's own
published DIA-NN case study fitted 0.59 and used it uncorrected. A shallow slope also
recovers *less* from missing values, so it is the conservative direction to be wrong in.

::: {.callout-note}
The visual fit is not the diagnostic. `plotDPC()` draws a curve in *unobserved* intensity
against proportions computed on *observed* intensity, so some discrepancy is structural.
Judge the slope.
:::

## Roll-up

`dpcQuant()` fits an additive precursor-plus-sample model per protein, maximising a
posterior that scores observed values by squared residual and missing values by their DPC
probability. The returned `$other$standard.error` is why this project uses limpa: `dpcDE()`
reads it as a precision weight, so a protein reconstructed largely from missing precursors
gets a value and a wide one. Handing `$E` to `lmFit()` discards that and treats every
protein as equally reliable.

```{r quantify}
#| cache: true
elapsed <- system.time(
  proteins <- dpcQuant(y, "Protein.Group", dpc = dpcest)
)
```

`r format(nrow(proteins), big.mark = ",")` protein groups from
`r format(nrow(y), big.mark = ",")` precursors in
`r round(elapsed[["elapsed"]] / 60, 1)` minutes. The matrix has no missing values; what
varies is how certain each value is.

## Detection filter

`filterByDetection()`'s help fixes the order: it runs after `dpcQuant()` and before DE, and
it returns a logical vector rather than an object. For a small experiment the help suggests
setting `n.samples` to the smallest group size. Here that is `BFR_T2` with 32 samples, one
short because nobody acquired S06's left leg at T2.

```{r detection}
group_sizes <- table(proteins$targets$group)
keep <- filterByDetection(proteins, n.samples = min(group_sizes))
proteins <- proteins[keep, ]

group_sizes
c(before = length(keep), kept = sum(keep), dropped = sum(!keep))
```

## What survives

```{r quality}
protein_quality <- tibble(
  protein = rownames(proteins),
  gene = proteins$genes$Genes,
  n_precursors = proteins$genes$NPrec,
  prop_observed = proteins$genes$PropObs,
  mean_log2 = rowMeans(proteins$E),
  mean_standard_error = rowMeans(proteins$other$standard.error),
  mean_n_observations = rowMeans(proteins$other$n.observations)
)
stopifnot(!anyNA(protein_quality))
summary(protein_quality[c("mean_log2", "mean_standard_error", "prop_observed")])
```

`NPrec` is what the installed version writes; the help page calls it `NPeptides`. A protein
with a high mean standard error and few observations is one whose estimate leans on the DPC
rather than on data.

```{r mds}
#| fig-height: 5
plotMDSUsingSEs(proteins,
  labels = proteins$targets$group,
  col = as.integer(factor(proteins$targets$group))
)
```

Samples are not expected to separate by group. Every participant contributes all four cells,
so between-participant variation dominates the leading dimensions and the participant term
in the design is what removes it. A sample sitting far from everything else is worth
investigating; a lack of group structure is not.

## Outputs

`proteins.rds` is read by both stages of `02_Differential_Expression`.

```{r write}
saveRDS(proteins, here("01_Preprocess", "02_Quantification", "c_data", "proteins.rds"))
write_csv(
  dpc_parameters,
  here("01_Preprocess", "02_Quantification", "c_data", "dpc_parameters.csv")
)
write_csv(
  protein_quality,
  here("01_Preprocess", "02_Quantification", "c_data", "protein_quality.csv")
)
```

```{r session}
sessionInfo()
```
