---
title: "02 · Quantification"
subtitle: "Detection probability curve, protein roll-up, detection filter, normalisation"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    df-print: kable
    fig-width: 7
    fig-height: 5
    fig-dpi: 150
    fig-align: center
execute:
  message: false
---

Half of a DIA matrix is empty, and fainter precursors go missing more often. `dpc()` fits that as
a curve, and `dpcQuant()` uses it to give every protein a value in every sample without imputing
anything. This stage follows limpa's vignette sequence.

```{r setup}
pacman::p_load(here, limpa, limma, purrr, tibble, writexl)

y <- readRDS(here("01_Preprocess", "01_Filtering", "c_data", "precursors_filtered.rds"))
dim(y)

checkpoints <- here("01_Preprocess", "02_Quantification", "c_data", "quant_runs")
primary_file <- file.path(checkpoints, "proteins_slope_0.7.rds")
sensitivity_file <- file.path(checkpoints, "proteins_slope_fitted.rds")
```

## Missingness

```{r ave-vs-mis}
plotAveVsMis(y)
```

A downward trend means detection depends on intensity and the DPC has something to fit. A
flat cloud would mean missingness is closer to random, where limpa still works but gains
less over simpler methods.

## The curve

```{r dpc}
dpcest <- dpc(y$E)
plotDPC(dpcest)

dpc_parameters <- tibble(
  intercept = dpcest$dpc[1],
  slope = dpcest$dpc[2]
)
round(dpc_parameters, 3)
```

`dpc()` defaults to `model = "cn"` and its help says it then returns `dpcCN()`, the
complete-normal estimator the limpa FAQ recommends for noisy data. Its `subset = 2000`
default fits on a systematic sample stratified by missingness and mean intensity rather
than all `r format(nrow(y), big.mark = ",")` rows. limpa documents `robust` as *"Only used
if `model='on'"*, so it does nothing here and the subsample supplies the robustness.

The fit comes out at `r round(dpcest$dpc[2], 3)`, below the 0.7-to-0.9 band the FAQ calls typical
for DIA-NN searched with match-between-runs, though inside the 0.1-to-1.0 range it calls usable.
The vignette notes the slope comes out low when peptides vary a lot, which a within-subject design
with heavy missingness might produce.

Quantification therefore uses the preset 0.7 that limpa's FAQ suggests: "If `dpcCN()` doesn't give
a value in that range, it would be reasonable to run `dpcQuant()` with a preset value
`dpc.slope=0.7`." The fitted value is carried into a sensitivity run at the end.

The FAQ says a slope set too low makes the analysis more conservative. On this data the reverse
held when it was last measured: the lower fitted slope gave smaller standard errors and more hits
than the preset. The preset remains the documented setting and, here, the cautious one.

::: {.callout-note}
The visual fit is not diagnostic. `plotDPC()` draws a curve in *unobserved* intensity
against proportions computed on *observed* intensity, so some discrepancy is structural.
Judge the slope.
:::

## Roll-up

`dpcQuant()` fits an additive precursor-plus-sample model per protein, maximising a
posterior that scores observed values by squared residual and missing values by their DPC
probability. The returned `$other$standard.error` is why this project uses limpa: `dpcDE()`
reads it as a precision weight, so a protein reconstructed largely from missing precursors
gets a wide standard error. Handing `$E` to `lmFit()` discards that and treats every protein as
equally reliable.

That takes about 100 minutes, so it lives in `a_script/02_quantify_run.R` rather than here.
This notebook loads what the script wrote, so a render doesn't begin a long run by accident.
Rerun the script by hand when the precursor matrix changes.

```{r quantify}
proteins <- readRDS(primary_file)
```

`r format(nrow(proteins), big.mark = ",")` protein groups from
`r format(nrow(y), big.mark = ",")` precursors. The matrix has no missing values; what varies
is each value's standard error.

## Detection filter

`filterByDetection()` runs after `dpcQuant()` and before DE, and returns a logical vector.
For a small experiment the help suggests setting `n.samples` to the smallest group size.
Here that is `BFR_T2` with 32 samples, one short because S06_L_T2 is missing.

The function counts detections across all samples and keeps a protein when that count clears 
the threshold, so the smallest group size sets a cohort-wide bar rather than guaranteeing 
detection within any one group.

```{r detection}
group_sizes <- table(proteins$targets$group)
keep <- filterByDetection(proteins, n.samples = min(group_sizes))
proteins <- proteins[keep, ]

group_sizes
c(before = length(keep), kept = sum(keep), dropped = sum(!keep))
```

Measured on an earlier run, the dropped proteins had a mean SE of 1.56 against 0.66 for 
those kept, and 0.15 detected precursors per sample against 7.48, so the typical sample 
measured nothing at all for them. Relaxing the threshold to limpa's default of 3 
recovered 371 of them, of which exactly one reached BH < 0.05 in any contrast, while the 
larger test set tightened the BH threshold enough to push 14 and 29 previously significant 
proteins back above 0.05 on the two training contrasts.

## Normalisation

limpa places normalisation after `dpcQuant()` and before `dpcDE()` and performs none itself,
because DIA-NN has usually normalised the precursors already. Normalised precursors do not make a
normalised protein roll-up.

A single per-sample shift fixes the spread in per-sample medians but not the MA curve width, which
cyclic loess may help with. Comparing a sample against the average of all samples gives a
difference per protein, and fitting that against abundance gives a curve. A flat curve means the
sample disagrees by a constant. A curve that travels means the disagreement depends on abundance.
Both are reported before and after.

```{r normalise}
ma_width <- function(m) {
  a <- rowMeans(m)
  apply(m, 2, \(x) diff(range(lowess(a, x - a, f = 0.5)$y)))
}
spread <- function(m) diff(range(apply(m, 2, median)))

before <- c(ma_width = median(ma_width(proteins$E)), median_spread = spread(proteins$E))
unnormalised <- proteins$E
proteins$E <- normalizeBetweenArrays(proteins$E, method = "cyclicloess")
after <- c(ma_width = median(ma_width(proteins$E)), median_spread = spread(proteins$E))

round(rbind(before, after), 3)
```

The top row shows every sample's intensity density before and after. The bottom row shows the
sample whose curve travelled furthest before normalisation, against the average of all samples.

```{r normalise-plots}
#| fig-height: 8
worst <- which.max(ma_width(unnormalised))
par(mfrow = c(2, 2))
plotDensities(unnormalised, legend = FALSE, main = "Before cyclic loess")
plotDensities(proteins$E, legend = FALSE, main = "After cyclic loess")
plotMD(unnormalised, column = worst, main = paste(colnames(unnormalised)[worst], "before"))
plotMD(proteins$E, column = worst, main = paste(colnames(proteins$E)[worst], "after"))
```

The curvature is unrelated to treatment or timepoint, so cyclic loess is not absorbing the effect
being tested. `proteins.rds` is written after this step and is the matrix the model is fitted to.

::: {.callout-warning}
## The standard errors stay on the old scale

Cyclic loess remaps `$E` non-linearly and leaves `$other$standard.error` untouched. After
normalisation the stored errors describe the unnormalised scale, and `dpcDE()` reads them as though
they described the new one. Smyth recommends quantile or cyclic-loess normalisation of limpa
protein quantifications when sample differences are technical
([Bioconductor support, 2026](https://support.bioconductor.org/p/9163187/#9163221)).
:::

## What survives

```{r quality}
protein_quality <- tibble(
  protein = rownames(proteins),
  gene = proteins$genes$Genes,
  n_precursors = proteins$genes$NPrec,
  prop_observed = proteins$genes$PropObs,
  mean_log2 = rowMeans(proteins$E),
  mean_standard_error = rowMeans(proteins$other$standard.error),
  mean_n_observations = rowMeans(proteins$other$n.observations)
)
summary(protein_quality[c("mean_log2", "mean_standard_error", "prop_observed")])
```

`NPrec` is what the installed version writes; the help page calls it `NPeptides`. A protein
with a high mean standard error and few observations is one whose estimate is drawn from DPC.

Points are coloured by participant, and the variance each factor explains in the leading dimension
is below the plot. 32 of 33 participants contribute all four cells, so between-participant
variation dominates; the participant term in the design removes it. Check any sample sitting far
from its own participant. No group structure is expected here.

```{r mds}
#| fig-height: 5
mds <- plotMDSUsingSEs(proteins,
  pch = 16,
  col = as.integer(factor(proteins$targets$participant))
)

r2 <- function(f) summary(lm(mds$x ~ factor(f)))$r.squared
round(c(
  participant = r2(proteins$targets$participant),
  group = r2(proteins$targets$group)
), 3)
```

## Slope sensitivity

Quantification is repeated against the fitted curve and compared with the primary run, with the
same detection filter and normalisation applied to both. The primary run passes `dpc.slope = 0.7` 
and lets limpa fit the rest, while the sensitivity run hands `dpcQuant()` the whole curve `dpc()` 
estimated, intercept included. `dpcDE()` reads the standard errors as precision weights, so
whichever run carries the larger mean standard error is the more conservative setting.

The median absolute difference in protein abundance follows the table. It measures how far values
move, not whether that movement survives a contrast: every comparison here is within one
participant, so a shift common to a participant's samples cancels out.

```{r slope-sensitivity}
#| eval: !expr file.exists(sensitivity_file)
sensitivity <- readRDS(sensitivity_file)
sensitivity <- sensitivity[filterByDetection(sensitivity, n.samples = min(group_sizes)), ]
sensitivity$E <- normalizeBetweenArrays(sensitivity$E, method = "cyclicloess")

shared <- intersect(rownames(proteins), rownames(sensitivity))
tibble(
  slope = c(0.7, round(dpcest$dpc[2], 3)),
  role = c("primary", "sensitivity"),
  n_proteins = c(nrow(proteins), nrow(sensitivity)),
  n_shared = length(shared),
  mean_standard_error = round(c(
    mean(proteins$other$standard.error[shared, ]),
    mean(sensitivity$other$standard.error[shared, ])
  ), 3)
)

round(median(abs(proteins$E[shared, ] - sensitivity$E[shared, ])), 4)
```

This section renders only when `c_data/quant_runs/proteins_slope_fitted.rds` is present, since
computing it costs about another 100 minutes. To produce it once:

```sh
Rscript 01_Preprocess/02_Quantification/a_script/02_quantify_run.R --sensitivity
```

After that it loads as the primary run does.

Refitting the contrasts under each slope is separate and not done here; the counts produced are 
recorded in this stage's README.

## Outputs

`proteins.rds` is read by both stages of `02_Differential_Expression`.

```{r write}
sheets <- list(dpc_parameters = dpc_parameters, protein_quality = protein_quality)
overview <- tibble(
  sheet = names(sheets),
  rows = map_int(sheets, nrow),
  columns = map_int(sheets, ncol),
  description = c(
    "Intercept and slope of the fitted detection probability curve",
    "Per protein: precursors, observed proportion, mean abundance, standard error, observations"
  )
)

out <- here("01_Preprocess", "02_Quantification", "c_data")
saveRDS(proteins, file.path(out, "proteins.rds"))
write_xlsx(c(list(overview = overview), sheets), file.path(out, "02_quantify.xlsx"))
```

```{r session}
sessionInfo()
```
