---
title: "Sensitivity: fixed detection-curve slope"
subtitle: "Quantification refitted at dpc.slope = 0.7"
format:
  html:
    toc: true
    embed-resources: true
    df-print: kable
execute:
  message: false
---

The fitted slope sits below the 0.7 to 0.9 band limpa reports as typical for DIA-NN searched with
match-between-runs. This refits quantification with the slope held at 0.7 and measures how far the
protein estimates move. Not part of the pipeline; it keeps its own cache and is run on demand.

```{r setup}
library(here)
library(limpa)
library(limma)
library(tibble)

y <- readRDS(here("01_Preprocess", "01_Filtering", "c_data", "precursors_filtered.rds"))
fitted_slope <- readRDS(here("01_Preprocess", "02_Quantification", "c_data", "proteins.rds"))
```

```{r quantify}
#| cache: true
slope_07 <- dpcQuant(y, "Protein.Group", dpc.slope = 0.7)
slope_07 <- slope_07[filterByDetection(
  slope_07, n.samples = min(table(slope_07$targets$group))
), ]
slope_07$E <- normalizeBetweenArrays(slope_07$E, method = "cyclicloess")
```

Normalised the same way as the main run, so the two matrices are comparable.

```{r compare}
shared <- intersect(rownames(fitted_slope), rownames(slope_07))

tibble(
  proteins_fitted = nrow(fitted_slope),
  proteins_slope_07 = nrow(slope_07),
  shared = length(shared),
  median_abs_difference = round(median(abs(fitted_slope$E[shared, ] - slope_07$E[shared, ])), 4),
  mean_se_fitted = round(mean(fitted_slope$other$standard.error), 4),
  mean_se_slope_07 = round(mean(slope_07$other$standard.error), 4)
)
```

A small median difference means the quantification is insensitive to the slope, which is what limpa
reports for a range of slope settings.

```{r session}
sessionInfo()
```
