---
title: "03 · Protein against phenotype"
subtitle: "Every protein correlated with the ultrasound outcomes"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    df-print: kable
    fig-width: 7
    fig-height: 5
    fig-dpi: 150
    fig-align: center
execute:
  message: false
---

`02_Differential` asks which proteins changed between groups. This asks whether a protein's change
tracks how much the muscle changed, which a contrast cannot answer: it compares group means and
carries no per-leg outcome.

Pooled correlates each protein's pre-to-post change with the phenotype change across all 65 legs.
Differential takes the 32 participants with both legs and correlates the BFR-minus-high-load
difference in protein with the same difference in phenotype, removing every between-participant
confound.

```{r setup}
pacman::p_load(here, dplyr, tidyr, purrr, tibble, ggplot2, readr, readxl, writexl)

proteins <- readRDS(here("01_Preprocess", "02_Quantification", "c_data", "proteins.rds"))
phenotype <- read_csv(here("00_Input", "phenotype.csv"), show_col_types = FALSE)
```

`r format(nrow(proteins), big.mark = ",")` proteins give `r round(nrow(proteins) * 0.05)` nominal
hits at p < 0.05 by chance, so no single protein is read at nominal p. BH within each analysis and
outcome is the reported quantity.

## Pairing

A leg needs both timepoints for a change score. S06's left leg has T1 only, so it drops here and
nowhere else.

```{r pairs}
legs <- proteins$targets |>
  mutate(leg_id = paste(participant, leg, sep = "_")) |>
  select(leg_id, participant, leg, treatment, timepoint, sample_id) |>
  pivot_wider(names_from = timepoint, values_from = sample_id) |>
  drop_na(T1, T2) |>
  arrange(participant, treatment)
leg_pairs <- legs |>
  select(participant, treatment, leg_id) |>
  pivot_wider(names_from = treatment, values_from = leg_id) |>
  drop_na(BFR, HLRT)

delta <- proteins$E[, legs$T2] - proteins$E[, legs$T1]
colnames(delta) <- legs$leg_id
paired <- delta[, leg_pairs$BFR] - delta[, leg_pairs$HLRT]
colnames(paired) <- leg_pairs$participant
c(legs = ncol(delta), participants = ncol(paired), proteins = nrow(delta))
```

## Outcomes

Four ultrasound measures, each as a post-minus-pre change. Cross-sectional area measures size; echo
intensity measures quality and falls as composition improves.

```{r outcomes}
leg_phenotype <- legs |>
  left_join(select(phenotype, -treatment), by = c("participant", "leg")) |>
  mutate(
    vl_csa = vl_csa_post_cm2 - vl_csa_pre_cm2,
    vl_echo = vl_echo_post_au - vl_echo_pre_au,
    rf_csa = rf_csa_post_cm2 - rf_csa_pre_cm2,
    rf_echo = rf_echo_post_au - rf_echo_pre_au
  )
outcomes <- c("vl_csa", "vl_echo", "rf_csa", "rf_echo")
paired_outcome <- map(set_names(outcomes), \(name) {
  value <- set_names(leg_phenotype[[name]], leg_phenotype$leg_id)
  value[leg_pairs$BFR] - value[leg_pairs$HLRT]
})
```

## Correlations

Neither change is expected to be normal, so every correlation is Spearman. Ties make `cor.test` fall
back to its approximation and warn on every call, so the 24,336 warnings are suppressed.

```{r correlate}
spearman_by_row <- function(values, outcome) {
  fits <- suppressWarnings(apply(values, 1, \(row) {
    test <- cor.test(row, outcome, method = "spearman")
    c(r = unname(test$estimate), p = test$p.value)
  }))
  tibble(
    protein = rownames(values), n = sum(!is.na(outcome)),
    r = unname(fits["r", ]), p = unname(fits["p", ])
  )
}
results <- map(set_names(outcomes), \(name) {
  bind_rows(
    pooled = spearman_by_row(delta, leg_phenotype[[name]]),
    differential = spearman_by_row(paired, paired_outcome[[name]]),
    .id = "analysis"
  )
}) |>
  list_rbind(names_to = "outcome") |>
  mutate(fdr = p.adjust(p, "BH"), .by = c(analysis, outcome)) |>
  mutate(gene = proteins$genes$Genes[match(protein, rownames(proteins$genes))]) |>
  relocate(analysis, outcome, protein, gene)
```

## What survives

```{r summary}
summary_table <- results |>
  summarise(
    proteins = n(), nominal = sum(p < 0.05), expected = round(n() * 0.05, 1),
    fdr_sig = sum(fdr < 0.05), best_fdr = signif(min(fdr), 3),
    .by = c(analysis, outcome, n)
  ) |>
  mutate(ratio = round(nominal / expected, 2)) |>
  arrange(analysis, outcome)
summary_table
```

Read `ratio` first: near 1, the nominal count is what `r format(nrow(proteins), big.mark = ",")`
tests return when nothing is there, whatever the smallest p.

```{r figure}
#| fig-height: 4
ggplot(summary_table, aes(ratio, outcome, fill = analysis)) +
  geom_vline(xintercept = 1, linewidth = 0.4, colour = "grey40") +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  scale_fill_manual(values = c(pooled = "#B2182B", differential = "#2166AC"), name = NULL) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(x = "nominal hits / chance expectation", y = NULL) +
  theme_minimal(base_size = 10) +
  theme(panel.grid.major.y = element_blank(), legend.position = "top")
```

## Training hits among the leads

A training hit is a protein at BH < 0.05 in either Post-Pre contrast of `02_Differential`. The table
counts training hits at nominal p per outcome against the count expected by chance, with a one-sided
Fisher test.

```{r training-overlap}
dep_matrix <- read_excel(
  here("02_Differential_Expression", "02_Differential", "c_data", "02_differential.xlsx"),
  "DEP_matrix"
)
training_hits <- dep_matrix |>
  filter(`BFR_Post-Pre_adj.P.Val` < 0.05 | `HLRT_Post-Pre_adj.P.Val` < 0.05) |>
  pull(protein)
training_overlap <- results |>
  mutate(training = protein %in% training_hits) |>
  summarise(
    training_hits = sum(training),
    nominal = sum(training & p < 0.05),
    expected = round(sum(training) * mean(p < 0.05), 1),
    fisher_p = signif(
      fisher.test(table(p < 0.05, training), alternative = "greater")$p.value, 2
    ),
    .by = c(analysis, outcome)
  ) |>
  arrange(analysis, outcome)
training_overlap
```

Both pooled CSA outcomes hold about twice the expected count; the other six sit at chance. Proteins
that changed with training vary more across legs, which gives them more power in a pooled
correlation.

```{r write}
sheets <- list(
  summary = summary_table,
  training_overlap = training_overlap,
  correlations = arrange(results, fdr)
)
overview <- data.frame(
  sheet = names(sheets), rows = map_int(sheets, nrow), columns = map_int(sheets, ncol),
  description = c(
    "Hits against chance per outcome",
    "Training hits at nominal p against chance",
    "Spearman r, p and FDR per protein"
  )
)
out <- here("02_Differential_Expression", "03_Phenotype", "c_data")
write_xlsx(c(list(overview = overview), sheets), file.path(out, "03_phenotype.xlsx"))
```

```{r session}
sessionInfo()
```
