---
title: "03 · Protein against phenotype"
subtitle: "Every protein correlated with the ultrasound outcomes"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    df-print: kable
    fig-width: 7
    fig-height: 5
    fig-dpi: 150
    fig-align: center
execute:
  message: false
---

`02_Differential` asks which proteins changed between groups. This asks whether a
protein's change tracks how much the muscle changed. The contrast model cannot
answer it, because a contrast compares group means and carries no per-leg outcome.

Two framings. **Pooled** takes all 65 legs and correlates each protein's pre-to-post change with
that leg's phenotype change. **Differential** takes the 32 participants with both legs and asks
whether the BFR-minus-high-load difference in a protein tracks the BFR-minus-high-load difference
in phenotype, which removes every between-participant confound.

3,042 proteins mean 152 nominal hits are expected at p < 0.05, so nothing here is read at nominal
p. BH within each analysis and outcome is the reported quantity.

```{r setup}
library(here)
library(dplyr)
library(tidyr)
library(purrr)
library(tibble)
library(ggplot2)

proteins <- readRDS(here("01_Preprocess/02_Quantification/c_data/proteins.rds"))
phenotype <- readr::read_csv(here("00_Input/phenotype.csv"), show_col_types = FALSE)
out <- here("02_Differential_Expression", "03_Phenotype", "c_data")
dir.create(out, recursive = TRUE, showWarnings = FALSE)
```

## Pairing

A leg contributes only when both timepoints were acquired, which is what makes a change score
definable. S06's left leg has T1 only, so it drops here and nowhere else.

```{r pairs}
targets <- proteins$targets |>
  mutate(leg_id = paste(participant, leg, sep = "_"))
by_timepoint <- targets |>
  select(leg_id, participant, leg, treatment, timepoint, sample_id) |>
  pivot_wider(names_from = timepoint, values_from = sample_id)
legs <- by_timepoint |>
  filter(!is.na(T1), !is.na(T2)) |>
  arrange(participant, treatment)
pairs <- legs |>
  select(participant, treatment, leg_id) |>
  pivot_wider(names_from = treatment, values_from = leg_id) |>
  filter(!is.na(BFR), !is.na(HLRT))

delta <- proteins$E[, legs$T2] - proteins$E[, legs$T1]
colnames(delta) <- legs$leg_id
paired <- delta[, pairs$BFR] - delta[, pairs$HLRT]
colnames(paired) <- pairs$participant
tibble(legs = ncol(delta), participants = ncol(paired), proteins = nrow(delta))
```

## Outcomes

Four ultrasound measures, each as a post-minus-pre change. Cross-sectional area is the size
outcome; echo intensity is the quality one, and falls when muscle composition improves.

```{r outcomes}
outcomes <- c(
  vl_csa = "vl_csa_post_cm2 - vl_csa_pre_cm2",
  vl_echo = "vl_echo_post_au - vl_echo_pre_au",
  rf_csa = "rf_csa_post_cm2 - rf_csa_pre_cm2",
  rf_echo = "rf_echo_post_au - rf_echo_pre_au"
)
leg_phenotype <- legs |>
  left_join(select(phenotype, -treatment), by = c("participant", "leg")) |>
  mutate(!!!set_names(map(outcomes, rlang::parse_expr), names(outcomes)))
paired_outcome <- map(set_names(names(outcomes)), function(name) {
  value <- set_names(leg_phenotype[[name]], leg_phenotype$leg_id)
  value[pairs$BFR] - value[pairs$HLRT]
})
```

## Correlations

Spearman throughout, since neither a protein change nor an echo intensity change is expected to
be normal. Ties make `cor.test` fall back from its exact p to the approximation and warn on every
call, so the warnings are suppressed rather than printed 24,336 times.

```{r correlate}
spearman_by_row <- function(values, outcome) {
  usable <- !is.na(outcome)
  values <- values[, usable, drop = FALSE]
  outcome <- outcome[usable]
  fits <- suppressWarnings(apply(values, 1, \(row) {
    test <- stats::cor.test(row, outcome, method = "spearman")
    c(r = unname(test$estimate), p = test$p.value)
  }))
  tibble(
    protein = rownames(values), n = length(outcome),
    r = unname(fits["r", ]), p = unname(fits["p", ])
  )
}
results <- imap(outcomes, \(expression, name) {
  bind_rows(
    mutate(spearman_by_row(delta, leg_phenotype[[name]]), analysis = "pooled"),
    mutate(spearman_by_row(paired, paired_outcome[[name]]), analysis = "differential")
  ) |>
    mutate(outcome = name)
}) |>
  list_rbind() |>
  mutate(fdr = p.adjust(p, "BH"), .by = c(analysis, outcome)) |>
  left_join(
    rownames_to_column(proteins$genes, "protein") |> select(protein, gene = Genes),
    by = "protein"
  ) |>
  relocate(analysis, outcome, protein, gene)
```

## What survives

```{r summary}
summary_table <- results |>
  summarise(
    proteins = n(), nominal = sum(p < 0.05), expected = round(n() * 0.05, 1),
    fdr_sig = sum(fdr < 0.05), best_fdr = signif(min(fdr), 3),
    .by = c(analysis, outcome, n)
  ) |>
  mutate(ratio = round(nominal / expected, 2)) |>
  arrange(analysis, outcome)
summary_table
```

Read `ratio` first. A value near 1 means the nominal count is what 3,042 tests return when
nothing is there, whatever the smallest p happens to be.

```{r figure}
#| fig-width: 7
#| fig-height: 4
ggplot(summary_table, aes(ratio, outcome, fill = analysis)) +
  geom_vline(xintercept = 1, linewidth = 0.4, colour = "grey40") +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  scale_fill_manual(values = c(pooled = "#B2182B", differential = "#2166AC"), name = NULL) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(x = "nominal hits / chance expectation", y = NULL) +
  theme_minimal(base_size = 10) +
  theme(panel.grid.major.y = element_blank(), legend.position = "top")
```

```{r write}
writexl::write_xlsx(
  list(summary = summary_table, correlations = arrange(results, fdr)),
  file.path(out, "03_phenotype.xlsx")
)
saveRDS(list(results = results, summary = summary_table), file.path(out, "phenotype.rds"),
  compress = "xz"
)
```
