---
title: "01 · Gene sets and protein volcanoes"
subtitle: "Frozen human pathways and the inputs for set testing"
params:
  msigdb_version: "2026.1.Hs"
  collections: [Hallmark, Reactome, GOBP]
  min_set_size: 15
  max_set_size: 500
  min_measured: 15
  jaccard_cutoff: 0.5
  protein_fdr: 0.05
  label_n: 5
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    code-fold: true
    code-tools: true
    df-print: kable
    fig-width: 7
    fig-height: 6.5
    fig-dpi: 150
    fig-align: center
execute:
  message: false
  error: false
  cache: false
---

This stage freezes pathway membership, matches it to the measured proteome,
and prepares the matrix and model for set testing. The volcanoes describe the
**existing protein-level results**. Pathways have not been tested here, so
there are no enrichment arcs or pathway p-values yet.

## Inputs and alignment

```{r setup}
required <- c(
  "here", "limma", "limpa", "dplyr", "tibble", "tidyr", "purrr", "readr",
  "writexl", "digest", "Matrix", "ggplot2", "enrichVolcano", "knitr"
)
missing <- required[!vapply(required, requireNamespace, logical(1),
  quietly = TRUE
)]
if (length(missing)) {
  stop("Install required R packages: ", paste(missing, collapse = ", "))
}
library(here)
library(dplyr)
library(tibble)
library(tidyr)
library(purrr)
library(ggplot2)
here::i_am("03_Pathway_Enrichment/01_Gene_Sets/a_script/01_gene_sets.qmd")

stage <- here("03_Pathway_Enrichment", "01_Gene_Sets")
out <- file.path(stage, "c_data")
cache_dir <- file.path(out, "cache")
figure_dir <- file.path(stage, "b_reports", "figures")
for (path in c(out, cache_dir, figure_dir)) {
  dir.create(path, recursive = TRUE, showWarnings = FALSE)
}
input_paths <- c(
  proteins = here("01_Preprocess/02_Quantification/c_data/proteins.rds"),
  fit = here("02_Differential_Expression/02_Differential/c_data/fit.rds"),
  design = here("02_Differential_Expression/01_Design/c_data/design.rds")
)
if (!all(file.exists(input_paths))) {
  stop("Run the upstream stages first. Missing: ",
    paste(input_paths[!file.exists(input_paths)], collapse = ", "))
}
proteins <- readRDS(input_paths[["proteins"]])
fit <- readRDS(input_paths[["fit"]])
d <- readRDS(input_paths[["design"]])
protein_ids <- rownames(proteins$E)
contrast_names <- colnames(d$contrasts)
training_contrasts <- c("BFR_post_vs_pre", "HLRT_post_vs_pre")
stopifnot(
  !anyDuplicated(protein_ids), !anyDuplicated(colnames(proteins$E)),
  identical(rownames(proteins$genes), protein_ids),
  identical(rownames(fit$coefficients), protein_ids),
  identical(colnames(proteins$E), rownames(d$design)),
  identical(colnames(proteins$E), proteins$targets$sample_id),
  identical(colnames(fit$coefficients), contrast_names),
  identical(fit$design, d$design), identical(fit$contrasts, d$contrasts),
  identical(fit$EList$E, proteins$E),
  identical(fit$EList$other, proteins$other),
  identical(fit$EList$genes, proteins$genes),
  all(is.finite(proteins$E)),
  identical(dimnames(proteins$other$standard.error), dimnames(proteins$E)),
  identical(dimnames(proteins$other$n.observations), dimnames(proteins$E)),
  all(is.finite(proteins$other$standard.error)),
  all(proteins$other$standard.error > 0),
  d$negative_control %in% contrast_names,
  all(training_contrasts %in% contrast_names)
)
if (!"volc_sig_col" %in% names(formals(enrichVolcano::volcano_ring))) {
  stop("enrichVolcano with volcano_ring(volc_sig_col = ...) is required. ",
    "Install the current Dustyn-T-Lewis/enrichVolcano package.")
}
stopifnot(
  is.numeric(params$min_set_size), params$min_set_size >= 1,
  params$min_set_size == as.integer(params$min_set_size),
  params$max_set_size >= params$min_set_size,
  params$min_measured >= 1,
  params$min_measured == as.integer(params$min_measured),
  params$min_measured <= params$max_set_size,
  params$jaccard_cutoff > 0, params$jaccard_cutoff <= 1,
  params$protein_fdr > 0, params$protein_fdr < 1,
  params$label_n >= 0, params$label_n == as.integer(params$label_n)
)
tibble(
  n_proteins = nrow(proteins$E), n_samples = ncol(proteins$E),
  n_participants = n_distinct(proteins$targets$participant),
  n_contrasts = length(contrast_names)
)
```

The matrix is the normalised log2 abundance matrix used by `dpcDE()`. The
handoff also preserves the complete EList, its standard errors and observation
counts, the weighted fit, and the fixed-participant design. This stage fits
no new model. The upstream approximation still applies: cyclic loess changes
abundances while the stored standard errors remain on the pre-normalisation
scale.

## Frozen gene-set cache

The source is human MSigDB, retrieved through
[msigdbr](https://igordot.github.io/msigdbr/reference/msigdbr.html).
The database release is pinned separately from the R package version. A
project-local RDS and checksum are sufficient for subsequent offline renders;
`msigdbr` is needed only when that snapshot is absent. Changing size or
overlap parameters never downloads a different release.

```{r frozen-cache}
collection_specs <- list(
  Hallmark = list(collection = "H", subcollection = NULL),
  Reactome = list(collection = "C2", subcollection = "CP:REACTOME"),
  GOBP = list(collection = "C5", subcollection = "GO:BP"),
  GOCC = list(collection = "C5", subcollection = "GO:CC"),
  GOMF = list(collection = "C5", subcollection = "GO:MF")
)
collections <- unlist(params$collections, use.names = FALSE)
stopifnot(
  length(collections) > 0, !anyDuplicated(collections),
  all(collections %in% names(collection_specs)),
  grepl("^[0-9]{4}\\.[0-9]+\\.Hs$", params$msigdb_version)
)
cache_request <- list(
  schema = 1L, species = "Homo sapiens", db_species = "HS",
  db_version = params$msigdb_version,
  collections = collection_specs[collections]
)
cache_key <- substr(digest::digest(cache_request, algo = "sha256"), 1, 12)
cache_file <- file.path(cache_dir, paste0(
  "msigdb_", params$msigdb_version, "_", cache_key, ".rds"
))
checksum_file <- paste0(cache_file, ".md5")

if (!file.exists(cache_file)) {
  if (!requireNamespace("msigdbr", quietly = TRUE)) {
    stop("The frozen cache is absent. Install msigdbr or restore ",
      "the matching RDS and .md5 files under c_data/cache/.")
  }
  fetched <- imap(collection_specs[collections], function(spec, database) {
    df <- msigdbr::msigdbr(
      db_species = "HS", species = "Homo sapiens",
      collection = spec$collection, subcollection = spec$subcollection
    )
    if (!identical(unique(df$db_version), params$msigdb_version)) {
      stop("Requested MSigDB ", params$msigdb_version,
        " but msigdbr returned ", paste(unique(df$db_version), collapse = ", "),
        ". Restore the requested snapshot or explicitly change the version.")
    }
    df |>
      transmute(
        database, pathway = gs_name, set_id = paste(database, gs_name, sep = "|"),
        gene = gene_symbol, msigdb_id = gs_id, source_id = gs_exact_source,
        description = gs_description, url = gs_url, db_version
      ) |>
      filter(!is.na(gene), nzchar(gene)) |>
      distinct()
  })
  frozen <- list(
    request = cache_request,
    created_utc = format(Sys.time(), tz = "UTC", usetz = TRUE),
    msigdbr_version = as.character(packageVersion("msigdbr")),
    source = "https://www.gsea-msigdb.org/gsea/msigdb/",
    membership = list_rbind(fetched)
  )
  # Write a complete snapshot before exposing the final cache filename.
  temporary <- tempfile(tmpdir = cache_dir, fileext = ".rds")
  saveRDS(frozen, temporary, compress = "xz")
  checksum <- unname(tools::md5sum(temporary))
  if (!file.rename(temporary, cache_file)) stop("Could not save MSigDB cache.")
  writeLines(checksum, checksum_file)
}
if (!file.exists(checksum_file) || !identical(
  unname(tools::md5sum(cache_file)), readLines(checksum_file, warn = FALSE)
)) {
  stop("MSigDB cache checksum is missing or mismatched. Restore the RDS ",
    "and its matching .md5; do not reuse an unverified snapshot.")
}
frozen <- readRDS(cache_file)
stopifnot(identical(frozen$request, cache_request))
membership <- frozen$membership
stopifnot(
  identical(unique(membership$db_version), params$msigdb_version),
  setequal(unique(membership$database), collections),
  !anyNA(membership$gene), all(nzchar(membership$gene))
)
tibble(
  database_release = frozen$request$db_version,
  msigdbr_at_freeze = frozen$msigdbr_version,
  frozen_utc = frozen$created_utc,
  md5 = unname(tools::md5sum(cache_file))
)
```

## Protein-to-gene mapping

Every protein remains in the protein matrix and volcano data. For gene-set
testing, rows with missing or multiple gene symbols are excluded rather than
expanded into independent genes. Symbols are matched exactly; aliases are
not silently rewritten. If several protein rows share one gene, the row
with the most observed precursors per sample is the representative, with
`NPrec` then accession breaking ties. This choice is made once across all
contrasts, without using fold changes or p-values.

```{r gene-map}
protein_map <- proteins$genes |>
  rownames_to_column("protein") |>
  mutate(
    mean_obs = rowMeans(proteins$other$n.observations),
    gene = trimws(Genes),
    mapping_status = case_when(
      is.na(gene) | !nzchar(gene) ~ "missing_symbol",
      grepl("[;,|]", gene) ~ "multiple_symbols",
      TRUE ~ "candidate"
    ),
    gene = if_else(mapping_status == "candidate", gene, NA_character_)
  )
representatives <- protein_map |>
  filter(mapping_status == "candidate") |>
  arrange(gene, desc(mean_obs), desc(NPrec), protein) |>
  distinct(gene, .keep_all = TRUE) |>
  pull(protein)
protein_map <- protein_map |>
  mutate(
    selected = protein %in% representatives,
    mapping_status = case_when(
      selected ~ "representative",
      mapping_status == "candidate" ~ "duplicate_gene",
      TRUE ~ mapping_status
    ),
    annotated = !is.na(gene) & gene %in% membership$gene,
    # Accessions distinguish duplicate symbols in protein-level plot labels.
    label = if_else(
      is.na(gene), protein,
      if_else(duplicated(gene) | duplicated(gene, fromLast = TRUE),
        paste0(gene, " (", protein, ")"), gene)
    )
  )
gene_map <- filter(protein_map, selected)
gene_universe <- gene_map$gene
stopifnot(!anyDuplicated(gene_universe), !anyDuplicated(protein_map$label))
gene_matrix <- proteins$E[gene_map$protein, , drop = FALSE]
rownames(gene_matrix) <- gene_universe
count(protein_map, mapping_status, annotated, name = "proteins")
```

An unannotated representative stays in the measured gene universe and the
gene matrix. Collection-specific annotated universes are also exported, so
a later enrichment test can state its background explicitly.

## Size and overlap filters

The source-size bounds are `r params$min_set_size`–`r params$max_set_size`
unique genes, with at least `r params$min_measured` measured representatives.
The second filter uses the current matrix, not a genome-wide database size.
Both the eligible collection and a reduced collection are exported.

Within each database, the reduced collection greedily removes sets whose
Jaccard overlap with an already retained set is at least
`r params$jaccard_cutoff`. Overlap uses measured members. Larger measured
sets take priority, then larger source sets, then the set ID. Every removed
set records its retained representative and overlap. Databases are reduced
separately so a GO term cannot erase a Hallmark set. Reduction limits
redundancy; it does not make the surviving sets independent.

```{r filter-sets}
set_members <- membership |> distinct(set_id, gene)
sets_full <- split(set_members$gene, set_members$set_id) |> map(sort)
sets_measured <- map(sets_full, intersect, y = gene_universe)
set_catalog <- membership |>
  distinct(set_id, database, pathway, msigdb_id, source_id, description, url) |>
  mutate(
    source_size = lengths(sets_full)[set_id],
    measured_size = lengths(sets_measured)[set_id],
    measured_fraction = measured_size / source_size,
    status = case_when(
      source_size < params$min_set_size ~ "source_too_small",
      source_size > params$max_set_size ~ "source_too_large",
      measured_size < params$min_measured ~ "too_few_measured",
      TRUE ~ "eligible"
    ),
    representative = NA_character_, jaccard = NA_real_
  )
stopifnot(!anyDuplicated(set_catalog$set_id))
eligible_ids <- set_catalog$set_id[set_catalog$status == "eligible"]
if (!length(eligible_ids)) stop("No gene sets passed the size/coverage filters.")

reduce_sets <- function(catalog) {
  catalog <- arrange(catalog, desc(measured_size), desc(source_size), set_id)
  sets <- sets_measured[catalog$set_id]
  incidence <- Matrix::sparseMatrix(
    i = match(unlist(sets, use.names = FALSE), gene_universe),
    j = rep(seq_along(sets), lengths(sets)), x = 1,
    dims = c(length(gene_universe), length(sets))
  )
  intersections <- Matrix::crossprod(incidence)
  kept <- integer()
  for (i in seq_along(sets)) {
    similarity <- numeric()
    if (length(kept)) {
      shared <- as.numeric(intersections[i, kept])
      similarity <- shared / (lengths(sets)[i] + lengths(sets)[kept] - shared)
    }
    if (length(similarity) && max(similarity) >= params$jaccard_cutoff) {
      j <- kept[which.max(similarity)]
      catalog$status[i] <- "redundant"
      catalog$representative[i] <- catalog$set_id[j]
      catalog$jaccard[i] <- max(similarity)
    } else {
      kept <- c(kept, i)
      catalog$status[i] <- "retained"
    }
  }
  catalog
}
reduced_catalog <- set_catalog |>
  filter(status == "eligible") |>
  split(~ database) |>
  map(reduce_sets) |>
  list_rbind()
set_catalog <- bind_rows(
  filter(set_catalog, status != "eligible"), reduced_catalog
) |>
  arrange(match(database, collections), set_id)
sets_eligible <- sets_measured[eligible_ids]
sets_retained <- sets_measured[set_catalog$set_id[set_catalog$status == "retained"]]
set_indices <- map(sets_retained, function(genes) {
  match(gene_map$protein[match(genes, gene_map$gene)], protein_ids)
})
annotated_universes <- setNames(map(collections, function(database) {
  intersect(gene_universe, membership$gene[membership$database == database])
}), collections)
set_summary <- set_catalog |>
  summarise(
    downloaded = n(),
    source_size_pass = sum(source_size >= params$min_set_size &
      source_size <= params$max_set_size),
    measured_size_pass = sum(status %in% c("retained", "redundant")),
    retained = sum(status == "retained"), .by = database
  )
set_summary
```

## Protein statistics and scores

`topTable()` reconstructs all five contrasts from the saved fit, preserving
the original BH family of all tested proteins within each contrast. FDR is
**not recalculated after gene mapping**. The inverted pi-score is
`P.Value^abs(logFC)` (smaller ranks higher); the equivalent published pi-value
is `abs(logFC) * -log10(P.Value)` (larger ranks higher). Neither is an
adjusted p-value. We export both, plus signed pi-values and moderated t
statistics, with their definitions in the handoff.

```{r protein-statistics}
protein_results <- map(contrast_names, function(contrast) {
  limma::topTable(fit, coef = contrast, number = Inf,
    adjust.method = "BH", sort.by = "none") |>
    rownames_to_column("protein") |>
    select(protein, logFC, AveExpr, t, P.Value, adj.P.Val, B) |>
    mutate(contrast = contrast, .before = 1)
}) |>
  list_rbind() |>
  left_join(protein_map, by = "protein", relationship = "many-to-one") |>
  mutate(
    padj = adj.P.Val,
    # Keep exact p-values in the export; bound only the logarithm for plotting.
    plot_p = pmax(P.Value, .Machine$double.xmin),
    pi_score = P.Value^abs(logFC),
    pi_value = if_else(logFC == 0, 0, abs(logFC) * -log10(plot_p)),
    signed_pi = sign(logFC) * pi_value,
    fdr_significant = padj < params$protein_fdr,
    direction = case_when(
      fdr_significant & logFC > 0 ~ "up",
      fdr_significant & logFC < 0 ~ "down",
      TRUE ~ "not_significant"
    )
  )
stopifnot(
  nrow(protein_results) == length(protein_ids) * length(contrast_names),
  !anyDuplicated(protein_results[c("contrast", "protein")]),
  all(is.finite(protein_results$logFC)),
  all(is.finite(protein_results$t)),
  all(is.finite(protein_results$P.Value)),
  all(protein_results$P.Value >= 0 & protein_results$P.Value <= 1),
  all(is.finite(protein_results$padj)),
  all(protein_results$padj >= 0 & protein_results$padj <= 1)
)
for (contrast in contrast_names) {
  result <- filter(protein_results, .data$contrast == .env$contrast)
  stopifnot(isTRUE(all.equal(result$padj, p.adjust(result$P.Value, "BH"))))
}
gene_results <- filter(protein_results, selected)
ranked_t <- setNames(map(contrast_names, function(contrast) {
  result <- filter(gene_results, .data$contrast == .env$contrast)
  sort(setNames(result$t, result$gene), decreasing = TRUE)
}), contrast_names)
ranked_pi <- setNames(map(training_contrasts, function(contrast) {
  result <- filter(gene_results, .data$contrast == .env$contrast)
  sort(setNames(result$signed_pi, result$gene), decreasing = TRUE)
}), training_contrasts)
contrast_summary <- protein_results |>
  summarise(
    tested = n(), fdr_up = sum(direction == "up"),
    fdr_down = sum(direction == "down"), min_fdr = min(padj),
    zero_p_values = sum(P.Value == 0), .by = contrast
  )
contrast_summary |> mutate(min_fdr = signif(min_fdr, 3))
```

The negative control, `r d$negative_control`, has
**`r with(filter(contrast_summary, contrast == d$negative_control), fdr_up + fdr_down)`**
proteins below BH FDR `r params$protein_fdr`. This is the reference for later
set tests, not a reason to relax thresholds when a contrast is empty.

## Protein volcanoes: BH FDR

[enrichVolcano](https://github.com/Dustyn-T-Lewis/enrichVolcano) draws these
protein volcanoes using an empty enrichment table. The x-axis is log2 fold
change and the y-axis is −log10(raw p). Colours and count badges use
**BH FDR < `r params$protein_fdr`**, with no added fold-change threshold.
Labels identify the lowest-FDR proteins among those passing that rule.
The package rescales each contrast to its own plotting area; cloud heights
and widths must not be compared across panels. Exact values are in the
workbook and CSV.

```{r plot-functions}
# Visual specification: all proteins; raw-p height; BH colours; no pathway
# claims. One 7 x 6.5 inch panel per contrast, PNG and vector PDF exports.
empty_enrichment <- tibble(
  pathway = character(), NES = double(), padj = double(), size = integer(),
  leading_edge = character()
)
# 0.3.0.9000 rejects zero enrichment rows at validation, even though the
# renderer handles them. Override that one check in a private function
# environment; leave the installed namespace and nonempty validation intact.
draw_volcano <- local({
  draw <- enrichVolcano::volcano_ring
  package_env <- environment(draw)
  validate <- get("validate_enrich_df", envir = package_env)
  adapter_env <- new.env(parent = package_env)
  adapter_env$validate_enrich_df <- function(enrich_df, cols, ...) {
    if (is.data.frame(enrich_df) && nrow(enrich_df) == 0L) {
      return(invisible(enrich_df))
    }
    validate(enrich_df, cols, ...)
  }
  environment(draw) <- adapter_env
  draw
})
plot_theme <- enrichVolcano::volcano_ring_theme(
  base_size = 12, base_family = "sans", palette = "okabe",
  ns = "#777777"
)
contrast_titles <- c(
  BFR_post_vs_pre = "BFR: post vs pre",
  HLRT_post_vs_pre = "High load: post vs pre",
  BFR_vs_HLRT_at_T1 = "BFR vs high load: pre-training control",
  BFR_vs_HLRT_at_T2 = "BFR vs high load: post-training",
  interaction = "Interaction: difference in training responses"
)
make_volcano <- function(contrast, rank_by = "fdr") {
  df <- filter(protein_results, .data$contrast == .env$contrast)
  if (rank_by == "pi") {
    labels <- df |>
      arrange(desc(pi_value), protein) |>
      slice_head(n = params$label_n) |>
      pull(label)
    subtitle <- "Labels ranked by pi-value; colours and counts use BH FDR"
  } else {
    labels <- df |>
      filter(fdr_significant) |>
      arrange(padj, P.Value, protein) |>
      slice_head(n = params$label_n) |>
      pull(label)
    subtitle <- paste0("Protein significance: BH FDR < ", params$protein_fdr)
  }
  draw_volcano(
    volc_df = select(df, label, logFC, plot_p, padj),
    enrich_df = empty_enrichment,
    gene_col = "label", pval_col = "plot_p", padj_col = "padj",
    volc_sig_col = "padj", genes_col = "leading_edge",
    p_threshold = params$protein_fdr, logfc_threshold = 0,
    title = unname(contrast_titles[contrast]), subtitle = subtitle,
    label_mode = if (length(labels)) "by_genes" else "none",
    label_genes = labels, label_n = params$label_n,
    label_size = 3.1, axis_size = 3.1, count_size = 3.3,
    point_size = 1.2, theme = plot_theme
  ) +
    theme(
      plot.title = element_text(size = 13, face = "bold"),
      plot.subtitle = element_text(size = 10, face = "plain"),
      plot.margin = margin(8, 8, 8, 8)
    )
}
save_volcano <- function(plot, name) {
  for (extension in c("png", "pdf")) {
    ggsave(file.path(figure_dir, paste0(name, ".", extension)),
      plot = plot, width = 7, height = 6.5, units = "in", dpi = 300,
      bg = "white")
  }
}
```

::: {.panel-tabset}

```{r fdr-volcanoes, results='asis'}
# Control first, then the primary interaction and descriptive contrasts.
plot_order <- c(d$negative_control, "interaction", training_contrasts,
  "BFR_vs_HLRT_at_T2")
for (contrast in plot_order) {
  cat("\n\n### ", contrast_titles[[contrast]], "\n\n", sep = "")
  plot <- make_volcano(contrast)
  save_volcano(plot, paste0("protein_volcano_fdr_", contrast))
  print(plot)
  cat("\n\n")
}
```

:::

## Training volcanoes: pi-ranked labels

These views use the same coordinates, FDR colours, and FDR counts. Only the
labels change: the top `r params$label_n` proteins by pi-value are labelled,
whether or not they pass FDR. **A pi label is a ranking, not a discovery.**
Pi-ranked views are limited to the two training contrasts, following the
differential-stage policy. Their tables include mean observed precursors
per sample because large fold changes can be driven by sparse detection.

::: {.panel-tabset}

```{r pi-volcanoes, results='asis'}
pi_top <- protein_results |>
  filter(contrast %in% training_contrasts) |>
  arrange(contrast, desc(pi_value), protein) |>
  slice_head(n = 10, by = contrast)
for (contrast in training_contrasts) {
  cat("\n\n### ", contrast_titles[[contrast]], "\n\n", sep = "")
  plot <- make_volcano(contrast, rank_by = "pi")
  save_volcano(plot, paste0("protein_volcano_pi_rank_", contrast))
  print(plot)
  cat("\n\n")
  displayed <- pi_top |>
    filter(.data$contrast == .env$contrast) |>
    select(protein, Genes, logFC, padj, pi_score, pi_value, mean_obs) |>
    mutate(across(c(logFC, pi_value, mean_obs), ~ round(.x, 2)),
      across(c(padj, pi_score), ~ signif(.x, 3)))
  print(knitr::kable(displayed, row.names = FALSE))
  cat("\n\n")
}
```

:::

## Outputs and handoff

All derived data are rebuilt on every render from the current inputs. The
source cache remains frozen. The workbook provides the protein matrix,
sample metadata, all protein and representative-gene statistics, the
mapping decisions, set catalog, and summaries. CSV exports retain full
precision. The RDS handoff keeps the model and uncertainty-bearing objects
that flat files cannot represent.

```{r write-outputs}
input_manifest <- tibble(
  input = names(input_paths),
  path = substring(unname(input_paths), nchar(here()) + 2),
  md5 = unname(tools::md5sum(input_paths))
)
package_versions <- tibble(
  package = required,
  version = vapply(required, function(package) {
    as.character(packageVersion(package))
  }, character(1))
)
provenance <- list(
  schema = 1L, created_utc = format(Sys.time(), tz = "UTC", usetz = TRUE),
  inputs = input_manifest, parameters = params,
  script_md5 = unname(tools::md5sum(file.path(
    stage, "a_script", "01_gene_sets.qmd"
  ))),
  cache_file = file.path("cache", basename(cache_file)),
  cache_md5 = unname(tools::md5sum(cache_file)),
  cache_request = frozen$request, packages = package_versions,
  mapping = paste("Exact single symbols; highest mean observed precursors,",
    "then NPrec, then accession; independent of differential statistics."),
  fdr = "BH across all tested proteins separately within each contrast",
  pi_score = "P.Value^abs(logFC); smaller ranks higher; no error control",
  pi_value = paste("abs(logFC) * -log10(max(P.Value, double.xmin));",
    "zero logFC gives zero; larger ranks higher; plotting bound only"),
  signed_pi = "sign(logFC) * pi_value; exploratory training ranks only",
  reduction = "Within database, measured-member Jaccard; deterministic greedy",
  pathway_tests_run = FALSE,
  plotting_adapter = "Private empty-enrichment validation adapter; namespace unchanged",
  enrichvolcano_renderer_sha256 = digest::digest(
    body(enrichVolcano::volcano_ring), algo = "sha256"
  )
)
gene_sets <- list(
  full = sets_full, eligible = sets_eligible, retained = sets_retained,
  catalog = set_catalog, measured_universe = gene_universe,
  annotated_universes = annotated_universes, provenance = provenance
)
pathway_inputs <- list(
  proteins = proteins, fit = fit, design = d,
  protein_results = protein_results, gene_results = gene_results,
  protein_map = protein_map, gene_matrix = gene_matrix,
  gene_sets = sets_retained, set_indices = set_indices,
  ranked_t = ranked_t, ranked_pi = ranked_pi,
  annotated_universes = annotated_universes, provenance = provenance
)
stopifnot(
  all(map_lgl(sets_retained, ~ all(.x %in% gene_universe))),
  all(lengths(sets_retained) >= params$min_measured),
  all(map_lgl(set_indices, ~ !anyNA(.x) && !anyDuplicated(.x))),
  all(map2_lgl(sets_retained, set_indices, function(genes, indices) {
    identical(genes, protein_map$gene[indices])
  })),
  identical(colnames(gene_matrix), rownames(d$design))
)
saveRDS(gene_sets, file.path(out, "gene_sets.rds"), compress = "xz")
saveRDS(pathway_inputs, file.path(out, "pathway_inputs.rds"), compress = "xz")
saveRDS(proteins$E, file.path(out, "protein_matrix.rds"), compress = "xz")
readr::write_csv(protein_results, file.path(out, "protein_statistics.csv"))
readr::write_csv(set_catalog, file.path(out, "gene_set_catalog.csv"))
readr::write_csv(input_manifest, file.path(out, "input_manifest.csv"))
matrix_table <- as.data.frame(proteins$E) |> rownames_to_column("protein")
writexl::write_xlsx(list(
  protein_matrix = matrix_table,
  samples = proteins$targets,
  protein_statistics = protein_results,
  gene_statistics = gene_results,
  protein_gene_map = protein_map,
  gene_set_catalog = set_catalog,
  set_summary = set_summary,
  contrast_summary = contrast_summary,
  pi_top_training = pi_top,
  input_manifest = input_manifest,
  package_versions = package_versions
), file.path(out, "01_gene_sets.xlsx"))
tibble(
  output = c("01_gene_sets.xlsx", "protein_matrix.rds", "gene_sets.rds",
    "pathway_inputs.rds", "protein_statistics.csv", "gene_set_catalog.csv"),
  size_mb = round(file.info(file.path(out, output))$size / 1024^2, 2)
)
```

`pathway_inputs$set_indices` refers to rows of
`pathway_inputs$proteins$E`, **not** rows of the gene matrix. Later set tests
must use the participant design and limpa's uncertainty/weights and report
the baseline-control result before interpreting the interaction. The rank
vectors are an input convenience, not an endorsement of a test that assumes
independent proteins. No ORA hit list is selected by pi-score here.

```{r session}
sessionInfo()
```
