---
title: "01 · Gene sets and protein volcanoes"
subtitle: "Frozen human pathways and the inputs for set testing"
params:
  msigdb_version: "2026.1.Hs"
  collections: [Hallmark, Reactome, GOBP]
  min_set_size: 15
  max_set_size: 500
  min_measured: 15
  jaccard_cutoff: 0.5
  protein_fdr: 0.05
  label_n: 5
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    code-fold: true
    code-tools: true
    df-print: kable
    fig-width: 7
    fig-height: 6.5
    fig-dpi: 150
    fig-align: center
execute:
  message: false
  error: false
  cache: false
---

This stage pins down which genes belong to which pathway, matches that list against the proteins
actually measured, and packages the matrix and the fitted model for whatever tests come next. The
volcanoes it draws show the protein results from `02_Differential`, unchanged. No pathway is
tested here, so no plot carries an enrichment arc or a pathway p-value.

## Inputs

```{r setup}
required <- c(
  "here", "limma", "limpa", "dplyr", "tibble", "tidyr", "purrr", "readr",
  "writexl", "digest", "Matrix", "ggplot2", "enrichVolcano", "knitr"
)
absent <- required[!vapply(required, requireNamespace, logical(1), quietly = TRUE)]
if (length(absent)) {
  stop("Install required R packages: ", paste(absent, collapse = ", "))
}
library(here)
library(dplyr)
library(tibble)
library(tidyr)
library(purrr)
library(ggplot2)
here::i_am("03_Pathway_Enrichment/01_Gene_Sets/a_script/01_gene_sets.qmd")

stage <- here("03_Pathway_Enrichment", "01_Gene_Sets")
out <- file.path(stage, "c_data")
cache_dir <- file.path(out, "cache")
figure_dir <- file.path(stage, "b_reports", "figures")
for (path in c(out, cache_dir, figure_dir)) {
  dir.create(path, recursive = TRUE, showWarnings = FALSE)
}
input_paths <- c(
  proteins = here("01_Preprocess/02_Quantification/c_data/proteins.rds"),
  fit = here("02_Differential_Expression/02_Differential/c_data/fit.rds"),
  design = here("02_Differential_Expression/01_Design/c_data/design.rds")
)
if (!all(file.exists(input_paths))) {
  stop(
    "Run the upstream stages first. Missing: ",
    paste(input_paths[!file.exists(input_paths)], collapse = ", ")
  )
}
proteins <- readRDS(input_paths[["proteins"]])
fit <- readRDS(input_paths[["fit"]])
d <- readRDS(input_paths[["design"]])
protein_ids <- rownames(proteins$E)
contrast_names <- colnames(d$contrasts)
training_contrasts <- c("BFR_post_vs_pre", "HLRT_post_vs_pre")
stopifnot(
  !anyDuplicated(protein_ids), !anyDuplicated(colnames(proteins$E)),
  identical(rownames(proteins$genes), protein_ids),
  identical(rownames(fit$coefficients), protein_ids),
  identical(colnames(proteins$E), rownames(d$design)),
  identical(colnames(proteins$E), proteins$targets$sample_id),
  identical(colnames(fit$coefficients), contrast_names),
  identical(fit$design, d$design), identical(fit$contrasts, d$contrasts),
  identical(fit$EList$E, proteins$E),
  identical(fit$EList$other, proteins$other),
  identical(fit$EList$genes, proteins$genes),
  all(is.finite(proteins$E)),
  identical(dimnames(proteins$other$standard.error), dimnames(proteins$E)),
  identical(dimnames(proteins$other$n.observations), dimnames(proteins$E)),
  all(is.finite(proteins$other$standard.error)),
  all(proteins$other$standard.error > 0),
  d$negative_control %in% contrast_names,
  all(training_contrasts %in% contrast_names)
)
if (!"volc_sig_col" %in% names(formals(enrichVolcano::volcano_ring))) {
  stop(
    "enrichVolcano with volcano_ring(volc_sig_col = ...) is required. ",
    "Install the current Dustyn-T-Lewis/enrichVolcano package."
  )
}
stopifnot(
  is.numeric(params$min_set_size), params$min_set_size >= 1,
  params$min_set_size == as.integer(params$min_set_size),
  params$max_set_size >= params$min_set_size,
  params$min_measured >= 1,
  params$min_measured == as.integer(params$min_measured),
  params$min_measured <= params$max_set_size,
  params$jaccard_cutoff > 0, params$jaccard_cutoff <= 1,
  params$protein_fdr > 0, params$protein_fdr < 1,
  params$label_n >= 0, params$label_n == as.integer(params$label_n)
)
tibble(
  n_proteins = nrow(proteins$E), n_samples = ncol(proteins$E),
  n_participants = n_distinct(proteins$targets$participant),
  n_contrasts = length(contrast_names)
)
```

The matrix is the one `dpcDE()` fitted, normalised and unaltered. The handoff carries the whole
EList with it: standard errors, observation counts, the weighted fit and the fixed-participant
design. Nothing is refitted here. The upstream approximation travels along too, since cyclic loess
moved the abundances and left the stored standard errors on the pre-normalisation scale.

## Freezing the gene sets

Set membership comes from human MSigDB through
[msigdbr](https://igordot.github.io/msigdbr/reference/msigdbr.html). The database release is
pinned on its own, separately from the version of the R package that fetches it, because the two
move independently. The first render writes a project-local RDS and an md5 beside it; every render
after that reads the snapshot and verifies the checksum, so it needs neither the network nor
`msigdbr`. Changing a size or overlap parameter re-filters the snapshot and never downloads a
different release.

```{r frozen-cache}
collection_specs <- list(
  Hallmark = list(collection = "H", subcollection = NULL),
  Reactome = list(collection = "C2", subcollection = "CP:REACTOME"),
  GOBP = list(collection = "C5", subcollection = "GO:BP"),
  GOCC = list(collection = "C5", subcollection = "GO:CC"),
  GOMF = list(collection = "C5", subcollection = "GO:MF")
)
collections <- unlist(params$collections, use.names = FALSE)
stopifnot(
  length(collections) > 0, !anyDuplicated(collections),
  all(collections %in% names(collection_specs)),
  grepl("^[0-9]{4}\\.[0-9]+\\.Hs$", params$msigdb_version)
)
cache_request <- list(
  schema = 1L, species = "Homo sapiens", db_species = "HS",
  db_version = params$msigdb_version,
  collections = collection_specs[collections]
)
cache_key <- substr(digest::digest(cache_request, algo = "sha256"), 1, 12)
cache_file <- file.path(cache_dir, paste0(
  "msigdb_", params$msigdb_version, "_", cache_key, ".rds"
))
checksum_file <- paste0(cache_file, ".md5")

if (!file.exists(cache_file)) {
  if (!requireNamespace("msigdbr", quietly = TRUE)) {
    stop(
      "The frozen cache is absent. Install msigdbr or restore ",
      "the matching RDS and .md5 files under c_data/cache/."
    )
  }
  fetched <- imap(collection_specs[collections], function(spec, database) {
    df <- msigdbr::msigdbr(
      db_species = "HS", species = "Homo sapiens",
      collection = spec$collection, subcollection = spec$subcollection
    )
    if (!identical(unique(df$db_version), params$msigdb_version)) {
      stop(
        "Requested MSigDB ", params$msigdb_version,
        " but msigdbr returned ", paste(unique(df$db_version), collapse = ", "),
        ". Restore the requested snapshot or explicitly change the version."
      )
    }
    df |>
      transmute(
        database,
        pathway = gs_name, set_id = paste(database, gs_name, sep = "|"),
        gene = gene_symbol, msigdb_id = gs_id, source_id = gs_exact_source,
        description = gs_description, url = gs_url, db_version
      ) |>
      filter(!is.na(gene), nzchar(gene)) |>
      distinct()
  })
  frozen <- list(
    request = cache_request,
    created_utc = format(Sys.time(), tz = "UTC", usetz = TRUE),
    msigdbr_version = as.character(packageVersion("msigdbr")),
    source = "https://www.gsea-msigdb.org/gsea/msigdb/",
    membership = list_rbind(fetched)
  )
  # Write a complete snapshot before exposing the final cache filename.
  temporary <- tempfile(tmpdir = cache_dir, fileext = ".rds")
  saveRDS(frozen, temporary, compress = "xz")
  checksum <- unname(tools::md5sum(temporary))
  if (!file.rename(temporary, cache_file)) stop("Could not save MSigDB cache.")
  writeLines(checksum, checksum_file)
}
if (!file.exists(checksum_file) || !identical(
  unname(tools::md5sum(cache_file)), readLines(checksum_file, warn = FALSE)
)) {
  stop(
    "MSigDB cache checksum is missing or mismatched. Restore the RDS ",
    "and its matching .md5; do not reuse an unverified snapshot."
  )
}
frozen <- readRDS(cache_file)
stopifnot(identical(frozen$request, cache_request))
membership <- frozen$membership
stopifnot(
  identical(unique(membership$db_version), params$msigdb_version),
  setequal(unique(membership$database), collections),
  !anyNA(membership$gene), all(nzchar(membership$gene))
)
tibble(
  database_release = frozen$request$db_version,
  msigdbr_at_freeze = frozen$msigdbr_version,
  frozen_utc = frozen$created_utc,
  md5 = unname(tools::md5sum(cache_file))
)
```

## One protein per gene

Set testing needs one row per gene, and the protein matrix does not supply that cleanly: some rows
carry no gene symbol, some carry several, and some genes appear on more than one row. Every
protein stays in the matrix and in the volcano data regardless. Only the gene-level view drops
them.

A row with no symbol, or with several, is left out rather than split into separate genes, since
splitting would invent measurements that were never made. Symbols match exactly and no alias is
rewritten behind your back. Where several rows share a gene, the row with the most observed
precursors per sample represents it, with `NPrec` and then accession breaking ties. That choice is
made once for all five contrasts and reads no fold change and no p-value.

```{r gene-map}
protein_map <- proteins$genes |>
  rownames_to_column("protein") |>
  mutate(
    mean_obs = rowMeans(proteins$other$n.observations),
    gene = trimws(Genes),
    mapping_status = case_when(
      is.na(gene) | !nzchar(gene) ~ "missing_symbol",
      grepl("[;,|]", gene) ~ "multiple_symbols",
      TRUE ~ "candidate"
    ),
    gene = if_else(mapping_status == "candidate", gene, NA_character_)
  )
representatives <- protein_map |>
  filter(mapping_status == "candidate") |>
  arrange(gene, desc(mean_obs), desc(NPrec), protein) |>
  distinct(gene, .keep_all = TRUE) |>
  pull(protein)
protein_map <- protein_map |>
  mutate(
    selected = protein %in% representatives,
    mapping_status = case_when(
      selected ~ "representative",
      mapping_status == "candidate" ~ "duplicate_gene",
      TRUE ~ mapping_status
    ),
    annotated = !is.na(gene) & gene %in% membership$gene,
    # Accessions distinguish duplicate symbols in protein-level plot labels.
    label = if_else(
      is.na(gene), protein,
      if_else(duplicated(gene) | duplicated(gene, fromLast = TRUE),
        paste0(gene, " (", protein, ")"), gene
      )
    )
  )
gene_map <- filter(protein_map, selected)
gene_universe <- gene_map$gene
stopifnot(!anyDuplicated(gene_universe), !anyDuplicated(protein_map$label))
gene_matrix <- proteins$E[gene_map$protein, , drop = FALSE]
rownames(gene_matrix) <- gene_universe
count(protein_map, mapping_status, annotated, name = "proteins")
```

A representative that no database annotates still counts as measured and still sits in the gene
matrix. Alongside it the stage exports one annotated background per collection, so a later
enrichment test can name the universe it tested against instead of assuming one.

## Which sets are kept

A set qualifies on two counts. Its source membership must fall between `r params$min_set_size` and
`r params$max_set_size` unique genes, and at least `r params$min_measured` of those genes must have
a representative in this matrix. The second bar reads the proteins actually measured here, which is
the number that governs power, rather than the set's size in the database.

Many sets then say almost the same thing. Within each database the stage walks the qualifying sets
from largest measured membership down and drops any whose Jaccard overlap with a set already kept
reaches `r params$jaccard_cutoff`, measuring overlap on measured members only. Ties break on source
size, then on set ID, so the result does not depend on row order. Each dropped set records which
set displaced it and by how much. Databases reduce separately, which stops a GO term from erasing a
Hallmark set that covers the same biology. Both the qualifying collection and the reduced one are
exported. Reduction cuts redundancy; it does not make what survives independent.

```{r filter-sets}
set_members <- membership |> distinct(set_id, gene)
sets_full <- split(set_members$gene, set_members$set_id) |> map(sort)
sets_measured <- map(sets_full, intersect, y = gene_universe)
set_catalog <- membership |>
  distinct(set_id, database, pathway, msigdb_id, source_id, description, url) |>
  mutate(
    source_size = lengths(sets_full)[set_id],
    measured_size = lengths(sets_measured)[set_id],
    measured_fraction = measured_size / source_size,
    status = case_when(
      source_size < params$min_set_size ~ "source_too_small",
      source_size > params$max_set_size ~ "source_too_large",
      measured_size < params$min_measured ~ "too_few_measured",
      TRUE ~ "eligible"
    ),
    representative = NA_character_, jaccard = NA_real_
  )
stopifnot(!anyDuplicated(set_catalog$set_id))
eligible_ids <- set_catalog$set_id[set_catalog$status == "eligible"]
if (!length(eligible_ids)) stop("No gene sets passed the size/coverage filters.")

reduce_sets <- function(catalog) {
  catalog <- arrange(catalog, desc(measured_size), desc(source_size), set_id)
  sets <- sets_measured[catalog$set_id]
  incidence <- Matrix::sparseMatrix(
    i = match(unlist(sets, use.names = FALSE), gene_universe),
    j = rep(seq_along(sets), lengths(sets)), x = 1,
    dims = c(length(gene_universe), length(sets))
  )
  intersections <- Matrix::crossprod(incidence)
  kept <- integer()
  for (i in seq_along(sets)) {
    similarity <- numeric()
    if (length(kept)) {
      shared <- as.numeric(intersections[i, kept])
      similarity <- shared / (lengths(sets)[i] + lengths(sets)[kept] - shared)
    }
    if (length(similarity) && max(similarity) >= params$jaccard_cutoff) {
      j <- kept[which.max(similarity)]
      catalog$status[i] <- "redundant"
      catalog$representative[i] <- catalog$set_id[j]
      catalog$jaccard[i] <- max(similarity)
    } else {
      kept <- c(kept, i)
      catalog$status[i] <- "retained"
    }
  }
  catalog
}
reduced_catalog <- set_catalog |>
  filter(status == "eligible") |>
  split(~database) |>
  map(reduce_sets) |>
  list_rbind()
set_catalog <- bind_rows(
  filter(set_catalog, status != "eligible"), reduced_catalog
) |>
  arrange(match(database, collections), set_id)
sets_eligible <- sets_measured[eligible_ids]
sets_retained <- sets_measured[set_catalog$set_id[set_catalog$status == "retained"]]
set_indices <- map(sets_retained, function(genes) {
  match(gene_map$protein[match(genes, gene_map$gene)], protein_ids)
})
annotated_universes <- setNames(map(collections, function(database) {
  intersect(gene_universe, membership$gene[membership$database == database])
}), collections)
set_summary <- set_catalog |>
  summarise(
    downloaded = n(),
    source_size_pass = sum(
      source_size >= params$min_set_size & source_size <= params$max_set_size
    ),
    measured_size_pass = sum(status %in% c("retained", "redundant")),
    retained = sum(status == "retained"), .by = database
  )
set_summary
```

## Protein statistics

`topTable()` rebuilds all five contrasts from the saved fit. The BH family stays as it was
upstream, one adjustment across every tested protein within each contrast, and gene mapping does
not trigger a recalculation. Adjusting again on the mapped subset would quietly change what every
FDR in this stage means.

Two forms of the pi score appear because the literature uses both. `pi_score` is
`P.Value^abs(logFC)`, where smaller ranks higher; `pi_value` is `abs(logFC) * -log10(P.Value)`,
where larger ranks higher. They order proteins identically. Neither is an adjusted p-value, and
neither controls anything. Both are exported, along with signed pi and the moderated t statistics,
and the handoff carries their definitions so a later stage cannot mistake one for an FDR.

```{r protein-statistics}
protein_results <- map(contrast_names, function(contrast) {
  limma::topTable(fit,
    coef = contrast, number = Inf,
    adjust.method = "BH", sort.by = "none"
  ) |>
    rownames_to_column("protein") |>
    select(protein, logFC, AveExpr, t, P.Value, adj.P.Val, B) |>
    mutate(contrast = contrast, .before = 1)
}) |>
  list_rbind() |>
  left_join(protein_map, by = "protein", relationship = "many-to-one") |>
  mutate(
    padj = adj.P.Val,
    # Keep exact p-values in the export; bound only the logarithm for plotting.
    plot_p = pmax(P.Value, .Machine$double.xmin),
    pi_score = P.Value^abs(logFC),
    pi_value = if_else(logFC == 0, 0, abs(logFC) * -log10(plot_p)),
    signed_pi = sign(logFC) * pi_value,
    fdr_significant = padj < params$protein_fdr,
    direction = case_when(
      fdr_significant & logFC > 0 ~ "up",
      fdr_significant & logFC < 0 ~ "down",
      TRUE ~ "not_significant"
    )
  )
stopifnot(
  nrow(protein_results) == length(protein_ids) * length(contrast_names),
  !anyDuplicated(protein_results[c("contrast", "protein")]),
  all(is.finite(protein_results$logFC)),
  all(is.finite(protein_results$t)),
  all(is.finite(protein_results$P.Value)),
  all(protein_results$P.Value >= 0 & protein_results$P.Value <= 1),
  all(is.finite(protein_results$padj)),
  all(protein_results$padj >= 0 & protein_results$padj <= 1)
)
for (contrast in contrast_names) {
  result <- filter(protein_results, .data$contrast == .env$contrast)
  stopifnot(isTRUE(all.equal(result$padj, p.adjust(result$P.Value, "BH"))))
}
gene_results <- filter(protein_results, selected)
ranked_t <- setNames(map(contrast_names, function(contrast) {
  result <- filter(gene_results, .data$contrast == .env$contrast)
  sort(setNames(result$t, result$gene), decreasing = TRUE)
}), contrast_names)
ranked_pi <- setNames(map(training_contrasts, function(contrast) {
  result <- filter(gene_results, .data$contrast == .env$contrast)
  sort(setNames(result$signed_pi, result$gene), decreasing = TRUE)
}), training_contrasts)
contrast_summary <- protein_results |>
  summarise(
    tested = n(), fdr_up = sum(direction == "up"),
    fdr_down = sum(direction == "down"), min_fdr = min(padj),
    zero_p_values = sum(P.Value == 0), .by = contrast
  )
contrast_summary |> mutate(min_fdr = signif(min_fdr, 3))
```

The negative control, `r d$negative_control`, returns
`r with(filter(contrast_summary, contrast == d$negative_control), fdr_up + fdr_down)`
proteins below BH FDR `r params$protein_fdr`. That count is the reference a later set test should
read its own null against. It is not licence to lower a threshold because some other contrast came
back empty.

## Volcanoes by FDR

[enrichVolcano](https://github.com/Dustyn-T-Lewis/enrichVolcano) draws these, handed an empty
enrichment table. Fold change runs along x and −log10 of the raw p-value up y. Colours and the
count badges read BH FDR < `r params$protein_fdr`, with no fold-change floor added on top. Labels
mark the lowest-FDR proteins among those that clear it.

One thing these plots cannot do is support a comparison between panels. The package rescales each
contrast to fill its own plotting area, so one cloud looking taller or wider than its neighbour
says nothing. Read the exact values from the workbook and the CSV.

```{r plot-functions}
empty_enrichment <- tibble(
  pathway = character(), NES = double(), padj = double(), size = integer(),
  leading_edge = character()
)
# 0.3.0.9000 rejects zero enrichment rows at validation, even though the
# renderer handles them. Override that one check in a private function
# environment; leave the installed namespace and nonempty validation intact.
draw_volcano <- local({
  draw <- enrichVolcano::volcano_ring
  package_env <- environment(draw)
  validate <- get("validate_enrich_df", envir = package_env)
  adapter_env <- new.env(parent = package_env)
  adapter_env$validate_enrich_df <- function(enrich_df, cols, ...) {
    if (is.data.frame(enrich_df) && nrow(enrich_df) == 0L) {
      return(invisible(enrich_df))
    }
    validate(enrich_df, cols, ...)
  }
  environment(draw) <- adapter_env
  draw
})
plot_theme <- enrichVolcano::volcano_ring_theme(
  base_size = 12, base_family = "sans", palette = "okabe",
  ns = "#777777"
)
contrast_titles <- c(
  BFR_post_vs_pre = "BFR: post vs pre",
  HLRT_post_vs_pre = "High load: post vs pre",
  BFR_vs_HLRT_at_T1 = "BFR vs high load: pre-training control",
  BFR_vs_HLRT_at_T2 = "BFR vs high load: post-training",
  interaction = "Interaction: difference in training responses"
)
make_volcano <- function(contrast, rank_by = "fdr") {
  df <- filter(protein_results, .data$contrast == .env$contrast)
  if (rank_by == "pi") {
    labels <- df |>
      arrange(desc(pi_value), protein) |>
      slice_head(n = params$label_n) |>
      pull(label)
    subtitle <- "Labels ranked by pi-value; colours and counts use BH FDR"
  } else {
    labels <- df |>
      filter(fdr_significant) |>
      arrange(padj, P.Value, protein) |>
      slice_head(n = params$label_n) |>
      pull(label)
    subtitle <- paste0("Protein significance: BH FDR < ", params$protein_fdr)
  }
  draw_volcano(
    volc_df = select(df, label, logFC, plot_p, padj),
    enrich_df = empty_enrichment,
    gene_col = "label", pval_col = "plot_p", padj_col = "padj",
    volc_sig_col = "padj", genes_col = "leading_edge",
    p_threshold = params$protein_fdr, logfc_threshold = 0,
    title = unname(contrast_titles[contrast]), subtitle = subtitle,
    label_mode = if (length(labels)) "by_genes" else "none",
    label_genes = labels, label_n = params$label_n,
    label_size = 3.1, axis_size = 3.1, count_size = 3.3,
    point_size = 1.2, theme = plot_theme
  ) +
    theme(
      plot.title = element_text(size = 13, face = "bold"),
      plot.subtitle = element_text(size = 10, face = "plain"),
      plot.margin = margin(8, 8, 8, 8)
    )
}
save_volcano <- function(plot, name) {
  for (extension in c("png", "pdf")) {
    ggsave(file.path(figure_dir, paste0(name, ".", extension)),
      plot = plot, width = 7, height = 6.5, units = "in", dpi = 300,
      bg = "white"
    )
  }
}
```

::: {.panel-tabset}

```{r fdr-volcanoes, results='asis'}
# Control first, then the primary interaction and descriptive contrasts.
plot_order <- c(
  d$negative_control, "interaction", training_contrasts,
  "BFR_vs_HLRT_at_T2"
)
for (contrast in plot_order) {
  cat("\n\n### ", contrast_titles[[contrast]], "\n\n", sep = "")
  plot <- make_volcano(contrast)
  save_volcano(plot, paste0("protein_volcano_fdr_", contrast))
  print(plot)
  cat("\n\n")
}
```

:::

## The same plots, ranked by pi

Same coordinates, same FDR colours, same counts. Only the labels move: here the top
`r params$label_n` proteins by pi-value are named, whether or not they pass FDR. A pi label ranks a
protein and selects nothing, so nothing labelled here has been discovered.

These views cover only the two training contrasts, which follows what `02_Differential` does and
for the same reason. Their tables report mean observed precursors per sample, because a protein
detected in almost no samples can post a large fold change on very little evidence.

::: {.panel-tabset}

```{r pi-volcanoes, results='asis'}
pi_top <- protein_results |>
  filter(contrast %in% training_contrasts) |>
  arrange(contrast, desc(pi_value), protein) |>
  slice_head(n = 10, by = contrast)
for (contrast in training_contrasts) {
  cat("\n\n### ", contrast_titles[[contrast]], "\n\n", sep = "")
  plot <- make_volcano(contrast, rank_by = "pi")
  save_volcano(plot, paste0("protein_volcano_pi_rank_", contrast))
  print(plot)
  cat("\n\n")
  displayed <- pi_top |>
    filter(.data$contrast == .env$contrast) |>
    select(protein, Genes, logFC, padj, pi_score, pi_value, mean_obs) |>
    mutate(
      across(c(logFC, pi_value, mean_obs), ~ round(.x, 2)),
      across(c(padj, pi_score), ~ signif(.x, 3))
    )
  print(knitr::kable(displayed, row.names = FALSE))
  cat("\n\n")
}
```

:::

## Outputs

Every derived file is rebuilt from the current inputs on each render, so no stale overlap filter or
old score can survive a parameter change. The frozen source cache is the one exception, which is
the point of freezing it.

The workbook is the readable copy: protein matrix, sample metadata, protein and gene statistics,
the mapping decisions, the set catalog and the summary tables. The CSVs carry the same numbers at
full precision for anything that needs to load rather than read them. The two RDS files hold what a
flat file cannot, namely the fit, the standard errors and the set indices.

```{r write-outputs}
input_manifest <- tibble(
  input = names(input_paths),
  path = substring(unname(input_paths), nchar(here()) + 2),
  md5 = unname(tools::md5sum(input_paths))
)
package_versions <- tibble(
  package = required,
  version = vapply(required, function(package) {
    as.character(packageVersion(package))
  }, character(1))
)
provenance <- list(
  schema = 1L, created_utc = format(Sys.time(), tz = "UTC", usetz = TRUE),
  inputs = input_manifest, parameters = params,
  script_md5 = unname(tools::md5sum(file.path(
    stage, "a_script", "01_gene_sets.qmd"
  ))),
  cache_file = file.path("cache", basename(cache_file)),
  cache_md5 = unname(tools::md5sum(cache_file)),
  cache_request = frozen$request, packages = package_versions,
  mapping = paste(
    "Exact single symbols; highest mean observed precursors,",
    "then NPrec, then accession; independent of differential statistics."
  ),
  fdr = "BH across all tested proteins separately within each contrast",
  pi_score = "P.Value^abs(logFC); smaller ranks higher; no error control",
  pi_value = paste(
    "abs(logFC) * -log10(max(P.Value, double.xmin));",
    "zero logFC gives zero; larger ranks higher; plotting bound only"
  ),
  signed_pi = "sign(logFC) * pi_value; exploratory training ranks only",
  reduction = "Within database, measured-member Jaccard; deterministic greedy",
  pathway_tests_run = FALSE,
  plotting_adapter = "Private empty-enrichment validation adapter; namespace unchanged",
  enrichvolcano_renderer_sha256 = digest::digest(
    body(enrichVolcano::volcano_ring),
    algo = "sha256"
  )
)
gene_sets <- list(
  full = sets_full, eligible = sets_eligible, retained = sets_retained,
  catalog = set_catalog, measured_universe = gene_universe,
  annotated_universes = annotated_universes, provenance = provenance
)
pathway_inputs <- list(
  proteins = proteins, fit = fit, design = d,
  protein_results = protein_results, gene_results = gene_results,
  protein_map = protein_map, gene_matrix = gene_matrix,
  gene_sets = sets_retained, set_indices = set_indices,
  ranked_t = ranked_t, ranked_pi = ranked_pi,
  annotated_universes = annotated_universes, provenance = provenance
)
stopifnot(
  all(map_lgl(sets_retained, ~ all(.x %in% gene_universe))),
  all(lengths(sets_retained) >= params$min_measured),
  all(map_lgl(set_indices, ~ !anyNA(.x) && !anyDuplicated(.x))),
  all(map2_lgl(sets_retained, set_indices, function(genes, indices) {
    identical(genes, protein_map$gene[indices])
  })),
  identical(colnames(gene_matrix), rownames(d$design))
)
saveRDS(gene_sets, file.path(out, "gene_sets.rds"), compress = "xz")
saveRDS(pathway_inputs, file.path(out, "pathway_inputs.rds"), compress = "xz")
saveRDS(proteins$E, file.path(out, "protein_matrix.rds"), compress = "xz")
readr::write_csv(protein_results, file.path(out, "protein_statistics.csv"))
readr::write_csv(set_catalog, file.path(out, "gene_set_catalog.csv"))
readr::write_csv(input_manifest, file.path(out, "input_manifest.csv"))
matrix_table <- as.data.frame(proteins$E) |> rownames_to_column("protein")
writexl::write_xlsx(list(
  protein_matrix = matrix_table,
  samples = proteins$targets,
  protein_statistics = protein_results,
  gene_statistics = gene_results,
  protein_gene_map = protein_map,
  gene_set_catalog = set_catalog,
  set_summary = set_summary,
  contrast_summary = contrast_summary,
  pi_top_training = pi_top,
  input_manifest = input_manifest,
  package_versions = package_versions
), file.path(out, "01_gene_sets.xlsx"))
tibble(
  output = c(
    "01_gene_sets.xlsx", "protein_matrix.rds", "gene_sets.rds",
    "pathway_inputs.rds", "protein_statistics.csv", "gene_set_catalog.csv"
  ),
  size_mb = round(file.info(file.path(out, output))$size / 1024^2, 2)
)
```

One trap to note before the next stage. `pathway_inputs$set_indices` indexes rows of
`pathway_inputs$proteins$E`, never rows of the gene matrix, and the two have different row counts.

A set test reading this handoff has to keep the participant design and limpa's weights, and it
should report the baseline control before anyone reads the interaction. The ranked vectors are here
because some tools want them, not because a test assuming independent proteins is appropriate for
this design. No hit list is selected by pi score at this stage.

```{r session}
sessionInfo()
```
