---
title: "01 · Gene sets and fgsea"
subtitle: "Frozen human pathways, and a ranked enrichment test on each"
params:
  msigdb_version: "2026.1.Hs"
  collections: [Hallmark, Reactome, GOBP]
  min_set_size: 15
  max_set_size: 500
  min_measured: 15
  jaccard_cutoff: 0.5
  protein_fdr: 0.05
  set_fdr: 0.05
  exclude_disease: true
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    code-fold: true
    code-tools: true
    df-print: kable
    fig-width: 7
    fig-height: 6.5
    fig-dpi: 150
    fig-align: center
execute:
  message: false
  error: false
  cache: false
---

This stage pins down which genes belong to which pathway, matches that list against the proteins
actually measured, and packages the matrix and the fitted model for whatever tests come next. The
volcanoes it draws show the protein results from `02_Differential`, unchanged. No pathway is
tested here, so no plot carries an enrichment arc or a pathway p-value.

## Inputs

```{r setup}
library(here)
library(dplyr)
library(tibble)
library(tidyr)
library(purrr)
library(ggplot2)
here::i_am("03_Pathway_Enrichment/01_Gene_Sets/a_script/01_gene_sets.qmd")
source(here("03_Pathway_Enrichment", "00_Functions", "00_functions.R"))

stage <- here("03_Pathway_Enrichment", "01_Gene_Sets")
out <- file.path(stage, "c_data")
cache_dir <- file.path(out, "cache")
for (path in c(out, cache_dir)) {
  dir.create(path, recursive = TRUE, showWarnings = FALSE)
}
inputs <- stage_inputs(
  proteins = "01_Preprocess/02_Quantification/c_data/proteins.rds",
  fit = "02_Differential_Expression/02_Differential/c_data/fit.rds",
  design = "02_Differential_Expression/01_Design/c_data/design.rds"
)
proteins <- inputs$proteins
fit <- inputs$fit
d <- inputs$design
protein_ids <- rownames(proteins$E)
contrast_names <- colnames(d$contrasts)
training_contrasts <- c("BFR_post_vs_pre", "HLRT_post_vs_pre")
# Three files written by three stages, any of which can be older than the others. A stale
# fit.rds still lines up by count, so every check below compares names rather than lengths.
stopifnot(
  identical(rownames(fit$coefficients), protein_ids),
  identical(colnames(fit$coefficients), contrast_names),
  identical(fit$design, d$design),
  identical(colnames(proteins$E), rownames(d$design)),
  d$negative_control %in% contrast_names,
  all(training_contrasts %in% contrast_names)
)
tibble(
  n_proteins = nrow(proteins$E), n_samples = ncol(proteins$E),
  n_participants = n_distinct(proteins$targets$participant),
  n_contrasts = length(contrast_names)
)
```

The matrix is the one `dpcDE()` fitted, normalised and unaltered, and nothing is refitted here.
The standard errors, the weighted fit and the fixed-participant design stay where the upstream
stages wrote them; this stage reads them by path and writes only what it derives. The upstream
approximation travels along, since cyclic loess moved the abundances and left the stored standard
errors on the pre-normalisation scale.

## Freezing the gene sets

Set membership comes from human MSigDB through
[msigdbr](https://igordot.github.io/msigdbr/reference/msigdbr.html). The database release is
pinned on its own, separately from the version of the R package that fetches it, because the two
move independently. The first render writes a project-local RDS and an md5 beside it; every render
after that reads the snapshot and verifies the checksum, so it needs neither the network nor
`msigdbr`. Changing a size or overlap parameter re-filters the snapshot and never downloads a
different release.

```{r frozen-cache}
collection_specs <- list(
  Hallmark = list(collection = "H", subcollection = NULL),
  Reactome = list(collection = "C2", subcollection = "CP:REACTOME"),
  GOBP = list(collection = "C5", subcollection = "GO:BP"),
  GOCC = list(collection = "C5", subcollection = "GO:CC"),
  GOMF = list(collection = "C5", subcollection = "GO:MF")
)
collections <- unlist(params$collections, use.names = FALSE)
stopifnot(all(collections %in% names(collection_specs)))
cache_request <- list(
  schema = 1L, species = "Homo sapiens", db_species = "HS",
  db_version = params$msigdb_version,
  collections = collection_specs[collections]
)
# The release and the collections are what make one snapshot different from another, so
# the filename says both. Ask for a combination you have not frozen and the name misses.
cache_file <- file.path(cache_dir, paste0(
  "msigdb_", params$msigdb_version, "_", paste(collections, collapse = "-"), ".rds"
))
checksum_file <- paste0(cache_file, ".md5")

if (!file.exists(cache_file)) {
  if (!requireNamespace("msigdbr", quietly = TRUE)) {
    stop(
      "The frozen cache is absent. Install msigdbr or restore ",
      "the matching RDS and .md5 files under c_data/cache/."
    )
  }
  fetched <- imap(collection_specs[collections], function(spec, database) {
    df <- msigdbr::msigdbr(
      db_species = "HS", species = "Homo sapiens",
      collection = spec$collection, subcollection = spec$subcollection
    )
    if (!identical(unique(df$db_version), params$msigdb_version)) {
      stop(
        "Requested MSigDB ", params$msigdb_version,
        " but msigdbr returned ", paste(unique(df$db_version), collapse = ", "),
        ". Restore the requested snapshot or explicitly change the version."
      )
    }
    df |>
      transmute(
        database,
        pathway = gs_name, set_id = paste(database, gs_name, sep = "|"),
        gene = gene_symbol, msigdb_id = gs_id, source_id = gs_exact_source,
        description = gs_description, url = gs_url, db_version
      ) |>
      filter(!is.na(gene), nzchar(gene)) |>
      distinct()
  })
  frozen <- list(
    request = cache_request,
    created_utc = format(Sys.time(), tz = "UTC", usetz = TRUE),
    msigdbr_version = as.character(packageVersion("msigdbr")),
    source = "https://www.gsea-msigdb.org/gsea/msigdb/",
    membership = list_rbind(fetched)
  )
  # Write a complete snapshot before exposing the final cache filename.
  temporary <- tempfile(tmpdir = cache_dir, fileext = ".rds")
  saveRDS(frozen, temporary, compress = "xz")
  checksum <- unname(tools::md5sum(temporary))
  if (!file.rename(temporary, cache_file)) stop("Could not save MSigDB cache.")
  writeLines(checksum, checksum_file)
}
if (!file.exists(checksum_file) || !identical(
  unname(tools::md5sum(cache_file)), readLines(checksum_file, warn = FALSE)
)) {
  stop(
    "MSigDB cache checksum is missing or mismatched. Restore the RDS ",
    "and its matching .md5; do not reuse an unverified snapshot."
  )
}
frozen <- readRDS(cache_file)
membership <- frozen$membership
tibble(
  database_release = frozen$request$db_version,
  msigdbr_at_freeze = frozen$msigdbr_version,
  frozen_utc = frozen$created_utc,
  md5 = unname(tools::md5sum(cache_file))
)
```

## One protein per gene

Set testing needs one row per gene, and the protein matrix does not supply that cleanly: some rows
carry no gene symbol, some carry several, and some genes appear on more than one row. Every
protein stays in the matrix and in the volcano data regardless. Only the gene-level view drops
them.

A row with no symbol, or with several, is left out rather than split into separate genes, since
splitting would invent measurements that were never made. Symbols match exactly and no alias is
rewritten behind your back. Where several rows share a gene, the row with the most observed
precursors per sample represents it, with `NPrec` and then accession breaking ties. That choice is
made once for all five contrasts and reads no fold change and no p-value.

```{r gene-map}
protein_map <- proteins$genes |>
  rownames_to_column("protein") |>
  mutate(
    mean_obs = rowMeans(proteins$other$n.observations),
    gene = trimws(Genes),
    mapping_status = case_when(
      is.na(gene) | !nzchar(gene) ~ "missing_symbol",
      grepl("[;,|]", gene) ~ "multiple_symbols",
      TRUE ~ "candidate"
    ),
    gene = if_else(mapping_status == "candidate", gene, NA_character_)
  )
representatives <- protein_map |>
  filter(mapping_status == "candidate") |>
  arrange(gene, desc(mean_obs), desc(NPrec), protein) |>
  distinct(gene, .keep_all = TRUE) |>
  pull(protein)
protein_map <- protein_map |>
  mutate(
    selected = protein %in% representatives,
    mapping_status = case_when(
      selected ~ "representative",
      mapping_status == "candidate" ~ "duplicate_gene",
      TRUE ~ mapping_status
    ),
    annotated = !is.na(gene) & gene %in% membership$gene,
    # Accessions distinguish duplicate symbols in protein-level plot labels.
    label = if_else(
      is.na(gene), protein,
      if_else(duplicated(gene) | duplicated(gene, fromLast = TRUE),
        paste0(gene, " (", protein, ")"), gene
      )
    )
  )
gene_map <- filter(protein_map, selected)
gene_universe <- gene_map$gene
# Labels are what the volcanoes print, and a collision would put two proteins on one point.
stopifnot(!anyDuplicated(protein_map$label))
count(protein_map, mapping_status, annotated, name = "proteins")
```

A representative that no database annotates still counts as measured and still sits in the gene
matrix. Alongside it the stage exports one annotated background per collection, so a later
enrichment test can name the universe it tested against instead of assuming one.

## Which sets are kept

A set qualifies on two counts. Its source membership must fall between `r params$min_set_size` and
`r params$max_set_size` unique genes, and at least `r params$min_measured` of those genes must have
a representative in this matrix. The second bar reads the proteins actually measured here, which is
the number that governs power, rather than the set's size in the database.

Many sets then say almost the same thing. Within each database the stage walks the qualifying sets
from largest measured membership down and drops any whose Jaccard overlap with a set already kept
reaches `r params$jaccard_cutoff`, measuring overlap on measured members only. Ties go to the larger
source set, then to the lower set ID, so the result does not depend on row order. Each dropped set records which
set displaced it and by how much. Databases reduce separately, which stops a GO term from erasing a
Hallmark set that covers the same biology. Both the qualifying collection and the reduced one are
exported. Reduction cuts redundancy; it does not make what survives independent.

```{r filter-sets}
stopifnot(params$max_set_size >= params$min_set_size)
set_members <- membership |> distinct(set_id, gene)
sets_full <- split(set_members$gene, set_members$set_id) |> map(sort)
sets_measured <- map(sets_full, intersect, y = gene_universe)
set_catalog <- membership |>
  distinct(set_id, database, pathway, msigdb_id, source_id, description, url) |>
  mutate(
    source_size = lengths(sets_full)[set_id],
    measured_size = lengths(sets_measured)[set_id],
    measured_fraction = measured_size / source_size,
    status = case_when(
      source_size < params$min_set_size ~ "source_too_small",
      source_size > params$max_set_size ~ "source_too_large",
      measured_size < params$min_measured ~ "too_few_measured",
      TRUE ~ "eligible"
    ),
    representative = NA_character_, jaccard = NA_real_
  )
stopifnot(!anyDuplicated(set_catalog$set_id))
eligible_ids <- set_catalog$set_id[set_catalog$status == "eligible"]
if (!length(eligible_ids)) stop("No gene sets passed the size/coverage filters.")

reduced_catalog <- set_catalog |>
  filter(status == "eligible") |>
  split(~database) |>
  map(\(part) reduce_sets(part, sets_measured, gene_universe, params$jaccard_cutoff)) |>
  list_rbind()
set_catalog <- bind_rows(
  filter(set_catalog, status != "eligible"), reduced_catalog
) |>
  arrange(match(database, collections), set_id)
sets_eligible <- sets_measured[eligible_ids]
sets_retained <- sets_measured[set_catalog$set_id[set_catalog$status == "retained"]]
set_indices <- map(sets_retained, function(genes) {
  match(gene_map$protein[match(genes, gene_map$gene)], protein_ids)
})
annotated_universes <- map(set_names(collections), function(database) {
  intersect(gene_universe, membership$gene[membership$database == database])
})
set_summary <- set_catalog |>
  summarise(
    downloaded = n(),
    source_size_pass = sum(
      source_size >= params$min_set_size & source_size <= params$max_set_size
    ),
    measured_size_pass = sum(status %in% c("retained", "redundant")),
    retained = sum(status == "retained"), .by = database
  )
set_summary
```

## Protein statistics

`topTable()` rebuilds all five contrasts from the saved fit. The BH family stays as it was
upstream, one adjustment across every tested protein within each contrast, and gene mapping does
not trigger a recalculation. Adjusting again on the mapped subset would quietly change what every
FDR in this stage means.

Two forms of the pi score appear because the literature uses both. `pi_score` is
`P.Value^abs(logFC)`, where smaller ranks higher; `pi_value` is `abs(logFC) * -log10(P.Value)`,
where larger ranks higher. They order proteins identically. Neither is an adjusted p-value, and
neither controls anything. Both are exported, along with signed pi and the moderated t statistics,
and the handoff carries their definitions so a later stage cannot mistake one for an FDR.

```{r protein-statistics}
protein_results <- map(contrast_names, function(contrast) {
  limma::topTable(fit,
    coef = contrast, number = Inf,
    adjust.method = "BH", sort.by = "none"
  ) |>
    rownames_to_column("protein") |>
    select(protein, logFC, AveExpr, t, P.Value, adj.P.Val, B) |>
    mutate(contrast = contrast, .before = 1)
}) |>
  list_rbind() |>
  left_join(protein_map, by = "protein", relationship = "many-to-one") |>
  mutate(
    padj = adj.P.Val,
    # Keep exact p-values in the export; bound only the logarithm for plotting.
    plot_p = pmax(P.Value, .Machine$double.xmin),
    pi_score = P.Value^abs(logFC),
    pi_value = abs(logFC) * -log10(plot_p),
    signed_pi = sign(logFC) * pi_value,
    fdr_significant = padj < params$protein_fdr,
    direction = case_when(
      fdr_significant & logFC > 0 ~ "up",
      fdr_significant & logFC < 0 ~ "down",
      TRUE ~ "not_significant"
    )
  )
# The join above is many-to-one, so a protein absent from the map would come back as NA
# rows rather than as an error.
stopifnot(nrow(protein_results) == length(protein_ids) * length(contrast_names))
gene_results <- filter(protein_results, selected)
contrast_summary <- protein_results |>
  summarise(
    tested = n(), fdr_up = sum(direction == "up"),
    fdr_down = sum(direction == "down"), min_fdr = min(padj),
    zero_p_values = sum(P.Value == 0), .by = contrast
  )
contrast_summary |> mutate(min_fdr = signif(min_fdr, 3))
```

The negative control, `r d$negative_control`, returns
`r with(filter(contrast_summary, contrast == d$negative_control), fdr_up + fdr_down)`
proteins below BH FDR `r params$protein_fdr`. That count is the reference a later set test should
read its own null against. It is not licence to lower a threshold because some other contrast came
back empty.

## Ranked enrichment

`fgsea()` walks the proteins ordered by moderated t and asks whether a set's members cluster
toward one end. It is fast, it needs no per-sample data, and it is what `enrichVolcano` consumes,
which is why it runs here. It also assumes the ranked proteins are exchangeable, and they are
not: mitochondrial and ribosomal sets move together by construction, and every participant
contributed four samples.

That assumption has a measurable cost on this data, so the negative control is reported before
any other result. `02_Set_Tests` runs `fry()`, which takes the design and the participant
structure, and that is the test to quote.

```{r ranks}
# One vector per contrast, representative proteins only, keyed by gene symbol so fgsea's
# leadingEdge comes back in the same namespace the volcanoes label with.
ranked <- protein_results |>
  filter(selected, !is.na(gene)) |>
  split(~contrast) |>
  map(\(result) set_names(result$t, result$gene))
ranked <- ranked[contrast_names]

gene_sets <- list(
  full = sets_full, eligible = sets_eligible, retained = sets_retained,
  catalog = set_catalog, measured_universe = gene_universe,
  annotated_universes = annotated_universes
)
# Per database, reduced within it, plus one pooled run where a Hallmark set and its GO twin
# compete. Pooling is what you want when several databases feed one figure.
per_database <- map(set_names(collections), \(db) {
  load_sets(gene_sets, db, scope = "within", cutoff = params$jaccard_cutoff)$sets
})
pooled <- load_sets(gene_sets, collections, scope = "across", cutoff = params$jaccard_cutoff)
set_lists <- c(per_database, list(combined = pooled$sets))
map_int(set_lists, length)
```

```{r fgsea}
fgsea_raw <- run_fgsea_all(
  ranked, set_lists,
  min_size = params$min_measured, max_size = params$max_set_size
)
filtered <- drop_disease_terms(
  fgsea_raw,
  patterns = if (isTRUE(params$exclude_disease)) DISEASE_PATTERNS else NULL
)
fgsea_results <- filtered$kept
excluded_terms <- filtered$excluded
c(tested = nrow(fgsea_raw), kept = nrow(fgsea_results), excluded = nrow(excluded_terms))
```

## The negative control, first

`r d$negative_control` compares two legs of the same person before either was trained, so it
cannot contain signal. Whatever fgsea calls significant here is what its independence assumption
costs on this data. Read every other count in this notebook against this number, not against zero.

```{r fgsea-control}
fgsea_results |>
  filter(contrast == d$negative_control) |>
  summarise(
    sets_tested = n(),
    fdr_significant = sum(padj < params$set_fdr),
    min_padj = signif(min(padj), 3), .by = collection
  )
```

## Counts

```{r fgsea-counts}
fgsea_results |>
  summarise(
    tested = n(),
    up = sum(padj < params$set_fdr & NES > 0),
    down = sum(padj < params$set_fdr & NES < 0),
    .by = c(collection, contrast)
  ) |>
  tidyr::pivot_wider(
    names_from = collection, values_from = c(up, down), values_fill = 0
  )
```

## What the name filter removed

Excluding a set because its name mentions a disease is a judgement, not a statistic. A set called
`REACTOME_INFLUENZA_INFECTION` is largely ribosome and translation machinery under a misleading
label, so removing it takes real biology with it. Every removal is listed here with the pattern
that caught it, the unfiltered table is exported beside the filtered one, and setting
`exclude_disease: false` reproduces the unfiltered result exactly.

```{r disease-excluded}
if (nrow(excluded_terms)) {
  excluded_terms |>
    distinct(collection, pathway, matched_pattern) |>
    count(collection, matched_pattern, name = "sets") |>
    arrange(desc(sets))
} else {
  tibble(note = "No name filter applied")
}
```

## Outputs

Everything derived is rebuilt from the current inputs on each render, so no stale overlap filter
or old score survives a parameter change. The frozen source cache is the one exception, which is
the point of freezing it.

The workbook carries the fgsea results three ways: unfiltered, name-filtered, and pooled across
databases. The protein-level tables stay, since `03_Volcanoes` draws its points and its labels
from them.

```{r write-outputs}
# leadingEdge is a list column, which no flat format holds. Collapse for the readable copies
# and keep the list form in the RDS.
flatten_edges <- function(df) {
  mutate(df, leadingEdge = map_chr(leadingEdge, paste, collapse = ";"))
}
gene_sets$provenance <- NULL
write_stage_outputs(
  out, "01_gene_sets",
  sheets = list(
    fgsea_unfiltered = flatten_edges(fgsea_raw),
    fgsea_filtered = flatten_edges(fgsea_results),
    fgsea_combined = flatten_edges(filter(fgsea_results, collection == "combined")),
    excluded_terms = flatten_edges(excluded_terms),
    samples = proteins$targets,
    protein_statistics = protein_results,
    gene_statistics = gene_results,
    protein_gene_map = protein_map,
    gene_set_catalog = set_catalog,
    set_summary = set_summary,
    contrast_summary = contrast_summary
  ),
  # What this stage derives, and nothing it merely read. proteins.rds and fit.rds are tracked
  # upstream, so copying them here would only let the copy go stale.
  rds = list(
    gene_sets = gene_sets,
    pathway_inputs = list(
      protein_results = protein_results, protein_map = protein_map,
      gene_sets = sets_retained, set_indices = set_indices,
      set_lists = set_lists, fgsea_results = fgsea_results
    )
  ),
  csv = list(
    fgsea_results = flatten_edges(fgsea_results),
    protein_statistics = protein_results,
    gene_set_catalog = set_catalog
  ),
  manifest = inputs$manifest, parameters = params,
  packages = c(
    "here", "limma", "limpa", "dplyr", "tibble", "tidyr", "purrr", "readr",
    "writexl", "fgsea", "knitr"
  ),
  notes = list(
    cache_file = file.path("cache", basename(cache_file)),
    cache_md5 = unname(tools::md5sum(cache_file)),
    cache_request = frozen$request,
    mapping = paste(
      "Exact single symbols; highest mean observed precursors,",
      "then NPrec, then accession; independent of differential statistics."
    ),
    fdr = "BH across all tested proteins separately within each contrast",
    pi_score = "P.Value^abs(logFC); smaller ranks higher; no error control",
    reduction = paste(
      "Measured-member Jaccard, deterministic greedy; within database per",
      "collection, across databases for `combined`"
    ),
    set_test = paste(
      "fgsea on moderated t, seed 1; assumes exchangeable proteins,",
      "so read its negative control"
    ),
    disease_filter = if (isTRUE(params$exclude_disease)) {
      paste("Sets excluded by name on:", paste(DISEASE_PATTERNS, collapse = ", "))
    } else {
      "None"
    },
    pathway_tests_run = TRUE
  )
)
```

Two traps before the next stage. `pathway_inputs$set_indices` indexes rows of `proteins$E` as the
upstream stage wrote it, so `02_Set_Tests` opens `proteins.rds` itself rather than indexing into
anything built here. And `fgsea_results` is keyed on gene symbols, which is what lets
`03_Volcanoes` match `leadingEdge` against its point labels; rank on anything else and the tick
lines vanish without a warning.

```{r session}
sessionInfo()
```
