---
title: "01 · Design"
subtitle: "Design matrix and five contrasts"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    df-print: kable
    fig-width: 7
    fig-height: 5
    fig-dpi: 150
    fig-align: center
execute:
  message: false
---

limpa works with any design limma accepts. This notebook builds one and checks it, apart from
the fit, so a design problem surfaces in seconds.

```{r setup}
pacman::p_load(here, limma, purrr, tibble, writexl)

proteins <- readRDS(here("01_Preprocess", "02_Quantification", "c_data", "proteins.rds"))
targets <- proteins$targets

with(targets, table(treatment, timepoint))
```

## Participant is a fixed block

Four cells, one per treatment-and-timepoint combination, plus participant as a fixed block: 131
rows by 36 columns, four group means and 32 participant dummies, 95 residual degrees of freedom.
Fixed rather than random because every contrast the study asks for sits within one person
(limma User's Guide §9.4.2), which is what makes pre-to-post paired. `sex` is constant within
participant, so this design cannot test it. One participant contributes three cells rather than
four, so the group columns are participant-adjusted cell means.

```{r design}
design <- model.matrix(~ 0 + group + participant, data = targets)
# makeContrasts() parses its arguments as R code, so a column left named "group2E-T1" would
# read as subtraction. The sub() strips the prefix and make.names() below checks the result.
colnames(design) <- sub("^group", "", colnames(design))
rownames(design) <- targets$sample_id

# Rank deficiency would make the contrasts inestimable and surface as silent NA coefficients.
stopifnot(
  is.fullrank(design),
  identical(make.names(colnames(design)), colnames(design))
)

c(
  samples = nrow(design), coefficients = ncol(design),
  residual_df = nrow(design) - ncol(design)
)
```

## The within-leg correlation is checked, not modelled

Samples share a participant, and within that they share a leg. The participant term removes the
first exactly; the second stays in the residual. If the leg-level correlation were large, the
within-leg contrasts would be tested too conservatively and the between-leg ones too liberally.

```{r correlation-strata}
leg_id <- paste(targets$participant, targets$leg, sep = "_")

strata <- tibble(
  level = "within leg, participant fixed (unweighted matrix correlation)",
  correlation = duplicateCorrelation(proteins$E, design, block = leg_id)$consensus.correlation
)
round(strata$correlation, 3)
```

The estimate runs on the bare matrix, without the precision and sample weights the fit uses, so it
describes the matrix rather than the model. That is enough for a near-zero check. When the number
sits near zero, as it has on this data, `02_Differential` calls `dpcDE()` without `block =`.
If the correlation ever rose, the fix is `block = leg_id` in the next notebook, but `leg_id` is
not saved, and no branch and no warning will tell you. Read the row on every rerun.

## One primary contrast, one control

```{r contrasts}
contrasts <- makeContrasts(
  `BFR_Post-Pre` = BFR_T2 - BFR_T1,
  `HLRT_Post-Pre` = HLRT_T2 - HLRT_T1,
  `BFR_Pre-HLRT_Pre` = BFR_T1 - HLRT_T1,
  `BFR_Post-HLRT_Post` = BFR_T2 - HLRT_T2,
  Modality_x_Time_Interaction = (BFR_T2 - BFR_T1) - (HLRT_T2 - HLRT_T1),
  levels = design
)
contrasts[rowSums(abs(contrasts)) > 0, ]
```

| Name | Role | Compares | Reads as |
|---|---|---|---|
| `Modality_x_Time_Interaction` | primary | difference of the two time effects | the study question |
| `BFR_Pre-HLRT_Pre` | negative control | two legs before training | must find nothing |
| `BFR_Post-Pre` | descriptive | one leg over time | what restricted training did |
| `HLRT_Post-Pre` | descriptive | the other leg over time | what conventional training did |
| `BFR_Post-HLRT_Post` | descriptive | two legs after training | leg difference at the end |

`Modality_x_Time_Interaction` equals `BFR_Post-HLRT_Post` minus `BFR_Pre-HLRT_Pre` exactly, so
these are not five independent questions. Report the primary and the control as the result; the
other three describe what happened on the way.

```{r write}
out <- here("02_Differential_Expression", "01_Design", "c_data")
saveRDS(
  list(
    design = design, contrasts = contrasts,
    negative_control = "BFR_Pre-HLRT_Pre"
  ),
  file.path(out, "design.rds")
)
sheets <- list(correlation_strata = strata)
overview <- tibble(
  sheet = names(sheets),
  rows = map_int(sheets, nrow),
  columns = map_int(sheets, ncol),
  description = "Within-leg correlation, participant fixed"
)
write_xlsx(c(list(overview = overview), sheets), file.path(out, "01_design.xlsx"))
```

```{r session}
sessionInfo()
```
