---
title: "01 · Design"
subtitle: "Design matrix and the five contrasts"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    df-print: kable
    fig-width: 7
    fig-height: 5
    fig-dpi: 150
    fig-align: center
execute:
  message: false
---

limpa works with any design limma accepts. This notebook builds one and checks it, apart from
the fit.

```{r setup}
pacman::p_load(here, limma, purrr, tibble, writexl)

proteins <- readRDS(here("01_Preprocess", "02_Quantification", "c_data", "proteins.rds"))
targets <- proteins$targets

with(targets, table(treatment, timepoint))
```

## Design

Four cells, one per treatment-and-timepoint combination, plus participant as a fixed block.
Fixed rather than random because every contrast the study asks for sits within one person
(limma User's Guide §9.4.2). `sex` is constant within participant, so this design cannot test it.

```{r design}
design <- model.matrix(~ 0 + group + participant, data = targets)
# makeContrasts() parses its arguments as R code, so a column left named "group2E-T1" would
# read as subtraction. The sub() strips the prefix and make.names() below checks the result.
colnames(design) <- sub("^group", "", colnames(design))
rownames(design) <- targets$sample_id

# Rank deficiency would make the contrasts inestimable and surface as silent NA coefficients.
stopifnot(
  is.fullrank(design),
  identical(make.names(colnames(design)), colnames(design))
)

c(
  samples = nrow(design), coefficients = ncol(design),
  residual_df = nrow(design) - ncol(design)
)
```

## Is a fixed participant term enough

Samples share a participant, and within that they share a leg. The participant term removes the
first exactly; the second stays in the residual. If the leg-level correlation were large, the
within-leg contrasts would be tested too conservatively and the between-leg ones too liberally.

```{r correlation-strata}
leg_id <- paste(targets$participant, targets$leg, sep = "_")

strata <- tibble(
  level = "within leg, participant fixed (unweighted matrix correlation)",
  correlation = duplicateCorrelation(proteins$E, design, block = leg_id)$consensus.correlation
)
round(strata$correlation, 3)
```

The estimate runs on the bare matrix, without the precision and sample weights the fit uses, so it
describes the matrix rather than the model. That is enough for a near-zero check. When the number
sits near zero, as it has on this data, `02_Differential` calls `dpcDE()` without `block =`.
If the correlation ever rose, the fix is `block = leg_id` in the next notebook, but no branch
and no warning will tell you. Read the row on every rerun.

## Contrasts

```{r contrasts}
contrasts <- makeContrasts(
  `BFR_Post-Pre` = BFR_T2 - BFR_T1,
  `HLRT_Post-Pre` = HLRT_T2 - HLRT_T1,
  `BFR_Pre-HLRT_Pre` = BFR_T1 - HLRT_T1,
  `BFR_Post-HLRT_Post` = BFR_T2 - HLRT_T2,
  Modality_x_Time_Interaction = (BFR_T2 - BFR_T1) - (HLRT_T2 - HLRT_T1),
  levels = design
)
contrasts[rowSums(abs(contrasts)) > 0, ]
```

The five carry three different roles, and the manuscript should read them that way. `Modality_x_Time_Interaction`
is the **primary** comparison and the question the study was designed to answer.
`BFR_Pre-HLRT_Pre` is the **negative control**: two legs of one person before either was trained,
so it must find nothing. The remaining three are **descriptive**. `Modality_x_Time_Interaction` equals
`BFR_Post-HLRT_Post` minus `BFR_Pre-HLRT_Pre` exactly, so these are not five independent
questions.

```{r write}
out <- here("02_Differential_Expression", "01_Design", "c_data")
saveRDS(
  list(
    design = design, contrasts = contrasts,
    negative_control = "BFR_Pre-HLRT_Pre"
  ),
  file.path(out, "design.rds")
)
sheets <- list(correlation_strata = strata)
overview <- tibble(
  sheet = names(sheets),
  rows = map_int(sheets, nrow),
  columns = map_int(sheets, ncol),
  description = "Consensus within-leg correlation with participant as a fixed term"
)
write_xlsx(c(list(overview = overview), sheets), file.path(out, "01_design.xlsx"))
```

```{r session}
sessionInfo()
```
