---
title: "01 · Design"
subtitle: "Design matrix and the five contrasts"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    df-print: kable
execute:
  message: false
---

limpa works with any design limma accepts. This notebook builds one and checks it, separately
from the fit, so a design problem surfaces in seconds rather than after an hour of modelling.

```{r setup}
library(here)
library(limpa)
library(limma)
library(dplyr)
library(readr)
library(stringr)
library(tibble)
library(nanoparquet)

proteins <- readRDS(here("01_Preprocess", "02_Quantification", "c_data", "proteins.rds"))
targets <- proteins$targets

with(targets, table(treatment, timepoint))
```

## Design

Four cell means, one per treatment-and-timepoint combination, plus participant as a fixed block.
Fixed rather than random because every contrast the study asks for is within one person, which is
the paired case of the limma User's Guide §9.4.2. `sex` is constant within participant, so this
model absorbs it and cannot test it.

```{r design}
design <- model.matrix(~ 0 + group + participant, data = targets)
colnames(design) <- sub("^group", "", colnames(design))
rownames(design) <- targets$sample_id

stopifnot(identical(make.names(colnames(design)), colnames(design)))

c(samples = nrow(design), coefficients = ncol(design),
  residual_df = nrow(design) - ncol(design))
```

`makeContrasts()` parses its arguments as R code, so the `sub()` above matters: a column named
like `2E-T1` would be read as subtraction.

## Is a fixed participant term enough

Samples share a participant, and within that they share a leg. The participant term removes the
first exactly; the second stays in the residual. If it were large, the within-leg contrasts would
be tested too conservatively and the between-leg ones too liberally.

```{r correlation-strata}
leg_id <- paste(targets$participant, targets$leg, sep = "_")

strata <- tibble(
  level = c("within participant (unweighted)", "within leg / participant fixed (unweighted)"),
  correlation = c(
    duplicateCorrelation(proteins$E, model.matrix(~ 0 + group, targets),
                         block = targets$participant)$consensus.correlation,
    duplicateCorrelation(proteins$E, design, block = leg_id)$consensus.correlation
  )
)
strata
```

Both are estimated on the bare matrix, so without the precision and sample weights the fit uses.
They describe the matrix, not the model, which is enough for a near-zero check. On this data the
second row is near zero, so `02_Differential` calls `dpcDE()` without `block =`. Nothing automates
that reading, so check the row on a rerun.

## Acquisition order

Nobody recorded when each sample was injected. `report.parquet` carries `Run.Index`, DIA-NN's own
run counter, which is not alphabetical order here (Spearman against alphabetical rank is 0.207) but
is not confirmed as acquisition order either. Plotted against PC1 for what it is worth: a trend
that tracks timepoint would put the two time contrasts in doubt, while the interaction would be
unaffected because both legs share a session.

```{r acquisition-order}
#| fig-height: 4
run_index <- read_parquet(here("00_Input", "report.parquet"),
                          col_select = c("Run", "Run.Index")) |>
  distinct() |>
  mutate(run_lower = str_to_lower(Run))
idx <- run_index$Run.Index[match(str_to_lower(targets$run), run_index$run_lower)]

pc1 <- prcomp(t(proteins$E), scale. = TRUE)$x[, 1]
plot(idx, pc1, col = as.factor(targets$timepoint), pch = 16,
     xlab = "Run.Index", ylab = "PC1")
```

## Contrasts

```{r contrasts}
contrasts <- makeContrasts(
  BFR_post_vs_pre   = BFR_T2 - BFR_T1,
  HLRT_post_vs_pre  = HLRT_T2 - HLRT_T1,
  BFR_vs_HLRT_at_T1 = BFR_T1 - HLRT_T1,
  BFR_vs_HLRT_at_T2 = BFR_T2 - HLRT_T2,
  interaction       = (BFR_T2 - BFR_T1) - (HLRT_T2 - HLRT_T1),
  levels = design
)
contrasts[rowSums(abs(contrasts)) > 0, ]
```

`BFR_vs_HLRT_at_T1` compares two legs of one person before either was trained, so it is the
negative control and must find nothing.

```{r write}
saveRDS(
  list(design = design, contrasts = contrasts,
       negative_control = "BFR_vs_HLRT_at_T1"),
  here("02_Differential_Expression", "01_Design", "c_data", "design.rds")
)
write_csv(strata,
          here("02_Differential_Expression", "01_Design", "c_data",
               "correlation_strata.csv"))
```

```{r session}
sessionInfo()
```
